### 2021-04-19 draft.Rmd
chunks

---
title: "draft"
author: "Daniel Saggau"
date: "4/19/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##

```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left of=A](UA){$U_A$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,left = 1cm of T](UT){$U_T$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (T);
\draw[->] (UA) -- (A);
\draw[->] (UT) -- (T);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

```{tikz,fig.cap="Atomic Intervention", fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=7pt,
    node distance=3cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left =1 cm of A](UA){$U_A$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5 cm)$) (T) {t};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,right=1 cm of A](UC){$U_C$};
\draw[->] (UA) -- (A);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

```{tikz,fig.cap="Policy Intervention",  fig.align= "center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left of=A](UA){$U_A$};
\node[my node, fill=gray!70] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {$T_\pi$};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,left = 1cm of T](UT){$U_T$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (T);
\draw[->] (UA) -- (A);
\draw[->] (UT) -- (T);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

# Counterfactuals

Process is described as follows:

(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
(b) Action: Exchange $(X = x)$ 
(c) Prediction: Compute $(Y = y)$

# Probabilistic Models vs SCMs

 |Method         | CBN |  SCM
|----------------|---------------------------|------------------------|
| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
|                |                                                                                                                      | 
| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
|                |                                                                                                                      | 
|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors

### 2021-04-19 draft.Rmd
different actions
draft

@@
 ---
-title: "draft"
-author: "Daniel Saggau"
-date: "4/19/2021"
-output: pdf_document
+title: "Structural Causal Models - A Gentle Introduction"
+author: 
+- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
+- Department of Statistics, Ludwig Maximilian University Munich, Germany
+date: "April 2021"
+output: 
+  pdf_document: 
+    number_sections: yes
+    fig_caption: yes
+citation_package: --biblatex
+fontsize: 12
+linestretch: 1.5
 ---
@@
 ```
-
-##
 
@@
 
+ |Method          | Action |  Example
+|-----------------|---------------------------|------------------------|
+| Prediction      | Observation/Co-occurrence | What happened...
+| Intervention    | Do-manipulation           | What happens if ...
+| Counterfactual  | Hypothetical Realities    | What would have happened if...
+
+Table: Pearl - Hierachy of Causation
+
 # Probabilistic Models vs SCMs

### 2021-04-22 paper/draft.Rmd
intro

--- 2021-04-19 draft.Rmd
@@
 ---
-title: "Structural Causal Models - A Gentle Introduction"
+title: "A Gentle Introduction into Structural Causal Models"
 author: 
@@
 linestretch: 1.5
+abstract: This paper provides a gentle introduction into causal inference, focusing on structural causal models and respective causal graphs.
+The first section provides introduces probabilities and will focus on clarifying the difference between observations and causal concepts.
+The second section discusses the hierachy of causation.
+The third section dicusses causal graphs.
+
 ---
@@
 ```
+
+Topic 1: Structural Causal Models (SCMs)
+Structural Causal Models are a general framework to describe data generating mechanisms. 
+The values of variables are described as functions of exogeneous and endogeneous variables. 
+These mechanisms can also be described with directed graphs. 
+Typically one requires for the endogenous variables to fulfill causal sufficiency and the corresponding graph to be acyclic.
+Also, one requires mechanisms to be independent.
+The student should convey and explain the notation on a number of examples and explain the motivation and role of the aforementioned assumptions.
+
+# Introduction
+
+ |Method          | Action |  Example
+|-----------------|---------------------------|------------------------|
+| Prediction      | Observation/Co-occurrence | What happened...
+| Intervention    | Do-manipulation           | What happens if ...
+| Counterfactual  | Hypothetical Realities    | What would have happened if...
+
+Table: Pearl - Hierachy of Causation
+
+## SEM and Structural Causal Model
+
+## Prediction
+
+## Intervention
 
@@
 
-# Counterfactuals
+## Counterfactuals
 
@@
 
- |Method          | Action |  Example
-|-----------------|---------------------------|------------------------|
-| Prediction      | Observation/Co-occurrence | What happened...
-| Intervention    | Do-manipulation           | What happens if ...
-| Counterfactual  | Hypothetical Realities    | What would have happened if...
+## Bayesian Models vs. SCM
 
-Table: Pearl - Hierachy of Causation
-
-# Probabilistic Models vs SCMs
+## Probabilistic Models vs SCMs
 

### 2021-04-22 references/SCM for dynamical Fairness Saggau.Rmd
intro

---
title: "**SCMs for Fairness in Dynamical Systems**"
author: 
- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
- Department of Statistics, Ludwig Maximilian University Munich, Germany
date: "April 2021"
output: 
  pdf_document: 
    number_sections: yes
    fig_caption: yes
citation_package: --biblatex
fontsize: 12
linestretch: 1.5
nocite: '@*'
bibliography: [fair.bib.bib]
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
abstract: Existing research on dynamic fairness builds on reinforced learning structures, in which we model based on agent-environment using markov decision processes. To complement this existing stream of research, Creager et al. (2020) reformulate these Markov-Decision Processes as structural causal models for fairness applications, labelling this method the Fair-Markov-Decision Process. The  difference between these specifications is that Creager et al. (2020) allow for sensitive attributes, within the model specification directly. By conditioning on these sensitive attributes, the authors ensure that the model remains Markovian. They apply this method to two case studies, firstly a case with off-policy intervention and secondly a case for modelling multiple agents. The objective of this paper is to provide an theoretical introduction of causal inference tools with specific emphasis on modelling interventions.
---

# Introduction

We are frequently unable to produce a specific set of information regarding a policy intervention via real-world experiments, due to ethical considerations.
One solution to obtain this set of information is using causal inference tools such as off-policy intervention.
To compute off-policy intervention we can use a structural causal model, using structural equations and modelling latent factors directly.
Meanwhile fairness research has focused on fairness in dynamical systems.
Existing research proposes agent-based models for fairness in dynamical systems.
While this stream of research has focused on modelling fairness as a process paying special attention to intermediate states prior to equilibrium outcomes, traditional stochastic agent-based models are unable to produce counterfactual observations.
Traditional structural causal models can produce off-policy interventions (and counterfactuals), but are primarily concerned with equilibrium outcomes and not with the states within processes.
To address and combine these two issues, this paper discusses the potential benefit casting the Markov Decision Process as a structural causal model to assess fairness in dynamical systems focusing on the recent publication by Creager et al. (2020).
Past literature has focused on rephrasing MDPs as structural causal models but there has been little emphasis on fairness specific domain adaptation. 
Creager et al. (2020) introduce a fair MDP by recasting a traditional MDP as a structural causal model with sensitive attributes within the model specification directly in our system in equations.
This paper will introduce their research as a unifying framework for fairness in dynamic systems.
The following paper proceeds as follows:
Section 2 will discuss fairness in dynamic systems from the perspective of Markov-Decision Processes.
Section 3 presents an introduction into causal inference.
Section 4 briefly compares MDPs and structural causal models. 
Section 5 outlines the current contribution to the literature by Creager et al. (2020).
Section 6 provides an illustration of other considerations and ongoing research.
Section 7 concludes.

# Agent-Based Models for Dynamic Fairness

Casting fairness as a dynamic concept has gained prominence in recent research.
D’ Amour et al. (2020)  model fairness, using a Markov-Decision Process (or in short MDP) Model.
A MDP model changes in the different system components, accounting for the different states.
The focus is not only on the outcome (equilibrium) but on the entire process.
The model undertakes manifold stochastic simulations to model the state after an intervention.
Subsequently, the model aggregates these simulations.
MDP are also very useful for other applications such as modelling complex systems. 
Here the focus is on MDP for a dynamic system (Ness et al., 2019).
Typically, a MDP consists of environmental states, action, transition probabilities (conditional probabilities) and rewards {S;A;P;R}.
The environmental states change as agents take different actions.
Agents can follow different objectives, such as e.g. acting greedy or following a mixed set of objectives.
This enabled the authors to conceptualize repercussions beyond scores such as prediction error.
Using this method, one can assess performance measures at each state of the process. 
D'Amour et al. (2020) study 3 examples namely dynamic lending, attention allocation with multiple agents and college admission as a Stackelberg game.^[For a detailed summary see the supplement or D'Amour et al. (2020)]
Their findings suggest that fairness in dynamic and static settings differentiate. 
Due to dynamics within the model such as a shifting population dynamic, various fairness objectives (e.g. equalizing the true positive rate at each step) converged slower than convergence in the static models.
Despite being stylized, this paper conceptualized shortcomings of exiting fairness notions.
Regardless, the authors do not use any causal methods which have more explanatory power and generally favorable when assessing different policies.
The following section provides a brief introduction to causal inference and causal graphs.

# Causal Inference

Research on causal inference is concerned with making inference on the relationship between variables beyond a given dataset. 
Specifically, causal inference tries to understand the relationship given changing conditions (Pearl, 2009).
Pearl argues that Causal inference requires a deeper understand beyond conditional probabilities.
Essentially, there are two ways to cast causal graphical models, either via a causal bayesian network (CBN) or via a structural causal model (SCM). 
CBNs use **conditional probabilities**, meanwhile the SCM uses a set of **structural equations**.^[for a full performance overview see the supplementary document]
CBNs are frequently used in modern Quantum Mechanics and are essentially stochastic models.
For a SCM, we need to specify a underlying relationship within our dataset beyond conditional independence. 
In a structural causal model (SCM, we define the entire data-generating process based on a set of structural equations.
Structural equation models stem from fields such as Econometrics, Physics, Psychology and other Social Sciences (Pearl, 2009).
While structural equation models are traditionally focused on a unique parametric specification, structural causal models are  non-parametric models and focus on computational graphical models. 
The pivotal advantage of this approach is that SCMs define our latent variables and their relationship with our other variables with our set of structural equations.
This means, that we define the observational distribution but also our intervention distribution. 
The interventions distribution defines our structural mechanism.
Using this distribution, we define e.g. average treatment effects, conditional average treatment effects and the effect of future interventions.
Further, we may define a structure for our hypothetical settings namely the counterfactual distribution.
This distribution is based on our posterior distribution. (Pearl, 2009)
in MDPs we do not explicitly define the latent variables, making it impossible to create counterfactual observations.

## Graphical Models

One common usage of SCMs is to compute computational graphs. 
Graphs entail nodes (or vertices) and edges connecting nodes (structural equations are our edges in the SCM).
One can think of nodes as our variables (latent and observed) and our edges as the causal relationship between these nodes. 
We can differentiate between a directed, partially directed and a un-directed graph.
There is also a distinction between a acyclic and a cyclic graph.
A un-directed graph has edges without directions, connecting the nodes.
A directed graph entails edges with arrows.
An edge starts at the parent and is directed to the child node.
A Node with common children is called a spouse node. 
Another manifestation is an edge with two arrows, which symbolize confounding factors. 
These graphs have bi-directional edges.
Directed graphs can either behave cyclical, e.g. in the case of mutual causation $(A \rightarrow T, T \rightarrow A)$ or the implementation of feedback, or behave acyclic.
Note that acyclic graphs by definition do not have cycles.
This means that there are no variables $(T \rightarrow T)$ with self-directed feedback loops.
Research has pre-dominantly focused on directed acyclic graphs (DAGs) for a number of reasons.^[Note that when we talk about DAGs in this paper we talk about Structural Causal Models. Another common name for them is a 'causal graphs'] 
Pearl (2009) suggests that:

* The framework allows for room for introspection
* Policy Interventions can be modeled easily because of flexible changes in our underlying model. (do-calculus)
* We can isolate policy interventions because we can easily hold other mechanisms constant (Pearl, 2009)
* Independence assumptions are more stable

Similar to Pearl, Creager et al. (2020) suggest that there are manifold benefits of using this DAG framework:

* DAGs visualize complex models, conveying assumptions to non-technical stakeholders
* DAGs assist in the process of making sense of observations from policy intervention

```{tikz,fig.cap= "Probabilistic DAG", fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
\node[my node,right = 1cm of T, fill=gray!30](C){C};
\draw[->] (A) -- (T);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

We will not go into detail about how to obtain information on the true data generating process and the underlying causal assumptions of the model.^[For further information on these issues, see the document of the 'pcalg' package and see algorithms like the IC-Algorithm.]

```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left of=A](UA){$U_A$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,left = 1cm of T](UT){$U_T$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (T);
\draw[->] (UA) -- (A);
\draw[->] (UT) -- (T);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

The unobserved latent factors are the variables filled in white.
The observed variables are the circles filled with gray color.
Essentially these latent factors differentiate the SCM from probabilistic approaches.
Keep in mind, in a SCM we have structural equations (functions). 
The probabilistic model is focused on conditional probabilities.

## Potential Queries

Essentially, there are 3 different queries, namely prediction, intervention and counterfactuals. 
**Prediction** would deal with questions like: Would the prison sentence be fair if we find the defendant to be black. (observation)
**Intervention** would instead deal with issues such as: Would the prison sentence be fair if we make sure that defendant is black. (do-calculus)

```{tikz,fig.cap="Atomic Intervention", fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=7pt,
    node distance=3cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left =1 cm of A](UA){$U_A$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5 cm)$) (T) {t};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,right=1 cm of A](UC){$U_C$};
\draw[->] (UA) -- (A);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

There are two popular types of intervention, one being the atomic intervention.
Atomic intervention means that we take a specific variable and replace it with a constant (T=t).
Another popular method is using policy intervention.
With policy intervention we replace a specific structural equation and define a new probability distribution $\pi$.
We can do that because we make the assumption that these structural equations are autonomous (Pearl, 2009).

Note, $\pi$ is different from historical policies ($f_T=\pi_{hist}$) and often not within our historical data at all ($\pi\neq\pi_{hist}$). (Creager et al., 2020)
When we do not have this historical data and make an estimate solely based on this historical information without having information on the targeted policy, we call this issue an off-policy intervention.

```{tikz,fig.cap="Policy Intervention",  fig.align= "center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left of=A](UA){$U_A$};
\node[my node, fill=gray!70] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {$T_\pi$};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,left = 1cm of T](UT){$U_T$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (T);
\draw[->] (UA) -- (A);
\draw[->] (UT) -- (T);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

To derive our estimates based on interventions, we can control whether our model fulfills certain conditions (Markov Condition) making estimation computationally easier for sequential and simultaneous interventions.^[For a detailed explanation of independence assumptions, Markov conditions and respective factorization for causal bayesian networks compared and structural causal models, see: Pearl, 2009)]
A model is considered semi-markovian, if the diagram derived from our causal model is acyclic.
A semi-markovian model has unobserved, latent variables and background-variables are not independent. 
The model is considered markovian if the error terms are jointly independent and the graph is acyclic (Pearl, 2009).
Based on this condition follows a truncated factorization.
This truncated factorization is defined as:

\begin{equation}
f(x)= \prod ^M_{i=1}f(x_i|pa_j)
\end{equation}

$pa_j$ is our parent node.
M is our set of structural equations.
Estimation with truncated factorization is very flexibly.(Peters et al.,2017)
Judea Pearl provides an extensive chapter (Cpt. 3) on this issue in his 'Causality' book from 2009.

Buidling on that, **Counterfactuals** deal with these questions like: *Would the prison sentence be fair had the defendant been black (given that the applicant is white and the sentence is fair)?*
This is the most challenging query because we need to define the relationship of our variables with latent factors and we need to understand the distribution of our latent factors.
Probabilistic Models cannot provide counterfactual observations because we have no information on our latent factors solely given conditional probabilities.^[see supplementary material]
The process is described as follows:

(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
(b) Action: Exchange $(X = x)$ 
(c) Prediction: Compute $(Y = y)$

In the first step we contextualize the past based on the event $\epsilon$.
In step two we compute $\Delta$ historical data^[minimally] by introducing hypothetical condition.
In step three we compute predictions for new historical observations.

# DAG Models vs. Agent Based Models

Ness et al. (2019) argue that a Markov Process is focused on underlying mechanisms within a system while being unable to undertake counterfactual inference.
Meanwhile, a DAG-based model like the structural causal model is able to undertake counterfactual inference but is unable to identify underlying structural mechanisms.
Furthermore, Hernan (2015) suggests that one can differentiate between the two approaches based on their reliance on data vs theory. 
DAGs are more dependent on data and Agent-based Models are more dependent on theory. 
Arnold et al. (2019) argue that DAGs primarily are interested in fixed effect estimations, looking at average treatment effects in particular. 
ABMs on the other hand are focusing on individual level models, also allowing for loops and spillovers (DAGs do not leave room for loops because of their structure, ensuring that no variable causes itself). 
The focus of ABMs is placed on modelling complexity.
Nevertheless, causal effects within ABMs have been subject to discourse and there is no consensus on how to ensure that estimates are causal within agent based models (Arnold et al., 2019).
Further, the authors argue that DAG-based models are less interested in distributional properties and patterns within natural heterogeneity.
This is pre-dominantly caused because of the strong assumptions underlying DAG-based models,leaving little room to breathe.

# Structural Causal Models for Dynamic Systems

The contribution of Creager et al. (2020) is connecting existing dynamic fairness models and rephrasing them as a structural causal model.^[For a more mathematical approach to the problem of how to recast a MDP as a SCM starting from random differential equations see: Bongers and Mooij, 2018] 
To cast a MDP as a SCM, one needs to modify certain assumptions. 
Firstly, the fair Markov-Decision Process suggested here, is not fully Markovian because the sensitive attributes persist across states.
Irrespective, one can converge the fair-MDP to a MDP if we condition on the sensitive attribute.
This means, we include the sensitive attributes directly in our set of structural equations.
By doing so, we enable fairness considerations for this attribute.

\begin{equation}
A = F_A(U_A)
\end{equation}
\begin{equation}
X^0=F_{X^0}(A, U_{X^0})
\end{equation}
\begin{equation}
T^{k+1} = f_X(X^k,T^k, A, U_{X^{k+1}}), k =0...K
\end{equation}

A is attribute (in the following paper A was used to assign people to group 1 or group 2).
U is our utility.
X is our state.
T is our actions.
$X^k$ is our final reward.

## Off Policy Evaluation

Off-Policy evaluation describes the circumstance where we want to evaluate an policy intervention but there is no data on this policy in our historical data while we are also unable to derive this information because an experimental study would be un-ethical.
Off-Policy evaluation for reinforced learning has been problematic due to issues such as confounding and the lack of introspection (Oberst and Sontag, 2019).
We can use different methods for off-policy evaluation.
Essentially, this paper will focus on model-based off policy evaluation and importance sampling.
There are further combinations of these methods.
Creager et al. (2020) compare such a hybrid model (doubly-estimator) with model-based policy evaluation via a standard regression model.
Model-based off-policy evaluation builds on a parametric model of an underlying Markov-Decision-Process (MDP).
The Markov-Structure definition specifies that our predictions are independent of the historical observations.
When we do not have observations we can also build a model where we make the assumption that they exist, without actually observing them.^[These models in turn are called partially observable Markov Decision Processes (POMDP) (Oberst, 2019)]
The general off policy evaluation process is structured as follows:
We need to firstly define our system of equations.
First, we define the distribution of our attributes (X), secondly we define the historical policies (T) and thirdly we define the outcome distribution (Y).

\begin{equation}
X=f_X(A,U_X)
\end{equation}
\begin{equation}
T=f_T(X,A,U_T)
\end{equation}
\begin{equation}
Y=f_Y(X,A,U_Y)
\end{equation}

This lending example is based on Liu et al. (2018). 
Liu et al. (2018) assume that the dynamics parameters^[transition probabilities in MDP] are known. 
But as suggested by Creager et al. (2020) these parameters are mostly unknown in a real world setting.
These parameters need to be estimated based on our historical data.
One possible estimation is a regression estimate. 
To further improve performance, we can also use the backdoor criterion.
Using the backdoor criterion we ensure that there are no confounding factors.^[As e.g. defined by Pearl (2009) this criterion has two conditions: *1. There are no objects within our admissible set that are descendant of X.* *2. Objects of our admissible set account for all back-door paths (examining X and Y) preventing any spurious relationships.*]

Creager et al. (2020) advocate for the usage of the doubly robust estimator, a hybrid estimator based on a regression- and inverse -propensity estimator.
The regression estimator in our doubly estimator is our control variate. 
The authors define the following steps in their approach:

(a) Define u as a causal quantity (outcome)
(b) Define structural causal model to detect variables suitable for the backdoor criterion
(c) Apply doubly estimator

D'Amour et al. (2020) use the equal opportunity parity equalizing the true positive rate between groups and the maximum profit parity.
We can define the equal opportunity parity as follows:

\begin{equation}
P(\hat{Y}|A=0, Y=1)=P(\hat{Y}|A=1, Y=1)
\end{equation}

Which adds the additional condition that Y has to be 1 in both probabilities.
Adding this additional condition makes equal opportunity the less strict constraint, because it only needs to hold for Y =1.

Creager et al. (2020) also use the demographic parity in their second example with multiple actors.
Mathematically, we can define the demographic parity as follows:

\begin{equation}
P(\hat{Y}|A=1)=P(\hat{Y}|A=0)
\end{equation}

$\hat{Y}$ is the predicted outcome and X the attribute.
This attribute could be a variable like e.g. gender or race. 
The predicted outcome should be equal for different states of the attribute (e.g. between males and females).
The authors use a trade-off parameter, lambda, weighing the importance of our equal opportunity objective relative to the maximum profit parity (Parameter U).

\begin{equation}
\delta_{EQOPP} = |(P(T=1|A=0, Y=1)-P(T=1|A=1,Y=1))|
\end{equation}
\begin{equation}
V_\pi =U - \lambda\delta_{EQOPP}
\end{equation}

Evidence suggest that the double robust estimator is much stronger than the regression estimator looking at the off-policy estimate relative to the trade-off parameter lambda.
As suggested in the paper, this example illustrates that we do not need concise information on system dynamics (transitional probabilities).

## Set Dynamics and Causal Structure with Multi-Agent Intervention

In their second application, they model the intervention of a third party, manually setting the credit score for everyone to 600.
The SCM defines the feature (Group) as a sensitive attribute related to creditworthiness.
We can see a number of interesting results within this model.
The Demographic Parity is more sensitive to intervention and offers more people loans.
The Equal Opportunity Parity is less sensitive because it obeys a less strict fairness constraints and offers less loans to people.
Offering more people loans in this use case also entails a worse outcome for many minority group members. 
Getting a loan but being unable to repay this loans leaves these members bankrupt, worsening their overall outcome.^[see the results section of Creager et al. (2020)]
One advantage of causal graphs specified as SCMs is that they are stable because they are not narrowed done or dependent on the specification of our parameters.
We are solely specifying causal mechanisms and not confining our model to a specific parametric structure.
Nevertheless, sensitivity is a substantial consideration, especially looking at specific model assumptions. 
To ensure that results are robust, they compare various interventional distribution shifts.
Minority group outcomes suggest to be very volatile to this shift over various time steps.
Institutional profits remain robust within their analysis, even when shifting interventional distribution.
This implies that the outcome of minorities is more susceptible to be affected by these fairness measures.

# Conclusions in Creager et al. (2020)

Creager et al. (2020) illustrate the versatility of structural causal models as an expressive simulator for dynamical systems which can be applied to model fairness parities.
These methods function in off-policy interventions and interventions with known dynamics. 
Their approach is an illustration of how causal inference unifies dynamic system modelling and interventions.
Further, different fairness parities behave differently and Creager et al. (2020) illustrate that currently it is not possible to maximize them simultaneously. 
Their findings suggest that the outcomes for the minority groups are **very sensitive** while the overall institutional profits in their example are stable.
The tools that have been introduced, allow for a deep analysis and optimization of various different interventions, including off-policy interventions where there is no historical data and where ethical constraints might render real world experiments troublesome. 
Regardless, these causal methods demand a high level of understanding and further assumptions that, due to the scope of this paper, have merely been touched upon in an incomplete and diluted manner.
Future research on fairness in machine learning could definitely utilize these methods, adding additional complexity providing even more realistic fairness models.

# Considerations

Note, that when working with SCMs need information on the causal structure, also entailing the relationship of unobserved characteristics.
We need to be certain that we have all necessary latent and observed variables within our specification.^[Further, we need some variables that fulfill the backdoor criterion, enabling one to estimate causal effects. In many disciplines such as medicine or economics that may be the case for manifold applications, but unquestionably this may not be the case. ]
While the SCM does not estimate causal effects, the quality is dependent on the quality of the underlying assumptions and structure that we have cast for the SCM.
To summarize, we need to already have a model and cannot determine this model via the SCM.
There are also issues with respect to identification e.g. the identification of our **counterfactual estimates** and the identification of the identification of **fixed solutions** for various initial conditions.
There might be multiple SCMs leading to the same interventional distribution but to different counterfactual estimates (e.g. Probability of Necessity - Probability that treatment was necessary for the outcome).
Some assumptions like the monotonicity assumption work for binary treatment but other treatments such as categorical treatment are less explored. 
Oberst and Sonntag suggest the **Gumbel-Max Structural Causal Models** (2019).^[This method is still rather novel and little has been done with respect to fairness applications.]
This might be an interesting model for applied fairness research in the future.
Sometimes SCMs are unable to provide complete specifications at the equilibrium within dynamical systems.
Blom et al. (2019) examine further generalizations of SCMs, specifically, the causal constraint model (CCM).
These models can give a more flexible representation when there is not a solvable solution.
Solutions might dependent on initial conditions which contradict common assumptions in SCMs.
CCMs provide a more flexible framework to model complex dynamical systems.
Especially looking at future research in this area, it is not unlikely that as fairness models become more complex, such generalizations of SCMs might gain traction.
Unfortunately, currently CCMs are not able to provide graphical representations like the SCM.^[The Authors of the paper suggested that they are currently working on computational graphs for CCMs]
Lastly, we might also be interested in cyclical graphs. 
Likely there are also many fairness applications where variables do reinforce themselves and have reinforcing feedback-loops^[thinking about e.g. 'Learn to Serve' and predictive Policing].
Lastly, Creager et al. (2020) did not explore counterfactuals for dynamical systems.

\newpage

# References

\footnotesize

### 2021-04-22 paper/draft.Rmd
draft stuff

@@
 
- |Method          | Action |  Example
-|-----------------|---------------------------|------------------------|
-| Prediction      | Observation/Co-occurrence | What happened...
-| Intervention    | Do-manipulation           | What happens if ...
-| Counterfactual  | Hypothetical Realities    | What would have happened if...
+## Time in Causality
 
-Table: Pearl - Hierachy of Causation
+In causal research, we are interested in the relationship of our variables outside of a fixed time frame. 
+We want to capture relationships that hold outside of an available dataset and results are intended to be generalizable to other time periods (other datasets).
 
@@
 
-## Prediction
+A lot of research ignores the distinction between structural equation models(SEM) and structural causal models (SCM).
+Essentially, the SCM builds on the SEM but there are some differences.
+Firstly, the structural equation model specification stems from disciplines like economics and psychology and is inherently a parametric specification. 
+The SCM on the other hand is a non-parametric specification.
+Secondly, the SCM allows one to create causal graphs.^[Note that we can also create causal graphs with other models like a bayesian causal network model]
 
-## Intervention
+Nevertheless, both models also have a lot in common.
+
+Structural Causal Models specify the relationship in functional forms.
+This is very different to working with conditional probabilities. 
+There are two popular schools of thought on how to specify causal mechanisms.
+
+There are a number of reasons why people are interested in structural causal models.
+We can specify exogenous and endogenous variables in our model specification. 
+When all our variables are fully specified within our model specification (endogenous variables only) we have a deterministic SCM.
+
+Note that while this is very unreasonable because we assume we can explain all exogenous error terms and further we assume that there are no unobserved variables. 
+Henceforth, when applying these models, they are seldom fully deterministic.
 
@@
 ```
+
+# Common Queries in Causal Models
+
+ |Method          | Action |  Example
+|-----------------|---------------------------|------------------------|
+| Prediction      | Observation/Co-occurrence | What happened...
+| Intervention    | Do-manipulation           | What happens if ...
+| Counterfactual  | Hypothetical Realities    | What would have happened if...
+
+Table: Pearl - Hierachy of Causation
+
+## Prediction
+
+## Intervention
 

### 2021-04-22 paper/draft.Rmd
text

@@
 The third section dicusses causal graphs.
-
 ---
@@
 
-Topic 1: Structural Causal Models (SCMs)
-Structural Causal Models are a general framework to describe data generating mechanisms. 
 The values of variables are described as functions of exogeneous and endogeneous variables. 
@@
 
-## Time in Causality
-
 In causal research, we are interested in the relationship of our variables outside of a fixed time frame. 
-We want to capture relationships that hold outside of an available dataset and results are intended to be generalizable to other time periods (other datasets).
-
-## SEM and Structural Causal Model
-
+We want to capture relationships that hold outside of an available dataset and results are intended to be generalizable to other time periods^[other datasets].
+This paper focuses on structural causal models. 
 A lot of research ignores the distinction between structural equation models(SEM) and structural causal models (SCM).
-Essentially, the SCM builds on the SEM but there are some differences.
+Essentially, the SCM builds on the SEM but there are some important differences.
 Firstly, the structural equation model specification stems from disciplines like economics and psychology and is inherently a parametric specification. 
 The SCM on the other hand is a non-parametric specification.
-Secondly, the SCM allows one to create causal graphs.^[Note that we can also create causal graphs with other models like a bayesian causal network model]
+Secondly, the SCM allows one to create causal graphs.^[Note that we can also create causal graphs with other models like a Bayesian Causal Network Model]
 
 Nevertheless, both models also have a lot in common. 
-
-Structural Causal Models specify the relationship in functional forms.
+Structural Causal Models specify the relationship in functional forms, including latent/unobserved variables.
 This is very different to working with conditional probabilities. 
@@
 There are a number of reasons why people are interested in structural causal models.
-We can specify exogenous and endogenous variables in our model specification. 
+We can specify exogenous and endogenous variables in our model specification for a SCM. 
 When all our variables are fully specified within our model specification (endogenous variables only) we have a deterministic SCM.
@@
 ## Intervention
+
+This paper focuses on atomic, policy and off-policy intervention.
 
@@
 
-## Probabilistic Models vs SCMs
+Bayesian approach causal models via conditional probabilities.
+Meanwhile structural causal models approach causality via functions.
+There are some different inherent to these two approaches.
+Conditional probabilities are less stable for causal methods (prediction and intervention).
+
+SCM are able to create counterfactuals, hypothetical knowledge. 
+CBN are unable to produce counterfactuals because we fail to specify exogenous variables in a probabilistic model.
+CBN only specify observable characteristics. 
+SCMs on the other hand specify latent factors too and are built on various distributional assumptions.
+In a SCM we may specify the counterfactual distribution from which we can draw counterfactual observations.
 

### 2021-04-22 paper/draft.Rmd
moving stuff around and writing about causal graphs

@@
 
-The values of variables are described as functions of exogeneous and endogeneous variables. 
-These mechanisms can also be described with directed graphs. 
-Typically one requires for the endogenous variables to fulfill causal sufficiency and the corresponding graph to be acyclic.
-Also, one requires mechanisms to be independent.
-The student should convey and explain the notation on a number of examples and explain the motivation and role of the aforementioned assumptions.
-
 # Introduction
@@
 This paper focuses on structural causal models. 
-A lot of research ignores the distinction between structural equation models(SEM) and structural causal models (SCM).
+A lot of research ignores the distinction between structural equation models (SEM) and structural causal models (SCM).
 Essentially, the SCM builds on the SEM but there are some important differences.
 Firstly, the structural equation model specification stems from disciplines like economics and psychology and is inherently a parametric specification. 
-The SCM on the other hand is a non-parametric specification.
-Secondly, the SCM allows one to create causal graphs.^[Note that we can also create causal graphs with other models like a Bayesian Causal Network Model]
+The SCM is a non-parametric specification with additional mechanisms to cast causal graphs.^[Note that we can also create causal graphs with other models like a Bayesian Causal Network Model]
 
 Nevertheless, both models also have a lot in common. 
+Structural equation models entail a fully specified set of equations.
+Henceforth the model looks at a system perspective of different variables.
+These different equations are autonomous which allows us to manipulate equations without modifying the rest of the system.
+
 Structural Causal Models specify the relationship in functional forms, including latent/unobserved variables.
@@
 There are two popular schools of thought on how to specify causal mechanisms.
+
+## Baysian Causal Models
+
+Meanwhile structural causal models approach causality via functions.
+There are some different inherent to these two approaches.
+Conditional probabilities are less stable for causal methods (prediction and intervention).
+
+SCM are able to create counterfactuals, hypothetical knowledge. 
+CBN are unable to produce counterfactuals because we fail to specify exogenous variables in a probabilistic model.
+CBN only specify observable characteristics. 
+SCMs on the other hand specify latent factors too and are built on various distributional assumptions.
+In a SCM we may specify the counterfactual distribution from which we can draw counterfactual observations.
+
+ |Method         | CBN |  SCM
+|----------------|---------------------------|------------------------|
+| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
+|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
+|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
+|                |                                                                                                                      | 
+| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
+|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
+|                |                                                                                                                      | 
+|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
+|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
 
@@
 
+We can apply a number of different queries in these different causal models.
+Note that one central advantage of SCMs is that we can cast counterfactuals.
+This is unique to structural causal models due to the specification of our distribution of unobserved variables.
+
  |Method          | Action |  Example
@@
 This paper focuses on atomic, policy and off-policy intervention.
+Atomic intervention means that we replace a variable with a constant so we replace variable $T$ with constant $t$.
+Policy intervention means we replace a variable with a new distribution $\pi$.
+Off-policy intervention is an intervention when we only have historical data and no data on the specific intervention. 
+Off-policy has gained traction in discplines like reinforced learning.
 
@@
 
-## Bayesian Models vs. SCM
+## Causal Graphs
 
-Bayesian approach causal models via conditional probabilities.
-Meanwhile structural causal models approach causality via functions.
-There are some different inherent to these two approaches.
-Conditional probabilities are less stable for causal methods (prediction and intervention).
+Causal graphs are graphical models to visualize causal models. 
+Causal graphs are based on various different causal models.
+A graphical models contains nodes and edges (vertices).
+If we are dealing with a causal graph based on a structural causal model, we usually see latent variables.
+These latent variables are typically differentiated by either using a different color and/or shape than the nodes in probabilistic models where we only special conditional probabilities.
+The absence of a edge between nodes can many different things depending on the underlying model.
+Absence in a SCM means that there is no causal relationship.
+Absence in a BCN means that there is no conditional probability.
 
-SCM are able to create counterfactuals, hypothetical knowledge. 
-CBN are unable to produce counterfactuals because we fail to specify exogenous variables in a probabilistic model.
-CBN only specify observable characteristics. 
-SCMs on the other hand specify latent factors too and are built on various distributional assumptions.
-In a SCM we may specify the counterfactual distribution from which we can draw counterfactual observations.
+The most common causal graph is the acyclic directed graph. 
+A directed graph is a graph where every edge has a direction.
+An un-directed graph has edges but there is no arrow which means that the relationship is unclear.
+An acyclic graph is a graph without cycles.
+That means that a variable does not cause itself, directly or indirectly.
+Given the way we look at time where the relationship between variables is indifferent to time, this representation provides a foundation for graphical models.
 
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+## Causal Mechanisms via Differential Equations
+
+Differential equations can describe a dynamic system accounting for time.
+
+# Conclusion
+
+This paper illustrates the versatile nature of structural causal models.
+Casting causal mechanisms via a functional specification entails many advantages.
+SCMs provide a stable foundation for causal graphs.
+
+# References

### 2021-04-22 references/peters_Causality.Rmd
moving stuff around and writing about causal graphs

---
title: "peters_elements_causal_inference"
author: "Daniel Saggau"
date: "4/22/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 1

Probability theory: 
Statistical learning:

### 2021-04-22 paper/draft.Rmd
prediction and criteria section

@@
 
+Note that prediction in a causal model and prediction in a machine learning context are not the same thing.
+In prediction in causal models we can only modify root nodes.
+We are unable to modify the child nodes in causal prediction.
+Meanwhile, in a machine learning context we can modify any variable.
+We do not have a underlying causal specification, henceforth we are able to modify any variable and undertake prediction. 
+One should note that clearly the prediction in causal models has more explanatory power than the prediction in a vanilla-machine learning model.
+
 ## Intervention
@@
 
+## Criteria
+
+## D-Seperation
+
+### Markovian Condition
+
+#### In Graphical Model
+
+#### in SCM
+
+### Backdoor criteria
+
+```{tikz, backdoor, fig.cap ="Backdoor Criterion",  fig.align="center", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+  \begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=7pt,
+    node distance=3cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\node[my node, fill=gray!30] (A){A};
+\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
+\node[my node,right of=T, fill=gray!30](C){C};
+\draw[->] (A) -- (T);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
+
 ## Causal Mechanisms via Differential Equations

### 2021-04-22 paper/draft.Rmd
draft1

@@
 linestretch: 1.5
-abstract: This paper provides a gentle introduction into causal inference, focusing on structural causal models and respective causal graphs.
-The first section provides introduces probabilities and will focus on clarifying the difference between observations and causal concepts.
-The second section discusses the hierachy of causation.
-The third section dicusses causal graphs.
+abstract: This paper provides a gentle introduction into causal inference focusing on structural causal models. 
 ---
@@
 We want to capture relationships that hold outside of an available dataset and results are intended to be generalizable to other time periods^[other datasets].
+The causal model entails a probabilistic specification but also entails additional information.
+This additional information is a specification for the model when intervention, outcome or feature distributions are modified (Peters et al.,2017).
+
 This paper focuses on structural causal models. 
@@
 There are two popular schools of thought on how to specify causal mechanisms.
+
+The first section provides introduces probabilities and will focus on clarifying the difference between observations and causal concepts.
+The second section discusses the hierachy of causation. 
+The third section dicusses causal graphs. 
+The fourth
 
@@
 
-Differential equations can describe a dynamic system accounting for time.
+Differential equations can describe a dynamic system accounting with time considerations.
 

### 2021-04-22 paper/draft.Rmd
essay

@@
   pdf_document:
+
     number_sections: yes
@@
 linestretch: 1.5
+header-includes:
+- \usepackage{amsmath}
+- \usepackage[makeroom]{cancel}
 abstract: This paper provides a gentle introduction into causal inference focusing on structural causal models. 
@@
 This additional information is a specification for the model when intervention, outcome or feature distributions are modified (Peters et al.,2017).
+Probability theory deals with outcomes given a specific underlying data generating process.
 
@@
 
-## Baysian Causal Models
+## Bayesian Causal Models
 
@@
 An un-directed graph has edges but there is no arrow which means that the relationship is unclear.
+We can depict confounding variables via a edge with two arrows.
 An acyclic graph is a graph without cycles.
@@
 ## Criteria
+
+## Reichenbach common cause principle
+
+If two [random] variables X and Y are dependent when ignoring z and are independent when accounting for z is the true underlying causal influence.
+We can observe Z by looking at the model independent of Z ($X\bcancel{\perp\!\!\!\perp}Y$) and once dependent on Z ($X\perp\!\!\!\perp Y|Z$).(Peters et al., 2017)
+Note that there may be a different explanation for the dependence (selection bias)
+
+## Independence Assumption
 

### 2021-04-22 references/peters_Causality.Rmd
draft1
essay

@@
 Probability theory:
-Statistical learning:
+
+Statistical learning: Capacity measures
+
+Peters et al. (2017) suggest that statistical learning theory has a inverse problem: We have data and want to understand the underlying data generating process and respective underlying structures. 
+This true underlying data generating process is unknown.
+We cannot observe the true DGP based on observational data, alone.
+Learning theory proposes the usage of functional specifications of our model to define the true underlying process that is rule based.
+These rules are pivotal because they allow us to specify the behavior of the model when exposed to unseen observation data.
+To ensure that this specification works, Peters et al. (2017) suggests that statistical learning theory uses the concept of capacity measures. 
+If we have high capacity, our functional specification is very applicable to different dataset.
+If we have low capacity, our specification is not representative of many datasets.
+
+Apart from having the problem of statistical ill posedness (once because the conditional expectation is undefined for values outside of our sample). 
+The second problem is that we usually have no full information for a underlying causal model even when having a fully specified observational distribution.

### 2021-04-24 references/peters_Causality.Rmd
xyz

@@
 The second problem is that we usually have no full information for a underlying causal model even when having a fully specified observational distribution.
+
+Importance of noise and additional complexity due to causal nature. 
+This is called causal learning.
+
+SCM using equations in terms of assignment rather than mathematical equations.
+SCM entails a joint distribution over all observables.
+
+# Time
+
+Time often necessary but not modelled specifically or well defined. 
+Usually relationship envolved dynamical process but it is not always well defined.
+Typically hard sciences involve more concise measures of time than social sciences.
+
+# Relationship mechanical models
+
+SCM somewhere betweren mechanisitic model with differential equations and statistical model
+
+## Chapter 2

### 2021-04-24 paper/draft.Rmd
xyz

@@
 Essentially, the SCM builds on the SEM but there are some important differences.
-Firstly, the structural equation model specification stems from disciplines like economics and psychology and is inherently a parametric specification.
-The SCM is a non-parametric specification with additional mechanisms to cast causal graphs.^[Note that we can also create causal graphs with other models like a Bayesian Causal Network Model]
+Firstly, the structural equation model specification stems from disciplines like economics and psychology and is inherently a parametric specification with algebraic equations.
+The SCM is a non-parametric specification (with assignment equations stemming from CS) with additional mechanisms to cast causal graphs.^[Note that we can also create causal graphs with other models like a Bayesian Causal Network Model]
 

### 2021-04-25 paper/draft.Rmd
notes

@@
 
+The first principle is independence of our mechanisms.
+As a reminder, each equation in a SCM is a separate independent mechanism.
+In a two-variable case of a causal model one can differentiate between the cause distribution and the effect distribution(Peters et al.,2017).
+
+Implications:
+
+1. Change mechanism without changing the rest of the model 
+2. Change of one mechanism does not contain information on other changes
+3. Systems can never be isolated from environment, henceforth weak dependence of noise possible without violating assumption about independent mechanisms
+
 ## D-Seperation

### 2021-04-25 references/peters_Causality.Rmd
xyz
notes

@@
 
-SCM somewhere betweren mechanisitic model with differential equations and statistical model
+SCM somewhere between mechanisitic model with differential equations and statistical model
 
 ## Chapter 2
+
+SEM vs SCM: SEM: Algebraic equations rather than assignments from CS in SCM
+
+Independent mechanisms
+
+Autonomous modules
+in Case example between cause distribution and effect distribution.
+
+Confusion economics + statistics
+
+Economics = E(Y|do(x)) - claim stats no substance 
+Statistics =E(Y|x) - claim SE have no meaning

### 2021-04-25 paper/draft.Rmd
role of time

@@
 The first section provides introduces probabilities and will focus on clarifying the difference between observations and causal concepts.
-The second section discusses the hierachy of causation. 
-The third section dicusses causal graphs. 
-The fourth
+
+## Criteria
+
+## Reichenbach common cause principle
+
+If two [random] variables X and Y are dependent when ignoring z and are independent when accounting for z is the true underlying causal influence.
+We can observe Z by looking at the model independent of Z ($X\bcancel{\perp\!\!\!\perp}Y$) and once dependent on Z ($X\perp\!\!\!\perp Y|Z$).(Peters et al., 2017)
+Note that there may be a different explanation for the dependence (selection bias)
+
+## Independence Assumption
+
+The first principle is independence of our mechanisms.
+As a reminder, each equation in a SCM is a separate independent mechanism.
+In a two-variable case of a causal model one can differentiate between the cause distribution and the effect distribution(Peters et al.,2017).
+
+Implications:
+
+1. Change mechanism without changing the rest of the model 
+2. Change of one mechanism does not contain information on other changes
+3. Systems can never be isolated from environment, henceforth weak dependence of noise possible without violating assumption about independent mechanisms
+
+### Markovian Condition
+
+### Backdoor criteria
+
+```{tikz, backdoor, fig.cap ="Backdoor Criterion",  fig.align="center", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+  \begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=7pt,
+    node distance=3cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\node[my node, fill=gray!30] (A){A};
+\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
+\node[my node,right of=T, fill=gray!30](C){C};
+\draw[->] (A) -- (T);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
+
+### (Causal) Minimality
+### Interventioanl equivalence 
+### Cyclic SCMs
+### d-seperation
+
+A baseline SCM consists of two assingments.
+
+\begin{equation}
+A = N_A
+\end{equation}
+\begin{equation}
+B = f_B(A, N_B)
+\end{equation}
+
+$N_A$ and $N_B$ are independent. 
+A is a direct cause of B.
+A-> B is causal graph (Peters et al., 2017).
+The intervention distribution is unequal to the observational distribution.
+
+**Canonical Representation**
+
+## Causal Graphs
+
+Causal graphs are graphical models to visualize causal models. 
+Causal graphs are based on various different causal models.
+A graphical models contains nodes and edges (vertices).
+If we are dealing with a causal graph based on a structural causal model, we usually see latent variables.
+These latent variables are typically differentiated by either using a different color and/or shape than the nodes in probabilistic models where we only special conditional probabilities.
+The absence of a edge between nodes can many different things depending on the underlying model.
+Absence in a SCM means that there is no causal relationship.
+Absence in a BCN means that there is no conditional probability.  
+The most common causal graph is the acyclic directed graph. 
+A directed graph is a graph where every edge has a direction.
+An un-directed graph has edges but there is no arrow which means that the relationship is unclear.
+We can depict confounding variables via a edge with two arrows.
+An acyclic graph is a graph without cycles.
+That means that a variable does not cause itself, directly or indirectly.
 
@@
 In a SCM we may specify the counterfactual distribution from which we can draw counterfactual observations.
-
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
 
@@
 Off-policy intervention is an intervention when we only have historical data and no data on the specific intervention. 
-Off-policy has gained traction in discplines like reinforced learning.
+Off-policy has gained traction in discplines like reinforced learning.^[Another type of intervention is soft intervention when we only change the noise distribution and keep the prior causal effect]
 
@@
 
-## Causal Graphs
+ |Method         | CBN |  SCM
+|----------------|---------------------------|------------------------|
+| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
+|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
+|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
+|                |                                                                                                                      | 
+| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
+|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
+|                |                                                                                                                      | 
+|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
+|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
 
-Causal graphs are graphical models to visualize causal models. 
-Causal graphs are based on various different causal models.
-A graphical models contains nodes and edges (vertices).
-If we are dealing with a causal graph based on a structural causal model, we usually see latent variables.
-These latent variables are typically differentiated by either using a different color and/or shape than the nodes in probabilistic models where we only special conditional probabilities.
-The absence of a edge between nodes can many different things depending on the underlying model.
-Absence in a SCM means that there is no causal relationship.
-Absence in a BCN means that there is no conditional probability.
-
-The most common causal graph is the acyclic directed graph. 
-A directed graph is a graph where every edge has a direction.
-An un-directed graph has edges but there is no arrow which means that the relationship is unclear.
-We can depict confounding variables via a edge with two arrows.
-An acyclic graph is a graph without cycles.
-That means that a variable does not cause itself, directly or indirectly.
 Given the way we look at time where the relationship between variables is indifferent to time, this representation provides a foundation for graphical models.
-
-## Criteria
-
-## Reichenbach common cause principle
-
-If two [random] variables X and Y are dependent when ignoring z and are independent when accounting for z is the true underlying causal influence.
-We can observe Z by looking at the model independent of Z ($X\bcancel{\perp\!\!\!\perp}Y$) and once dependent on Z ($X\perp\!\!\!\perp Y|Z$).(Peters et al., 2017)
-Note that there may be a different explanation for the dependence (selection bias)
-
-## Independence Assumption
-
-The first principle is independence of our mechanisms.
-As a reminder, each equation in a SCM is a separate independent mechanism.
-In a two-variable case of a causal model one can differentiate between the cause distribution and the effect distribution(Peters et al.,2017).
-
-Implications:
-
-1. Change mechanism without changing the rest of the model 
-2. Change of one mechanism does not contain information on other changes
-3. Systems can never be isolated from environment, henceforth weak dependence of noise possible without violating assumption about independent mechanisms
-
-## D-Seperation
-
-### Markovian Condition
-
-#### In Graphical Model
-
-#### in SCM
-
-### Backdoor criteria
-
-```{tikz, backdoor, fig.cap ="Backdoor Criterion",  fig.align="center", echo =F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-  \begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (A){A};
-\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
-\node[my node,right of=T, fill=gray!30](C){C};
-\draw[->] (A) -- (T);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
 
@@
 
-Differential equations can describe a dynamic system accounting with time considerations.
+SCMs are abstractions of physical processes.
+Differential equations can describe a dynamic cyclic system accounting with more sophisticated time considerations.
+Henceforth SCMs are predominately used for acyclic causal graphs.
+Note that acyclic causal graphs are still dynamical models but they do not have loops where variables reinforce themselves.
 

### 2021-04-25 references/peters_Causality.Rmd
role of time

@@
 Statistics =E(Y|x) - claim SE have no meaning
+
+**statistical vs. algorithmic independence**
+
+# The role of Time
+
+When interested in general cyclic system, we can use differential equations rather than SCMs as the SCM is an abstraction of underlying physical processes. 
+Econometrics decided to view structural equations as algebraic equations.
+Ignore interpretation as abstractions.
+Natural intervention does not violate independence. 
+Others may be complex.
+
+When interested in cyclic systems differential equation unavoidable.
+
+```{r, fig.cap="Peters et al. (2017) Ranking SCMs", include =F}
+library(knitr)
+include_graphics("~/Desktop/SOSE2021/causal_ml/paper/relation_physics.png")
+```

### 2021-04-25 paper/draft.Rmd

--- 2021-04-24 paper/draft.Rmd
@@
 
+## Time in Causality
+
 In causal research, we are interested in the relationship of our variables outside of a fixed time frame. 
@@
 
+## SEM and Structural Causal Model
+
+A lot of research ignores the distinction between structural equation models(SEM) and structural causal models (SCM).
+Essentially, the SCM builds on the SEM but there are some differences.
+Firstly, the structural equation model specification stems from disciplines like economics and psychology and is inherently a parametric specification. 
+The SCM on the other hand is a non-parametric specification.
+Secondly, the SCM allows one to create causal graphs.^[Note that we can also create causal graphs with other models like a bayesian causal network model]
+
+Nevertheless, both models also have a lot in common.
+
+Structural Causal Models specify the relationship in functional forms.
+This is very different to working with conditional probabilities. 
+There are two popular schools of thought on how to specify causal mechanisms.
+
+There are a number of reasons why people are interested in structural causal models.
+We can specify exogenous and endogenous variables in our model specification. 
+When all our variables are fully specified within our model specification (endogenous variables only) we have a deterministic SCM.
+
+Note that while this is very unreasonable because we assume we can explain all exogenous error terms and further we assume that there are no unobserved variables. 
+Henceforth, when applying these models, they are seldom fully deterministic. 
 This paper focuses on structural causal models. 
@@
 Essentially, the SCM builds on the SEM but there are some important differences.
-Firstly, the structural equation model specification stems from disciplines like economics and psychology and is inherently a parametric specification with algebraic equations.
-The SCM is a non-parametric specification (with assignment equations stemming from CS) with additional mechanisms to cast causal graphs.^[Note that we can also create causal graphs with other models like a Bayesian Causal Network Model]
+Firstly, the structural equation model specification stems from disciplines like economics and psychology and is inherently a parametric specification.
+The SCM is a non-parametric specification with additional mechanisms to cast causal graphs.^[Note that we can also create causal graphs with other models like a Bayesian Causal Network Model]
 
@@
 Off-policy has gained traction in discplines like reinforced learning.
+
+# Common Queries in Causal Models
+
+ |Method          | Action |  Example
+|-----------------|---------------------------|------------------------|
+| Prediction      | Observation/Co-occurrence | What happened...
+| Intervention    | Do-manipulation           | What happens if ...
+| Counterfactual  | Hypothetical Realities    | What would have happened if...
+
+Table: Pearl - Hierachy of Causation
+
+## Prediction
+
+## Intervention
 

### 2021-04-25 paper/draft.Rmd

--- 2021-04-25 paper/draft.Rmd
@@
 
+# Common Queries in Causal Models
+
+ |Method          | Action |  Example
+|-----------------|---------------------------|------------------------|
+| Prediction      | Observation/Co-occurrence | What happened...
+| Intervention    | Do-manipulation           | What happens if ...
+| Counterfactual  | Hypothetical Realities    | What would have happened if...
+
+Table: Pearl - Hierachy of Causation
+
+## Prediction
+
+## Intervention
+
 ```{tikz,fig.cap="Atomic Intervention", fig.align="center", echo =F}

### 2021-04-26 references/vanilla_presentation .Rmd
vanilla presentation for fairness ml

---
title: "Fairness in Machine Learning WS20/21"
subtitle: "Fostering Notions of Dynamic Fairness via Structural Causal Models"
date: "03/05/2021"
author: "Daniel Saggau"
output:
  beamer_presentation:
    theme: Madrid
    colortheme: spruce
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

- Fairness in dynamical system has become important because equalizing true positive rate at each step does not converge as fast in systems with e.g. population dynamics 
- Research on dynamical systems has focused on markov decision processes
- Prior research on causal fairness has focused in static systems 
- Recently Creager et al. (2020) proposed a structural causal model for dynamical systems

# Why care about causal fairness models?

- Fairness is not static (D'Amour et al. 2020) introduce shortcomings of existing fairness correction measures
- From a modelling perspective, we can improve the specification (Markov Decision Process)
- Regular models may express a model in conditional probabilities (probabilistic model) or may be expressed as differential equation 
- SCMs are cast in functional form which is more stable (but may also be expressed as differential equation)
- Additionally, we specify the model beyond conditional probabilities -> we specify the latent variables (exogenous variables not observable within our dataset)

# What is a SCM

- Functional specification of our model including latent factors
- Probabilistic specification + additional knowledge

# SCMs for Fairness in dynamical Systems

- Fair-MDP
- Judea Pearl:do-calculus
-

# SCM vs Causal Bayesian Network (Probabilistic Model)

# Exisiting Considerations

- Counterfactuals for Fairness in Dynamical Systems 
- Off-policy estimation (model based (regression) or model free estimation (propensity weight))

# Time and Causal Inference

- Structural causal models are abstractions of physical processes
-

# Adding to existing Literature

**Methodological:**

- Cyclic Structural Causal Models with actual reinforcing loops
- Semi-Deterministic SCMs (deterministic -> all variables are known)

**Application:**

- Off Policy Interventions

### 2021-04-26 paper/draft.Rmd

--- 2021-04-25 paper/draft.Rmd
@@
 abstract: This paper provides a gentle introduction into causal inference focusing on structural causal models.
+Structural Causal Models are abstractions of underlying physical mechanisms.
+SCMs are used for causal inference.
+This paper briefly introduces the role of time in causality.
+I will use a two and three variable model to illustrate underlying assumptions of structural causal models.
 ---
@@
 
-In causal research, we are interested in the relationship of our variables outside of a fixed time frame. 
-We want to capture relationships that hold outside of an available dataset and results are intended to be generalizable to other time periods^[other datasets].
+In causal research,we want to capture relationships that hold outside of an available dataset and results are intended to be generalizable to other time periods^[other datasets].
 The causal model entails a probabilistic specification but also entails additional information.
 This additional information is a specification for the model when intervention, outcome or feature distributions are modified (Peters et al.,2017).
-Probability theory deals with outcomes given a specific underlying data generating process.
 
+A lot of research ignores the distinction between structural equation models(SEM) and structural causal models (SCM).
+Essentially, the SCM builds on the SEM but there are some differences.
+
+Structural Causal Models specify the relationship in functional forms.
+This is very different to working with conditional probabilities. 
+There are two popular schools of thought on how to specify causal mechanisms.
+
+There are a number of reasons why people are interested in structural causal models.
+We can specify exogenous and endogenous variables in our model specification. 
+When all our variables are fully specified within our model specification (endogenous variables only) we have a deterministic SCM.
+
+Note that while this is very unreasonable because we assume we can explain all exogenous error terms and further we assume that there are no unobserved variables. 
+Henceforth, when applying these models, they are seldom fully deterministic. 
 This paper focuses on structural causal models.
-A lot of research ignores the distinction between structural equation models (SEM) and structural causal models (SCM).
-Essentially, the SCM builds on the SEM but there are some important differences.
+
 Firstly, the structural equation model specification stems from disciplines like economics and psychology and is inherently a parametric specification with algebraic equations.
@@
 Henceforth the model looks at a system perspective of different variables.
-These different equations are autonomous which allows us to manipulate equations without modifying the rest of the system.
+These different equations are autonomous.
+This allows us to manipulate equations without modifying the rest of the system.
 
@@
 This is very different to working with conditional probabilities. 
-There are two popular schools of thought on how to specify causal mechanisms.
-
+There are two popular schools of thought on how to specify causal mechanisms
 The first section provides introduces probabilities and will focus on clarifying the difference between observations and causal concepts.

### 2021-04-30 paper/draft.Rmd
abstract

@@
 abstract: This paper provides a gentle introduction into causal inference focusing on structural causal models.
-Structural Causal Models are abstractions of underlying physical mechanisms.
-SCMs are used for causal inference.
-This paper briefly introduces the role of time in causality.
-I will use a two and three variable model to illustrate underlying assumptions of structural causal models.
+Structural Causal Models (SCM) are abstractions of underlying physical mechanisms, extending probabilisitc models by providing an exhaustive specification for the data generating process.
+Probabilistic models specify conditional probabilities, meanwhile SCMs specify the underlying structure of both endogenous and exogenous variables.
+Structural Causal Models are an generalization of structural equation models with some further extensions to compute causal graphs.
+This essay will explore fundamental pillars within causal inference starting with the differences between probablisitic models and causal models, followed by the hierachy of causation.
+I will briefly dicuss the role of time in causal models and provide instructions on various causal graphs focusing on acyclic graphical models, the most prominent causal graph model.
 ---

### 2021-04-30 paper/draft.Rmd
yz

--- 2021-04-26 paper/draft.Rmd
@@
 - \usepackage[makeroom]{cancel}
-abstract: This paper provides a gentle introduction into causal inference focusing on structural causal models.
-Structural Causal Models are abstractions of underlying physical mechanisms.
-SCMs are used for causal inference.
-This paper briefly introduces the role of time in causality.
-I will use a two and three variable model to illustrate underlying assumptions of structural causal models.
+abstract: Interest in a deeper understand of relationships between variables has increased the popularity of causal models. There are various different causal model specifications.
+I will explore two models namely (1) Bayesian causal networks (BCN) and (2) structural causal models (SCM).
+Probabilisitic specifications such as a BCN cast a model based on conditional probabilities.
+SCMs cast a model based on functions.
+SCMs extend probabilisitic models by specifying the data generating process rather than solely focusing on conditional probabilities.
+These SCMs can be applied to number of different queries.
+Using Pearl's hierachy of causation (2009), I focus on prediction, intervention and counterfactuals.
+To further clarify central notions of causal modelling, this paper briefly discusses the role of time in causality and will illustrate an general overview of the history of causal modelling to contexualize the SCM modelling approach.
+I will also introduce causal graphs, specifically focusing on the popular directed acyclical graphical (DAG) models, to highlight differences in various queries and modelling assumptions.
 ---
@@
 # Introduction
+
+"Correlation does not imply Causation" is one of the most pivotal lessons many students learn during introductory undergraduate statistics courses.
 

### 2021-04-30 paper/abstract.Rmd
abstract draft

---
title: "A Gentle Introduction into Structural Causal Models"
author: 
- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
- Department of Statistics, Ludwig Maximilian University Munich, Germany
date: "April 2021"
output: 
  pdf_document:

    number_sections: yes
    fig_caption: yes
citation_package: --biblatex
fontsize: 12
linestretch: 1.5
header-includes:
- \usepackage{amsmath}
- \usepackage[makeroom]{cancel}

---

# Abstract

Interest in a deeper understand of relationships between variables has increased the popularity of causal models. I will explore two popular causal specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM) focusing on the latter. Probabilistic specifications such as a BCN cast a model based on conditional probabilities. SCMs cast a model based on functions. Further, SCMs extend probabilistic models by specifying the data generating process rather than solely employing on conditional probabilities. Causal models can be applied to number of different modelling problems. Using Pearl's hierarchy of causation (2009), I focus on predictions, interventions and counterfactuals. To clarify central notions of causal modelling, this paper briefly discusses the role of time in causality and provides a general overview of the history of causal modelling to contextualize the SCM modelling approach. I will also introduce causal graphs, concentrating on the popular directed acyclical graphical (DAG) models, to highlight differences in mentioned queries and modelling approaches.

### 2021-04-30 paper/draft.Rmd
abstract draft

@@
 - \usepackage[makeroom]{cancel}
-abstract: Interest in a deeper understand of relationships between variables has increased the popularity of causal models. There are various different causal model specifications.
-I will explore two models namely (1) Bayesian causal networks (BCN) and (2) structural causal models (SCM).
+abstract: Interest in a deeper understand of relationships between variables has increased the popularity of causal models.
+There are various different causal model specifications.I will explore two models namely (1) Bayesian causal networks (BCN) and (2) structural causal models (SCM).
 Probabilisitic specifications such as a BCN cast a model based on conditional probabilities.

### 2021-04-30 paper/abstract.Rmd
paper

@@
 
-Interest in a deeper understand of relationships between variables has increased the popularity of causal models.
-I will explore two popular causal specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM) focusing on the latter.
+The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling. 
+To provide a comprehensive understanding of causal modelling, I introduce two prominent causal model specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM), focusing on the latter.
 Probabilistic specifications such as a BCN cast a model based on conditional probabilities. 
-SCMs cast a model based on functions.
-Further, SCMs extend probabilistic models by specifying the data generating process rather than solely employing on conditional probabilities.
-Causal models can be applied to number of different modelling problems.
-Using Pearl's hierarchy of causation (2009), I focus on predictions, interventions and counterfactuals.
-To clarify central notions of causal modelling, this paper briefly discusses the role of time in causality and provides a general overview of the history of causal modelling to contextualize the SCM modelling approach.
-I will also introduce causal graphs, concentrating on the popular directed acyclical graphical (DAG) models, to highlight differences in mentioned queries and modelling approaches.
+SCMs cast a model based on assignment functions and extend probabilistic models by specifying the data generating process rather than solely utilizing conditional probabilities.
+Another difference between these models is their ability to address different queries such as *predictions*, *interventions* and *counterfactuals*.
+These queries are part of Pearl's causal hierarchy (2009). 
+Pearl matches these queries with their respective actions namely *observing*, *doing* and *imagining*.
+I compare the feasibility of addressing these queries and undertaking respective actions for both specifications.
+To contextualize SCMs within the field of causality, I also discuss the role of time in causal inference and provide a brief overview of the history of causal modelling.
+This paper uses various directed acyclic graphs to highlight the differences in these modelling approaches.
+The insights of this paper can be used as a baseline for subsequent research on structural causal models.

### 2021-05-02 history causal modelling.Rmd
history beginning

---
title: "Introduction"
author: "Daniel Saggau"
date: "5/2/2021"
output: 
  pdf_document:
    number_sections: yes
    fig_caption: yes
citation_package: --biblatex
bibliography: [ref.bib]
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# History of Causal Models

There is no clear consensus on when exactly  strcutal causal modelling originated.
Judea Pearl argues that path analysis, a method proposed by Sewall Wright in the 1920s is the foundation of modern structural causal models.
Sewall Wright, a Statistican and Geneticist, sparked the interest in causal modelling with his path analysis which subsequently was adopted by various other disciplines such as econometrics, psychometrics and sociology.
This path analysis is a structural equation model with one variable per indicator.

Thereafter the Norwegian Economist Trygve Haavelmo (1944) provided a modelling framework for the field of economics in is paper 'The Probability Approach in Econometrics.'
One could argue that this paper provided a new of level of sophistication to the discpline of Econometrics as a distinct field. 
Haavelmo was awarded the Nobel Memorial Prize in Economic Science in 1989 for his work on this issue. 
His work focused on

Tjalling Koopmans

## Bayesian Networks

## SEMs

## SCMs

# References

### 2021-05-02 intro.Rmd
intro

---
title: "A Gentle Introduction into Structural Causal Models"
author: 
- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
- Department of Statistics, Ludwig Maximilian University Munich, Germany
date: " April 5th, 2021"
output: 
  pdf_document:
        number_sections: yes
        fig_caption: yes
fontsize: 12
linestretch: 1.5
citation_package: --biblatex
bibliography: [ref.bib]
nocite: '@*'
abstract: The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling.  To provide a comprehensive understanding of causal modelling, I introduce two prominent causal model specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM), focusing on the latter. Probabilistic specifications such as a BCN cast a model based on conditional probabilities. SCMs cast a model based on assignment functions and extend probabilistic models by specifying the data generating process rather than solely utilizing conditional probabilities. Another difference between these models is their ability to address different queries such as *predictions*, *interventions* and *counterfactuals*. These queries are part of Pearl's causal hierarchy (2009).Pearl matches these queries with their respective actions namely *observing*, *doing* and *imagining*. I compare the feasibility of addressing these queries and undertaking respective actions for both specifications. To contextualize SCMs within the field of causality, I also discuss the role of time in causal inference and provide a brief overview of the history of causal modelling. This paper uses various directed acyclic graphs to highlight the differences in these modelling approaches. The insights of this paper can be used as a baseline for subsequent research on structural causal models.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Most undergraduate students that take a class in introductory statistics have heard the phrase 'correlation does not imply causation'.
While correlation implies co-occurrence, people are frequently interesting a causal understanding of relationships between variables.
One may argue that especially in high stake settings, algorithmic decision making based on co-occurrence is insufficient.
Causal modelling has become a pivotal tool for interpretable modelling.
There are different schools of thought on how to optimally model causal relationships.
Structural Causal Models and Bayesian Causal Networks are two prominent causal models.
Bayesian Causal Networks cast a model based on conditional probabilities.
All relationships are defined in conditional probabilities, excluding exogenous variables.

Structural Causal Models (SCM) are a nonparametric modification of structural equations models.
Structural equations models (SEM) are parametric causal models which are very popular in fields like economics, psychology and sociology.
Here we differentiate between these two models because I focus on SCMs rather than SEMs.
While both have a similar basis, they are some differences that I will further addressed in this essay.
One should point out that there is a lot of controversy around SEMs because many scholars challenge the value of these parameters.
As pointed out by Pearl (2012), SEMs have been deemed useless because these parametric specifications are never correct.
The central idea of a SCM is to provide a model specifying an underlying causal process, without getting caught up in redundant parameter specifications that are never reflective of the true data generating process regardless of specification. 
Henceforth SCMs can be seen as heuristic versions of a SEM.
SCMs use assignment equations to specify an underlying data-generation-process, or in short DGP.^[There is a lot of confusion around equations and what exactly they imply. Here we call them assignment equations rather than just equations to ensure that there is no misconception. I want to clarify that I mean equations as used in a computer science rather than an equation in mathematics.]

Beyond the school of thought, there is also a hierarchy of causation which Pearl introduced in 2009.
This hierarchy

Another important factor in causal research is the understanding of time. 
Causal models mostly disregard concise notions of time and make the strong assumption that relationships between variables hold beyond the confinement of time.
Nevertheless, some research has also looked at causal models with a more concrete specification of time, treating time more like one would in a physical mechanism in hard sciences. 
In these specifications, we can use differential equations rather than assignment equations.

Understanding statistical models beyond black-box specifications sparked the wave of interpretable machine learning.
One stream of research in interpretable machine learning is causal modelling. 
This paper addresses the linkage between standard machine learning approaches and causal machine learning.

Existing literature has provided an excellent introduction to structural causal models and bayesian causal networks. 
Judea Pearl (2009) provided a comprehensive and accessible introduction into this topic with his book on causality.
Peters et al. (2017) added some elements, paying a lot of attention to the similarities and differences between causal modelling and physical sciences.

This paper brings together these contributions and attempts to provide a comprehensive summary of structural causal models.
The insights in this paper are an abstraction of these pieces of work, offering the reading a basic understanding of the topic for future research and an overview of the current research.

The subsequent section examines central assumptions in causal modelling. 
The second section looks at the difference SCMs and BCNs.
The section thereafter studies the hierachy of causation, looking at the different queries in the SCM and the BCN.

# References

### 2021-05-02 paper/abstract.Rmd
intro

@@
   pdf_document: 
-
     number_sections: yes

### 2021-05-02 intro.Rmd
graphs

@@
 Bayesian Causal Networks cast a model based on conditional probabilities.
+
 All relationships are defined in conditional probabilities, excluding exogenous variables.
+In Figure 1 we can see that an example of what a probabilistic model looks like.
+All variables in our graph are observable variables. 
+We do not specify an exogenous variables.
+
+```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=7pt,
+    node distance=3cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\node[my node] (A){A};
+\node[my node,right =1 cm of A](C){C};
+\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
+\draw[->] (A) -- (T);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
 
@@
 I want to clarify that I mean equations as used in a computer science rather than an equation in mathematics.]
+
+```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+  \begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=6pt,
+    node distance=2cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
+\node[my node, fill=gray!30] (A){A};
+\node[latent,left of=A](UA){$U_A$};
+\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
+\node[my node,right = 1 cm of T, fill=gray!30](C){C};
+\node[latent,left = 1cm of T](UT){$U_T$};
+\node[latent,right = 1cm of A](UC){$U_C$};
+\draw[->] (A) -- (T);
+\draw[->] (UA) -- (A);
+\draw[->] (UT) -- (T);
+\draw[->] (UC) -- (C);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
 
@@
 
+The rest of the paper is structured as follows: 
 The subsequent section examines central assumptions in causal modelling. 

### 2021-05-02 intro.Rmd
hierachy of causation

@@
 
+ |Method          | Action |  Example
+|-----------------|---------------------------|------------------------|
+| Prediction      | Observation/Co-occurrence | What happened...
+| Intervention    | Do-manipulation           | What happens if ...
+| Counterfactual  | Hypothetical Realities    | What would have happened if...
+
 Another important factor in causal research is the understanding of time. 
@@
 The subsequent section examines central assumptions in causal modelling. 
-The second section looks at the difference SCMs and BCNs.
 The section thereafter studies the hierachy of causation, looking at the different queries in the SCM and the BCN.

### 2021-05-02 intro.Rmd
intro

@@
 Beyond the school of thought, there is also a hierarchy of causation which Pearl introduced in 2009.
-This hierarchy
 
@@
 | Counterfactual  | Hypothetical Realities    | What would have happened if...
+
+This hierarchy examines different queries, their actions and the respective hierachical rank.
+The first query is prediction where we examine based on observations.
+The second query is interventions, where we can use Pearls (2009) do-calculus to examine what happens if certain conditions holds.
+Note that there are various different types of intervention, such as atomic intervention, where we set a variable to a constant, or e.g.
+policy intervention where we specify a different function for this specific variable.
+
+```{tikz,fig.cap="Policy Intervention",  fig.align= "center", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+  \begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=6pt,
+    node distance=2cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
+\node[my node, fill=gray!30] (A){A};
+\node[latent,left of=A](UA){$U_A$};
+\node[my node, fill=gray!70] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {$T_\pi$};
+\node[my node,right = 1 cm of T, fill=gray!30](C){C};
+\node[latent,left = 1cm of T](UT){$U_T$};
+\node[latent,right = 1cm of A](UC){$U_C$};
+\draw[->] (A) -- (T);
+\draw[->] (UA) -- (A);
+\draw[->] (UT) -- (T);
+\draw[->] (UC) -- (C);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
 

### 2021-05-03 intro.Rmd
intro stuff

@@
 
- |Method          | Action |  Example
-|-----------------|---------------------------|------------------------|
-| Prediction      | Observation/Co-occurrence | What happened...
-| Intervention    | Do-manipulation           | What happens if ...
-| Counterfactual  | Hypothetical Realities    | What would have happened if...
+\footnotesize
+ |Method          | Action |  Example | ML Method | 
+|---------------------|-------------|-------------------------|-------------------|
+| Prediction $P(a|b)$               | Co-occurrence             | What happened...               | Supervised/Unsupervised  
+| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            | Reinforcement Learning      
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if... | Reinforcement Learning/Interpretable ML
+
+Table: Pearls Hierachy of Causation (2009)
+
+\normalsize
 

### 2021-05-03 appendix.Rmd
intro + appendix

--- 2021-05-03 intro.Rmd
@@
 ---
-title: "A Gentle Introduction into Structural Causal Models"
-author: 
-- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
-- Department of Statistics, Ludwig Maximilian University Munich, Germany
-date: " April 5th, 2021"
-output: 
-  pdf_document:
-        number_sections: yes
-        fig_caption: yes
-fontsize: 12
-linestretch: 1.5
-citation_package: --biblatex
-bibliography: [ref.bib]
-nocite: '@*'
-abstract: The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling.
-To provide a comprehensive understanding of causal modelling, I introduce two prominent causal model specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM), focusing on the latter.
-Probabilistic specifications such as a BCN cast a model based on conditional probabilities.
-SCMs cast a model based on assignment functions and extend probabilistic models by specifying the data generating process rather than solely utilizing conditional probabilities.
-Another difference between these models is their ability to address different queries such as *predictions*, *interventions* and *counterfactuals*.
-These queries are part of Pearl's causal hierarchy (2009).Pearl matches these queries with their respective actions namely *observing*, *doing* and *imagining*.
-I compare the feasibility of addressing these queries and undertaking respective actions for both specifications.
-To contextualize SCMs within the field of causality, I also discuss the role of time in causal inference and provide a brief overview of the history of causal modelling.
-This paper uses various directed acyclic graphs to highlight the differences in these modelling approaches.
-The insights of this paper can be used as a baseline for subsequent research on structural causal models.
+title: "appendix"
+author: "Daniel Saggau"
+date: "5/3/2021"
+output: pdf_document
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-# Introduction
-
-Most undergraduate students that take a class in introductory statistics have heard the phrase 'correlation does not imply causation'.
-While correlation implies co-occurrence, people are frequently interesting a causal understanding of relationships between variables.
-One may argue that especially in high stake settings, algorithmic decision making based on co-occurrence is insufficient.
-Causal modelling has become a pivotal tool for interpretable modelling.
-There are different schools of thought on how to optimally model causal relationships.
-Structural Causal Models and Bayesian Causal Networks are two prominent causal models.
-Bayesian Causal Networks cast a model based on conditional probabilities.
-
-All relationships are defined in conditional probabilities, excluding exogenous variables.
-In Figure 1 we can see that an example of what a probabilistic model looks like.
-All variables in our graph are observable variables. 
-We do not specify an exogenous variables.
-
-```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\node[my node] (A){A};
-\node[my node,right =1 cm of A](C){C};
-\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
-\draw[->] (A) -- (T);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
-
-Structural Causal Models (SCM) are a nonparametric modification of structural equations models.
-Structural equations models (SEM) are parametric causal models which are very popular in fields like economics, psychology and sociology.
-Here we differentiate between these two models because I focus on SCMs rather than SEMs.
-While both have a similar basis, they are some differences that I will further addressed in this essay.
-One should point out that there is a lot of controversy around SEMs because many scholars challenge the value of these parameters.
-As pointed out by Pearl (2012), SEMs have been deemed useless because these parametric specifications are never correct.
-The central idea of a SCM is to provide a model specifying an underlying causal process, without getting caught up in redundant parameter specifications that are never reflective of the true data generating process regardless of specification.
-Henceforth SCMs can be seen as heuristic versions of a SEM.
-SCMs use assignment equations to specify an underlying data-generation-process, or in short DGP.^[There is a lot of confusion around equations and what exactly they imply.
-Here we call them assignment equations rather than just equations to ensure that there is no misconception.
-I want to clarify that I mean equations as used in a computer science rather than an equation in mathematics.]
-
-```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-  \begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=6pt,
-    node distance=2cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (A){A};
-\node[latent,left of=A](UA){$U_A$};
-\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
-\node[my node,right = 1 cm of T, fill=gray!30](C){C};
-\node[latent,left = 1cm of T](UT){$U_T$};
-\node[latent,right = 1cm of A](UC){$U_C$};
-\draw[->] (A) -- (T);
-\draw[->] (UA) -- (A);
-\draw[->] (UT) -- (T);
-\draw[->] (UC) -- (C);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
-
-Beyond the school of thought, there is also a hierarchy of causation which Pearl introduced in 2009.
 
@@
 | Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            | Reinforcement Learning      
-| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if... | Reinforcement Learning/Interpretable ML
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if... | Reinforcement Learning / Interpretable ML
 
 Table: Pearls Hierachy of Causation (2009)
-
 \normalsize
-
-This hierarchy examines different queries, their actions and the respective hierachical rank.
-The first query is prediction where we examine based on observations.
-The second query is interventions, where we can use Pearls (2009) do-calculus to examine what happens if certain conditions holds.
-Note that there are various different types of intervention, such as atomic intervention, where we set a variable to a constant, or e.g.
-policy intervention where we specify a different function for this specific variable.
-
-```{tikz,fig.cap="Policy Intervention",  fig.align= "center", echo =F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-  \begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=6pt,
-    node distance=2cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (A){A};
-\node[latent,left of=A](UA){$U_A$};
-\node[my node, fill=gray!70] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {$T_\pi$};
-\node[my node,right = 1 cm of T, fill=gray!30](C){C};
-\node[latent,left = 1cm of T](UT){$U_T$};
-\node[latent,right = 1cm of A](UC){$U_C$};
-\draw[->] (A) -- (T);
-\draw[->] (UA) -- (A);
-\draw[->] (UT) -- (T);
-\draw[->] (UC) -- (C);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
-
-Another important factor in causal research is the understanding of time. 
-Causal models mostly disregard concise notions of time and make the strong assumption that relationships between variables hold beyond the confinement of time.
-Nevertheless, some research has also looked at causal models with a more concrete specification of time, treating time more like one would in a physical mechanism in hard sciences. 
-In these specifications, we can use differential equations rather than assignment equations.
-
-Understanding statistical models beyond black-box specifications sparked the wave of interpretable machine learning.
-One stream of research in interpretable machine learning is causal modelling. 
-This paper addresses the linkage between standard machine learning approaches and causal machine learning.
-
-Existing literature has provided an excellent introduction to structural causal models and bayesian causal networks. 
-Judea Pearl (2009) provided a comprehensive and accessible introduction into this topic with his book on causality.
-Peters et al. (2017) added some elements, paying a lot of attention to the similarities and differences between causal modelling and physical sciences.
-
-This paper brings together these contributions and attempts to provide a comprehensive summary of structural causal models.
-The insights in this paper are an abstraction of these pieces of work, offering the reading a basic understanding of the topic for future research and an overview of the current research.
-
-The rest of the paper is structured as follows: 
-The subsequent section examines central assumptions in causal modelling. 
-The section thereafter studies the hierachy of causation, looking at the different queries in the SCM and the BCN.
-
-# References

### 2021-05-03 intro.Rmd
intro + appendix

--- 2021-05-02 intro.Rmd
@@
 
+In figure 2 we can see a structural causal model. 
+The square nodes represent the latent variables
+The circle nodes represent the observed variables.
+The arrows are our structural equations and depict the relationship between two variables.
+We can immediately differentiate probabilistic and structural equation models by looking for latent variables in our directed acyclic graph.
+^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks.
+For further information see Pearl (2009).
+Due to the confined scope of this paper, I will not explore cyclic structures.]
+
 Beyond the school of thought, there is also a hierarchy of causation which Pearl introduced in 2009.
 
- |Method          | Action |  Example
-|-----------------|---------------------------|------------------------|
-| Prediction      | Observation/Co-occurrence | What happened...
-| Intervention    | Do-manipulation           | What happens if ...
-| Counterfactual  | Hypothetical Realities    | What would have happened if...
+Note that there are further applications of these different queries, but they are applied to assosciational learning. 
+One example is the application of counterfactuals for deep unsupervised learning. 
+Qin et. al (2019) use counterfactuals for story generation.
+While they employ counterfactuals for their loss function, technically they still learn associational structures namely word embeddings.
 

### 2021-05-03 intro.Rmd
intro

@@
 All relationships are defined in conditional probabilities, excluding exogenous variables.
-In Figure 1 we can see that an example of what a probabilistic model looks like.
-All variables in our graph are observable variables. 
-We do not specify an exogenous variables.
-
 ```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
@@
 ```
+In Figure 1 we can see that an example of what a probabilistic model looks like.
+All variables in our graph are observable variables. 
+We do not specify an exogenous variables.
 
@@
 
+\footnotesize
+ |Method          | Action |  Example | Usage | 
+|------------------|-------------|--------------------|-------------------|
+| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
+| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           |SCM
+
+Table: Pearls Hierachy of Causation (2009)
+
+\normalsize
+
 Beyond the school of thought, there is also a hierarchy of causation which Pearl introduced in 2009.
-
-Note that there are further applications of these different queries, but they are applied to assosciational learning. 
-One example is the application of counterfactuals for deep unsupervised learning. 
-Qin et. al (2019) use counterfactuals for story generation.
-While they employ counterfactuals for their loss function, technically they still learn associational structures namely word embeddings.
-
+Pearl focuses on three layers namely association, intervention and counterfactuals. 
+A higher level implies a deeper more detailed knowledge of the relationship between your variables.
 This hierarchy examines different queries, their actions and the respective hierachical rank.
 The first query is prediction where we examine based on observations.
+We can see that vanilla-machine learning methods, bayesian networks and classical regression are at the lowest level in the causal hierarchy.
 The second query is interventions, where we can use Pearls (2009) do-calculus to examine what happens if certain conditions holds.
@@
 policy intervention where we specify a different function for this specific variable.
+Causal bayesian networks, markov decision processes and reinforcement learning frequent these intervention methods for e.g. policy estimation.
 

### 2021-05-03 intro.Rmd
intro

@@
 nocite: '@*'
-abstract: The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling.
-To provide a comprehensive understanding of causal modelling, I introduce two prominent causal model specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM), focusing on the latter.
-Probabilistic specifications such as a BCN cast a model based on conditional probabilities.
-SCMs cast a model based on assignment functions and extend probabilistic models by specifying the data generating process rather than solely utilizing conditional probabilities.
-Another difference between these models is their ability to address different queries such as *predictions*, *interventions* and *counterfactuals*.
-These queries are part of Pearl's causal hierarchy (2009).Pearl matches these queries with their respective actions namely *observing*, *doing* and *imagining*.
-I compare the feasibility of addressing these queries and undertaking respective actions for both specifications.
-To contextualize SCMs within the field of causality, I also discuss the role of time in causal inference and provide a brief overview of the history of causal modelling.
-This paper uses various directed acyclic graphs to highlight the differences in these modelling approaches.
-The insights of this paper can be used as a baseline for subsequent research on structural causal models.
 ---
@@
 There are different schools of thought on how to optimally model causal relationships.
+While various fields have tried to provide parametric causal models, there are various challenges.
+One example is the consistency of causal models based on observational data.
+The consistency of observational data is more often than not violated, making the correct parametric estimation challenging. 
+Henceforth, models with non-parametric specification are gaining prominence within scholarshop.
 Structural Causal Models and Bayesian Causal Networks are two prominent causal models.
 Bayesian Causal Networks cast a model based on conditional probabilities.
+All relationships are defined in conditional probabilities, excluding exogenous variables.
+Because we ignore the interventional distribution, we cannot create counterfactuals based on bayesian causal networks.
 
-All relationships are defined in conditional probabilities, excluding exogenous variables.
 ```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
@@
 ```
+
 In Figure 1 we can see that an example of what a probabilistic model looks like.
@@
 
-Structural Causal Models (SCM) are a nonparametric modification of structural equations models.
+Compared to the BCN, the SCM specifies the intervention distribution for data outside of our observations and also specifies relationships based on equations.
+Structural Causal Models (SCM) are a nonparametric modification of structural equations models with some extensions.
 Structural equations models (SEM) are parametric causal models which are very popular in fields like economics, psychology and sociology.
-Here we differentiate between these two models because I focus on SCMs rather than SEMs.
-While both have a similar basis, they are some differences that I will further addressed in this essay.
 One should point out that there is a lot of controversy around SEMs because many scholars challenge the value of these parameters.
 As pointed out by Pearl (2012), SEMs have been deemed useless because these parametric specifications are never correct.
-The central idea of a SCM is to provide a model specifying an underlying causal process, without getting caught up in redundant parameter specifications that are never reflective of the true data generating process regardless of specification.
+The central idea of a SCM is to provide a model specifying an underlying causal process, without getting caught up in redundant parameter specifications.
+The SCM combines central ideas from the potential outcome framework, the structural equation model and graphical models that have developed a langauge to test some assumptions that are not testable in the SEM.
 Henceforth SCMs can be seen as heuristic versions of a SEM.
-SCMs use assignment equations to specify an underlying data-generation-process, or in short DGP.^[There is a lot of confusion around equations and what exactly they imply.
-Here we call them assignment equations rather than just equations to ensure that there is no misconception.
-I want to clarify that I mean equations as used in a computer science rather than an equation in mathematics.]
+
+SCMs use assignment equations to specify an underlying data-generation-process, or in short DGP.
+Note that these equations have not always had a concise mathematical notation. 
+Intially there was no sign to express the assignment equation and people used the '=' and would e.g. write A=B.
+But treating an equation as a algebraic equation, an equation that is symmetric makes no sense in the causality context because it would imply that B=A.
+Henceforth, the initial = sign was replaced with thw ':=' which is asymmetric (Pearl, 2009).
 
@@
 | Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
-| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           |SCM
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           |SCM ,PO
 

### 2021-05-03 intro.Rmd
xyz

@@
 One may argue that especially in high stake settings, algorithmic decision making based on co-occurrence is insufficient.
-Causal modelling has become a pivotal tool for interpretable modelling.
+This paper tries to clarify structural differences between association-based models and causal models. 
 There are different schools of thought on how to optimally model causal relationships.
-While various fields have tried to provide parametric causal models, there are various challenges.
-One example is the consistency of causal models based on observational data.
-The consistency of observational data is more often than not violated, making the correct parametric estimation challenging. 
-Henceforth, models with non-parametric specification are gaining prominence within scholarshop.
 Structural Causal Models and Bayesian Causal Networks are two prominent causal models.
-Bayesian Causal Networks cast a model based on conditional probabilities.
-All relationships are defined in conditional probabilities, excluding exogenous variables.
-Because we ignore the interventional distribution, we cannot create counterfactuals based on bayesian causal networks.
+Bayesian Causal Networks (BCN) cast a model based on conditional probabilities.
+All relationships are defined in conditional probabilities.
+BCNs ignore exogenous variables.
 
@@
 
-Compared to the BCN, the SCM specifies the intervention distribution for data outside of our observations and also specifies relationships based on equations.
-Structural Causal Models (SCM) are a nonparametric modification of structural equations models with some extensions.
+Compared to the BCN, the SCM specifies the intervention distribution for data outside of our observations and specifies relationships based on equations.
+Structural Causal Models (SCM) are a nonparametric modification of structural equations models with some extensions such as graphical methods and the implementation of features from the potential outcome framework.
 Structural equations models (SEM) are parametric causal models which are very popular in fields like economics, psychology and sociology.
@@
 The central idea of a SCM is to provide a model specifying an underlying causal process, without getting caught up in redundant parameter specifications.
+To accommodate the lack of data on exogenous variables, the field of causal inference introduced tools to combine various different sources of knowledge such as theoretical knowledge.
 The SCM combines central ideas from the potential outcome framework, the structural equation model and graphical models that have developed a langauge to test some assumptions that are not testable in the SEM.
 Henceforth SCMs can be seen as heuristic versions of a SEM.
-
 SCMs use assignment equations to specify an underlying data-generation-process, or in short DGP.
-Note that these equations have not always had a concise mathematical notation. 
-Intially there was no sign to express the assignment equation and people used the '=' and would e.g. write A=B.
-But treating an equation as a algebraic equation, an equation that is symmetric makes no sense in the causality context because it would imply that B=A.
-Henceforth, the initial = sign was replaced with thw ':=' which is asymmetric (Pearl, 2009).
+Note that these equations have not always had a concise mathematical notation (Pearl, 2009).
+Initially there was no sign to express the assignment equation and people used the '=' and one would e.g. write A=B.
+But treating an equation as a algebraic equation, an equation that is symmetric, makes no sense in the causality context.
+This would imply that B=A because the order has no concrete meaning in algebraic equations.
+Henceforth, the initial = sign was replaced with the ':=' which is asymmetric (Pearl, 2009).
+This misconception has caused a lot of challenges which I will adddress further on in this paper.
 
@@
 
-In figure 2 we can see a structural causal model. 
+In figure 2 we can see a vanilla-structural causal model. 
 The square nodes represent the latent variables
@@
 Due to the confined scope of this paper, I will not explore cyclic structures.]
+
+Beyond the school of thought, there is also a hierarchy of causation which Pearl introduced in 2009.
+Pearl focuses on three layers namely association, intervention and counterfactuals. 
+A higher level implies a deeper more detailed knowledge of the relationship between your variables.
+This hierarchy examines different queries, their actions and the respective hierachical rank.
+The first query is prediction where we examine based on observations.
+We can see that vanilla-machine learning methods, bayesian networks and classical regression are at the lowest level in the causal hierarchy.
 
@@
 \normalsize
-
-Beyond the school of thought, there is also a hierarchy of causation which Pearl introduced in 2009.
-Pearl focuses on three layers namely association, intervention and counterfactuals. 
-A higher level implies a deeper more detailed knowledge of the relationship between your variables.
-This hierarchy examines different queries, their actions and the respective hierachical rank.
-The first query is prediction where we examine based on observations.
-We can see that vanilla-machine learning methods, bayesian networks and classical regression are at the lowest level in the causal hierarchy.
 The second query is interventions, where we can use Pearls (2009) do-calculus to examine what happens if certain conditions holds.

### 2021-05-03 topic_1_firstdraft.Rmd
xyz

---
title: "A Gentle Introduction into Structural Causal Models"
author: 
- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
- Department of Statistics, Ludwig Maximilian University Munich, Germany
date: " April 5th, 2021"
output: 
  pdf_document:
        number_sections: yes
        fig_caption: yes
fontsize: 12
linestretch: 1.5
citation_package: --biblatex
bibliography: [ref.bib]
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Most undergraduate students that take a class in introductory statistics have heard the phrase 'correlation does not imply causation'.
While correlation implies co-occurrence, people are frequently interested in a causal understanding of relationships between variables.
One may argue that especially in high stake settings, algorithmic decision making based on co-occurrence is insufficient [@bareinboim2020pearl].
There are different approaches on how to model causal relationships.
Structural Causal Models and Bayesian Causal Networks are arguably the two most prominent causal models.
Bayesian Causal Networks (BCN) cast a model based on conditional probabilities.
All relationships are defined in **conditional probabilities.**
BCNs ignore exogenous variables [@pearl2009causality].
In Figure 1 we can see an example of what a probabilistic model looks like in a directed acylic graph.
Our nodes are the white circles.
The edges are the arrows, defined by the conditional probabilities.
All variables in our graph are observed variables. 
C is our **collider** variable because the affect of A on C and T on C collide [@pearl2009causality].

```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=7pt,
    node distance=3cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\node[my node] (A){A};
\node[my node,right =1 cm of A](C){C};
\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
\draw[->] (A) -- (T);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

The field of causal inference introduced tools to combine various different sources of knowledge such as theoretical knowledge [@morgan_winship_2014] for more realistic models beyond the confinement of our observations.
Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality] including unobserved variables.
SCMs are a nonparametric modification of structural equations models (a set of functional equations) with some extensions.
These extensions are a combination of the mathematical language from graphical models and the implementation of the potential outcome framework [@pearl2012causal].
Structural equations models (SEM) are parametric causal models which are very popular in fields like economics, psychology and sociology (Pearl 2009).
One should point out that there is a lot of controversy around SEMs because many scholars challenge the value of the parameters in a SEM [@pearl2012causal].
The central idea of SCMs is to provide a model specifying an underlying data generating process, without the redundancy of creating inconsistent parameter specifications.^[Inconsistency refers to inherent nature of creating parameters based of observational data. One inherent feature of observational data is that it is seldom consistent and henceforth respective estimates are questionable.For further information see @hernan2008does] 
SCMs use assignment equations to specify an underlying data-generation-process.
Note that these equations have not always had a concise mathematical notation (Pearl, 2009).
Initially there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
But treating an equation as a **algebraic equation**, makes no sense in the causality context because this equation has no causal information.
This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
The problem is that the equation is symmetric.
The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
This misconception has caused a lot of challenges which I will address further on in this paper. 
As mentioned, we define variables as functions therefore a variable for example $A=f_A(B,U_A)$.
A is defined by B and the latent factor $U_A$.
To summarize, a SCM consists of a set of (autonomous) equations to generate (a) endogenous variables and (b) exogenous variables.
In figure 2 we can see a plain-vanilla structural causal model. 
The square nodes represent the latent variables.
The circle nodes represent the observed variables.
The arrows are our structural equations and depict the relationship between variables.
^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks. For further information see Pearl (2009). Due to the confined scope of this paper, I will not explore cyclic structures.]

```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left of=A](UA){$U_A$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,left = 1cm of T](UT){$U_T$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (T);
\draw[->] (UA) -- (A);
\draw[->] (UT) -- (T);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

Beyond the different causal model specifications, Judea Pearl (2009) introduced the hierarchy of causation.
Pearl focuses on three layers namely association, intervention and counterfactuals. 
A higher level implies a more detailed knowledge of the relationship between the variables.
The first query is association, where we examine relationships based on observations.

\footnotesize
 |Method          | Action |  Example | Usage | 
|------------------|-------------|--------------------|-------------------|
| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO

Table: Pearls Hierachy of Causation (2009)

\normalsize

In table 1 we can see that machine learning (ML) methods, bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy.
These methods depend on association alone and lack information on what happens if we make changes in our root nodes (our nodes that are not caused by other variables) and what would happen if we had a different setting. 
The second query deals with interventions.
Here we can use Pearls (2009) do-calculus to examine what happens if root nodes are manipulated.
Note that there are various different types of intervention, such as atomic intervention, where we set a variable to a constant, or e.g. policy intervention where we specify a different function for a specific equation in our set of equations.
Additionally there is also off-policy intervention, where we model a different intervention that is not in our historical data based on the historical data [@oberst2019counterfactual].
Causal bayesian networks , markov decision processes (MDP) and reinforcement learning frequent these intervention methods.
The third query is counterfactual modelling. 
Here we deal with hypothetical settings. 
SCMs and potential outcome models allow for counterfactual modelling.
The reason why these models can model counterfactuals is because they also include information on the interventional distribution [@oberst2019counterfactual].
Given that e.g. BCN only entail conditional probabilities and no information on relationships outside of the observations in the data, they cannot create counterfactuals (Pearl 2009).

Another issue in this paper is understanding the relationship of time and causal modelling.
Causal models mostly disregard concise notions of time and make the strong assumption that relationships between variables hold beyond the confinement of time (Peters et al. 2017).
These vague definitions of time is more prevalent in social sciences. 
Meanwhile hard sciences deal with time in a more concise manner. 
Some research has also looked at causal models with a more concise specification of time using differential equations (Mooij et al. 2013).

To summarize, existing literature has provided an excellent introduction to structural causal models and bayesian causal networks. 
Judea Pearl (2009) provided a comprehensive and accessible introduction into this topic with his book on causality.
His work has addressed various misconceptions in social sciences, economics, causality and statistics. 
Peters et al. (2017) added to this discourse by addressing the relationship of causal modelling to physical sciences and causal inference for observational data.
Additionally, other scholars e.g. [@bareinboim2020pearl] have worked on the application of causal methods in machine learning. 
@tarka2018overview discuss the history of SEMs and SCMs.
This paper brings together these contributions. 
The objective of this paper is to provide a comprehensive summary of structural causal models and respective applications in social sciences, physical sciences and machine learning.
The insights in this paper are an abstraction of these pieces of work, offering the reading a basic understanding of the topic for future research and an overview of the current research.
The rest of the paper is structured as follows: 
The upcoming section studies the hierarchy of causation, looking at the different queries in the SCM and the BCN and what implications the functional/conditional probability specification carry.
The third section provides further reference to the intersection of machine learning and structural causal models. 
The fourth section studies historical differences and similarities between SCMs and causal models (SEM,PO) that emerged in the social sciences. 
The fifth section examines the relationship of physical sciences and SCMs, addressing the issue of time and modelling based on differential equations.

# References

### 2021-05-06 topic_1_firstdraft.Rmd
next chapters

@@
 nocite: '@*'
+abtract: The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling.
+To provide a comprehensive understanding of causal modelling, I introduce two prominent causal model specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM), focusing on the latter.Probabilistic specifications such as a BCN cast a model based on conditional probabilities.
+SCMs cast a model based on assignment functions and extend probabilistic models by specifying the data generating process rather than solely utilizing conditional probabilities.
+Another difference between these models is their ability to address different queries such as *predictions*, *interventions* and *counterfactuals*.
+These queries are part of Pearl's causal hierarchy (2009).
+Pearl matches these queries with their respective actions namely *observing*, *doing* and *imagining*.
+I compare the feasibility of addressing these queries and undertaking respective actions for both specifications.
+To contextualize SCMs within the field of causality, I also discuss the role of time in causal inference and provide a brief overview of the history of causal modelling.
+This paper uses various directed acyclic graphs to highlight the differences in these modelling approaches.
+The insights of this paper can be used as a baseline for subsequent research on structural causal models.
+
 ---
@@
 
+# The Causal Hierachy Compared
+
+ |Method         | CBN |  SCM
+|----------------|---------------------------|------------------------|
+| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
+|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
+|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
+|                |                                                                                                                      | 
+| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
+|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
+|                |                                                                                                                      | 
+|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
+|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+
+**Counterfactuals**
+
+Process is described as follows:
+
+(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
+(b) Action: Exchange $(X = x)$ 
+(c) Prediction: Compute $(Y = y)$
+
+# Causal Modelling and Machine Learning
+
+# Causal Models and Social Sciences
+
+Judea Pearl argues that path analysis, a method proposed by Sewall Wright in the 1920s is the foundation of modern structural causal models.
+Sewall Wright, a Statistican and Geneticist, sparked the interest in causal modelling with his path analysis which subsequently was adopted by various other disciplines such as econometrics, psychometrics and sociology.
+This path analysis is a structural equation model with one variable per indicator.
+
+Thereafter the Norwegian Economist Trygve Haavelmo (1944) provided a modelling framework for the field of economics in is paper 'The Probability Approach in Econometrics.'
+One could argue that this paper provided a new of level of sophistication to the discipline of Econometrics as a distinct field. 
+Haavelmo was awarded the Nobel Memorial Prize in Economic Science in 1989 for his work on this issue. 
+His work focused on
+
+# Causal Models and Physical Sciences
+
+# Conclusion
+
 # References

### 2021-05-07 topic_1_firstdraft.Rmd
restructure

@@
 The rest of the paper is structured as follows:
-The upcoming section studies the hierarchy of causation, looking at the different queries in the SCM and the BCN and what implications the functional/conditional probability specification carry.
-The third section provides further reference to the intersection of machine learning and structural causal models. 
-The fourth section studies historical differences and similarities between SCMs and causal models (SEM,PO) that emerged in the social sciences. 
-The fifth section examines the relationship of physical sciences and SCMs, addressing the issue of time and modelling based on differential equations.
 
-# The Causal Hierachy Compared
+# Foundations of Structural Causal Models
+
+Judea Pearl argues that path analysis, a method proposed by Sewall Wright in the 1920s is the foundation of modern structural causal models.
+Sewall Wright, a Statistican and Geneticist, sparked the interest in causal modelling with his path analysis which subsequently was adopted by various other disciplines such as econometrics, psychometrics and sociology.
+This path analysis is a structural equation model with one variable per indicator.
+
+Thereafter the Norwegian Economist Trygve Haavelmo (1944) provided a modelling framework for the field of economics in is paper 'The Probability Approach in Econometrics.'
+One could argue that this paper provided a new of level of sophistication to the discipline of Econometrics as a distinct field. 
+Haavelmo was awarded the Nobel Memorial Prize in Economic Science in 1989 for his work on this issue.
+
+## Assumptions
+
+Independence
+
+## Pearl Causal Hierachy
 
@@
 
-# Causal Modelling and Machine Learning
+## Causal Inference and Time
 
-# Causal Models and Social Sciences
-
-Judea Pearl argues that path analysis, a method proposed by Sewall Wright in the 1920s is the foundation of modern structural causal models.
-Sewall Wright, a Statistican and Geneticist, sparked the interest in causal modelling with his path analysis which subsequently was adopted by various other disciplines such as econometrics, psychometrics and sociology.
-This path analysis is a structural equation model with one variable per indicator.
-
-Thereafter the Norwegian Economist Trygve Haavelmo (1944) provided a modelling framework for the field of economics in is paper 'The Probability Approach in Econometrics.'
-One could argue that this paper provided a new of level of sophistication to the discipline of Econometrics as a distinct field. 
-Haavelmo was awarded the Nobel Memorial Prize in Economic Science in 1989 for his work on this issue. 
-His work focused on
-
-# Causal Models and Physical Sciences
+## Graphical Models
 

### 2021-05-07 topic_1_firstdraft.Rmd
topic 1 table

@@
 
+Aside from path diagrams also introduced graph rules for writing down relationships.
+
 Thereafter the Norwegian Economist Trygve Haavelmo (1944) provided a modelling framework for the field of economics in is paper 'The Probability Approach in Econometrics.'
@@
 
+Causal assumptions are encoded in missing links.
+
 ## Assumptions
 
-Independence
+Causal assumptions differentiate causal models from association learning methods.
+
+ | Association-based Concepts |  Causal Concepts
+|--------------------------|------------------------|
+| Correlation              | Randomization  
+| Regression               | Confounding
+| Conditional Independence | Disturbance
+| Likelihood               | Error Terms      
+| Odds Ratio               | Structural Coefficients       
+| Propensity Score         | Spurious Correlation
+
+Table: Concepts in Causality and Association concepts
+
+These causal concepts are not expressible based on distribution functions/statistical associations. (Pearl 2010)
+
+Disturbance in SCM: Correlated and causal factor ; responsible for variation
+Disturbance in Regression: Uncorrelated
+
+Causal assumption not testable (e.g. $Cov(U_a, U_b)=0$).
+d-seperation to test assumptions in totality (cannot make assumptions in isolation).
+
+exploit invariant characteristics of SEM whithout committing to shape. 
+structural if function autonomous and invariant to change in form of other functions.
 

### 2021-05-08 topic_1_firstdraft.Rmd
random conclusion

@@
 
+This paper provides a gentle introduction into structural causal models.
+SCMs entail many features, complementing research on association learning by providing depth. 
+This in turn, is of particular benefit for high stake decision settings. 
+SCMs differentiate from other methods through the specification of endogenous and exogenous variables, treating the exogenous factors as pivotal components of the actual model rather than assuming they are ommitable errors that are uncorrelated.
+As suggested by Pearls Hierachy, there are different levels to learning and each higher step can do anything the prior step can but with more detail and information. 
+Henceforth, as machin
+
 # References

### 2021-05-09 references/vanilla_presentation .Rmd
presentation slides

@@
 ---
-title: "Fairness in Machine Learning WS20/21"
-subtitle: "Fostering Notions of Dynamic Fairness via Structural Causal Models"
+title: "Fairness in Machine Learning SS21"
+subtitle: "Beyond Static Fairness"
 date: "03/05/2021"
@@
 
+# Dynamical Repercussions
+
+- Apple card discriminating women
+
 # Introduction
@@
 - Prior research on causal fairness has focused in static systems
-- Recently Creager et al. (2020) proposed a structural causal model for dynamical systems
+
+# Fairness-Gym
+
+- Environment to simulate fairness
 
@@
 - Judea Pearl:do-calculus
--
 
-# SCM vs Causal Bayesian Network (Probabilistic Model)
+# Graphical Illustration
+
+```{tikz,fig.align="center",fig.cap="Probabilistic Model", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=7pt,
+    node distance=3cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\node[my node] (A){A};
+\node[my node,right =1 cm of A](C){C};
+\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
+\draw[->] (A) -- (T);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
+
+```{tikz,fig.cap="Structural Causal Model", fig.align="center", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+  \begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=6pt,
+    node distance=2cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
+\node[my node, fill=gray!30] (A){A};
+\node[latent,left of=A](UA){$U_A$};
+\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
+\node[my node,right = 1 cm of T, fill=gray!30](C){C};
+\node[latent,left = 1cm of T](UT){$U_T$};
+\node[latent,right = 1cm of A](UC){$U_C$};
+\draw[->] (A) -- (T);
+\draw[->] (UA) -- (A);
+\draw[->] (UT) -- (T);
+\draw[->] (UC) -- (C);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
+
+# Interventions
+
+- Atomic Intervention 
+- Policy Intervention 
+- Off-Policy Intervention (model-based; model-free)
+
+# Results Off Policy Intervention - Lambda
+
+* Introduce trade off parameter lambda 
+* V of Pi is the overall objective, Pi is our policy and U is utility
+
+$$V_\pi =U - \lambda\delta_{EQOPP}$$
+
+# Causal Hierachy
+
+ |Method          | Action |  Example | Usage | 
+|------------------|-------------|--------------------|-------------------|
+| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
+| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
+
+Table: Pearls Hierachy of Causation (2009)
 
@@
 - Off-policy estimation (model based (regression) or model free estimation (propensity weight))
-
-# Time and Causal Inference
-
-- Structural causal models are abstractions of physical processes
--
 
@@
 - Semi-Deterministic SCMs (deterministic -> all variables are known)
+-
 

### 2021-05-11 references/vanilla_presentation .Rmd
presentation baseline

@@
 
-- Apple card discriminating women
-
-# Introduction
+- What to do once we learned a certain statistical model, we implement an action
+- Workflow: Typically we create fairness algorithms such as equal opportunity parities or demographic parities to evaluate classifiers
+- We can evaluate these methods via simulation studies
+- fairness for dynamical systems is nothing new but recently study pointed out that simulations are great tool for this setting
+- This approach models fairness as dynamical system with agents and environments
 
 - Fairness in dynamical system has become important because equalizing true positive rate at each step does not converge as fast in systems with e.g. population dynamics 
+- Widdening gap between two policies 
 - Research on dynamical systems has focused on markov decision processes

### 2021-05-11 scm.Rmd
presentation baseline

--- 2021-05-07 topic_1_firstdraft.Rmd
@@
 ---
-title: "A Gentle Introduction into Structural Causal Models"
-author: 
-- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
-- Department of Statistics, Ludwig Maximilian University Munich, Germany
-date: " April 5th, 2021"
-output: 
-  pdf_document:
-        number_sections: yes
-        fig_caption: yes
-fontsize: 12
-linestretch: 1.5
-citation_package: --biblatex
-bibliography: [ref.bib]
-nocite: '@*'
-abtract: The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling.
-To provide a comprehensive understanding of causal modelling, I introduce two prominent causal model specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM), focusing on the latter.Probabilistic specifications such as a BCN cast a model based on conditional probabilities.
-SCMs cast a model based on assignment functions and extend probabilistic models by specifying the data generating process rather than solely utilizing conditional probabilities.
-Another difference between these models is their ability to address different queries such as *predictions*, *interventions* and *counterfactuals*.
-These queries are part of Pearl's causal hierarchy (2009).
-Pearl matches these queries with their respective actions namely *observing*, *doing* and *imagining*.
-I compare the feasibility of addressing these queries and undertaking respective actions for both specifications.
-To contextualize SCMs within the field of causality, I also discuss the role of time in causal inference and provide a brief overview of the history of causal modelling.
-This paper uses various directed acyclic graphs to highlight the differences in these modelling approaches.
-The insights of this paper can be used as a baseline for subsequent research on structural causal models.
-
+title: "scm_presentation"
+author: "Daniel Saggau"
+date: "5/11/2021"
+output: beamer_presentation
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
@@
 
-Most undergraduate students that take a class in introductory statistics have heard the phrase 'correlation does not imply causation'.
-While correlation implies co-occurrence, people are frequently interested in a causal understanding of relationships between variables.
-One may argue that especially in high stake settings, algorithmic decision making based on co-occurrence is insufficient [@bareinboim2020pearl].
-There are different approaches on how to model causal relationships.
-Structural Causal Models and Bayesian Causal Networks are arguably the two most prominent causal models.
-Bayesian Causal Networks (BCN) cast a model based on conditional probabilities.
-All relationships are defined in **conditional probabilities.**
-BCNs ignore exogenous variables [@pearl2009causality].
-In Figure 1 we can see an example of what a probabilistic model looks like in a directed acylic graph.
-Our nodes are the white circles.
-The edges are the arrows, defined by the conditional probabilities.
-All variables in our graph are observed variables. 
-C is our **collider** variable because the affect of A on C and T on C collide [@pearl2009causality].
+Causal assumptions differentiate causal models from association learning methods.
 
-```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
+ | Association-based Concepts |  Causal Concepts
+|--------------------------|------------------------|
+| Correlation              | Randomization  
+| Regression               | Confounding
+| Conditional Independence | Disturbance
+| Likelihood               | Error Terms      
+| Odds Ratio               | Structural Coefficients       
+| Propensity Score         | Spurious Correlation
+
+# Fundamental Differences to begin
+
+SCM:
+
+- is a nonparametric SEM 
+- has functional form rather than using probabilities
+- entails features from the PO framework and graphical representation 
+- conflict whether to use graphs or not
+
+* A SEM is a parametric specification used in applied sciences (parameters contested)
+* A Bayesian causal network is another popular causal model using conditional probabilities and NO functions 
+* Differences in performance between BCN and SCM
+
+#
+
+ |Method         | CBN |  SCM
+|----------------|---------------------------|------------------------|
+| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
+|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
+|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
+|                |                                                                                                                      | 
+| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
+|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
+|                |                                                                                                                      | 
+|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
+|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+
+# Graphical Illustration
+
+```{tikz,fig.align="center",fig.cap="Probabilistic Model", echo =F}
 \usetikzlibrary{calc}
@@
 
-The field of causal inference introduced tools to combine various different sources of knowledge such as theoretical knowledge [@morgan_winship_2014] for more realistic models beyond the confinement of our observations.
-Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality] including unobserved variables.
-SCMs are a nonparametric modification of structural equations models (a set of functional equations) with some extensions.
-These extensions are a combination of the mathematical language from graphical models and the implementation of the potential outcome framework [@pearl2012causal].
-Structural equations models (SEM) are parametric causal models which are very popular in fields like economics, psychology and sociology (Pearl 2009).
-One should point out that there is a lot of controversy around SEMs because many scholars challenge the value of the parameters in a SEM [@pearl2012causal].
-The central idea of SCMs is to provide a model specifying an underlying data generating process, without the redundancy of creating inconsistent parameter specifications.^[Inconsistency refers to inherent nature of creating parameters based of observational data.
-One inherent feature of observational data is that it is seldom consistent and henceforth respective estimates are questionable.For further information see @hernan2008does]
-SCMs use assignment equations to specify an underlying data-generation-process.
-Note that these equations have not always had a concise mathematical notation (Pearl, 2009).
-Initially there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
-But treating an equation as a **algebraic equation**, makes no sense in the causality context because this equation has no causal information.
-This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
-The problem is that the equation is symmetric.
-The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
-This misconception has caused a lot of challenges which I will address further on in this paper. 
-As mentioned, we define variables as functions therefore a variable for example $A=f_A(B,U_A)$.
-A is defined by B and the latent factor $U_A$.
-To summarize, a SCM consists of a set of (autonomous) equations to generate (a) endogenous variables and (b) exogenous variables.
-In figure 2 we can see a plain-vanilla structural causal model. 
-The square nodes represent the latent variables.
-The circle nodes represent the observed variables.
-The arrows are our structural equations and depict the relationship between variables.
-^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks.
-For further information see Pearl (2009).
-Due to the confined scope of this paper, I will not explore cyclic structures.]
-
-```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
+```{tikz,fig.cap="Structural Causal Model", fig.align="center", echo =F}
 \usetikzlibrary{calc}
@@
 
-Beyond the different causal model specifications, Judea Pearl (2009) introduced the hierarchy of causation.
-Pearl focuses on three layers namely association, intervention and counterfactuals. 
-A higher level implies a more detailed knowledge of the relationship between the variables.
-The first query is association, where we examine relationships based on observations.
+# Pearls Causal Hierachy
 
@@
 \normalsize
-
-In table 1 we can see that machine learning (ML) methods, bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy.
-These methods depend on association alone and lack information on what happens if we make changes in our root nodes (our nodes that are not caused by other variables) and what would happen if we had a different setting.
-The second query deals with interventions.
-Here we can use Pearls (2009) do-calculus to examine what happens if root nodes are manipulated.
-Note that there are various different types of intervention, such as atomic intervention, where we set a variable to a constant, or e.g.
-policy intervention where we specify a different function for a specific equation in our set of equations.
-Additionally there is also off-policy intervention, where we model a different intervention that is not in our historical data based on the historical data [@oberst2019counterfactual].
-Causal bayesian networks , markov decision processes (MDP) and reinforcement learning frequent these intervention methods.
-The third query is counterfactual modelling. 
-Here we deal with hypothetical settings. 
-SCMs and potential outcome models allow for counterfactual modelling.
-The reason why these models can model counterfactuals is because they also include information on the interventional distribution [@oberst2019counterfactual].
-Given that e.g. BCN only entail conditional probabilities and no information on relationships outside of the observations in the data, they cannot create counterfactuals (Pearl 2009).
-
-Another issue in this paper is understanding the relationship of time and causal modelling.
-Causal models mostly disregard concise notions of time and make the strong assumption that relationships between variables hold beyond the confinement of time (Peters et al. 2017).
-These vague definitions of time is more prevalent in social sciences. 
-Meanwhile hard sciences deal with time in a more concise manner. 
-Some research has also looked at causal models with a more concise specification of time using differential equations (Mooij et al. 2013).
-
-To summarize, existing literature has provided an excellent introduction to structural causal models and bayesian causal networks. 
-Judea Pearl (2009) provided a comprehensive and accessible introduction into this topic with his book on causality.
-His work has addressed various misconceptions in social sciences, economics, causality and statistics. 
-Peters et al. (2017) added to this discourse by addressing the relationship of causal modelling to physical sciences and causal inference for observational data.
-Additionally, other scholars e.g. [@bareinboim2020pearl] have worked on the application of causal methods in machine learning. 
-@tarka2018overview discuss the history of SEMs and SCMs.
-This paper brings together these contributions. 
-The objective of this paper is to provide a comprehensive summary of structural causal models and respective applications in social sciences, physical sciences and machine learning.
-The insights in this paper are an abstraction of these pieces of work, offering the reading a basic understanding of the topic for future research and an overview of the current research.
-The rest of the paper is structured as follows:
-
-# Foundations of Structural Causal Models
-
-Judea Pearl argues that path analysis, a method proposed by Sewall Wright in the 1920s is the foundation of modern structural causal models.
-Sewall Wright, a Statistican and Geneticist, sparked the interest in causal modelling with his path analysis which subsequently was adopted by various other disciplines such as econometrics, psychometrics and sociology.
-This path analysis is a structural equation model with one variable per indicator.
-
-Thereafter the Norwegian Economist Trygve Haavelmo (1944) provided a modelling framework for the field of economics in is paper 'The Probability Approach in Econometrics.'
-One could argue that this paper provided a new of level of sophistication to the discipline of Econometrics as a distinct field. 
-Haavelmo was awarded the Nobel Memorial Prize in Economic Science in 1989 for his work on this issue.
-
-## Assumptions
-
-Independence
-
-## Pearl Causal Hierachy
-
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
-
-**Counterfactuals**
-
-Process is described as follows:
-
-(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
-(b) Action: Exchange $(X = x)$ 
-(c) Prediction: Compute $(Y = y)$
-
-## Causal Inference and Time
-
-## Graphical Models
-
-# Conclusion
-
-# References

### 2021-05-11 scm.Rmd
scm

@@
 ---
-title: "scm_presentation"
+title: "A Gentle Introduction to Structural Causal Models"
 author: "Daniel Saggau"
@@
 
-# Introduction
+# Introduction (1)
+
+- Machine learning has provided many insights into different research problems 
+- One issue is the consideration of 'What are we actually predicting'?
+- Mainstream tools are build on association-based learning 
+- Associations are not enough for high stake settings 
+- In disciplines like psychology or economics people are less interested in associational learning
+- We want causation and not correlation
+
+# Introduction (2)
 
@@
 
-# Fundamental Differences to begin
+# History
 
-SCM:
+* Path Analysis -> SEM -> SCM
+*
 
+# How to build a SCM
+
+- Consists of system of equations 
+- Assignment equation ':=' rather than regular equation '=' 
 - is a nonparametric SEM 
@@
 - entails features from the PO framework and graphical representation 
-- conflict whether to use graphs or not
+- Exogenous factors are part of the model specification
 
-* A SEM is a parametric specification used in applied sciences (parameters contested)
-* A Bayesian causal network is another popular causal model using conditional probabilities and NO functions 
-* Differences in performance between BCN and SCM
+# SCM Applications:
 
-#
-
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+- Flexible simulations for higher order problems (intervention, counterfactual)
+- Graphical visualization via directed acyclic graph
+-
 
@@
 
+# Fundamental Differences (1)
+
+- conflict whether to use graphs or not 
+* A SEM is a parametric specification used in applied sciences (parameters contested)
+* A Bayesian causal network is another popular causal model using conditional probabilities and NO functions 
+* Differences in performance between BCN and SCM
+
+# Performance Evaluation
+
+ |Method         | CBN |  SCM
+|----------------|---------------------------|------------------------|
+| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
+|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
+|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
+|                |                                                                                                                      | 
+| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
+|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
+|                |                                                                                                                      | 
+|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
+|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+
+# Assumptions
+
+**Error terms**
+
+Regression: Omittable outside factor 
+SCM/SEM; Latent influencial factor that is pivotal for the model specification but not observable
+
 # Pearls Causal Hierachy
@@
 \normalsize
+
+# Causality -In Between Physical and Social Sciences
+
+# Causality and Time
+
+- Time in Physical Sciences: Mechanical and exact 
+- Time in Social Sciences: Often Vague 
+- Regular Time Specification is also more vague 
+- To accomodate that issue, research on differential equation based SCMs started

### 2021-05-13 references/vanilla_presentation .Rmd
yellow

--- 2021-05-09 references/vanilla_presentation .Rmd
@@
     theme: Madrid
-    colortheme: spruce
+    colortheme: crane
 ---

### 2021-05-13 topic_1_firstdraft.Rmd
topic1
presentation slides

@@
 In table 1 we can see that machine learning (ML) methods, bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy.
-These methods depend on association alone and lack information on what happens if we make changes in our root nodes (our nodes that are not caused by other variables) and what would happen if we had a different setting.
+These methods depend on association alone.
+Associational methods neglect information on changes in our root nodes. 
+Root nodes are nodes that are not caused by any other variable.
 The second query deals with interventions.
-Here we can use Pearls (2009) do-calculus to examine what happens if root nodes are manipulated.
-Note that there are various different types of intervention, such as atomic intervention, where we set a variable to a constant, or e.g.
-policy intervention where we specify a different function for a specific equation in our set of equations.
+Here we can use Pearls (2009) do-calculus.
+The do-calculus enables us to study the manipulation of parent nodes.
+ There are various types of intervention.
+One example is **atomic intervention,** where we set a variable to a constant.
+In **policy intervention** we specify a different function for an equation.
+
 Additionally there is also off-policy intervention, where we model a different intervention that is not in our historical data based on the historical data [@oberst2019counterfactual].
@@
 
+### Causal Inference on Observational Data
+
+Select Admissable set based on backdoor criterion.
+Backdoor criterion missing in PO framework where subset is selected based on mental reasoning.
+Conditional Ignorability is mirror of backdoor criterion in PO-Framework.
+
 ## Pearl Causal Hierachy

### 2021-05-14 topic_1_firstdraft.Rmd
xyz

@@
 
-Most undergraduate students that take a class in introductory statistics have heard the phrase 'correlation does not imply causation'.
-While correlation implies co-occurrence, people are frequently interested in a causal understanding of relationships between variables.
-One may argue that especially in high stake settings, algorithmic decision making based on co-occurrence is insufficient [@bareinboim2020pearl].
+Most students have heard the phrase 'correlation does not imply causation.'
+While correlation implies co-occurrence, for many problems correlation is not enough. 
+Algorithmic decision making based on co-occurrence is insufficient  in high stake settings [@bareinboim2020pearl].
+For many problems we want to understand causal relationships between variables.
 There are different approaches on how to model causal relationships.
-Structural Causal Models and Bayesian Causal Networks are arguably the two most prominent causal models.
-Bayesian Causal Networks (BCN) cast a model based on conditional probabilities.
-All relationships are defined in **conditional probabilities.**
-BCNs ignore exogenous variables [@pearl2009causality].
-In Figure 1 we can see an example of what a probabilistic model looks like in a directed acylic graph.
-Our nodes are the white circles.
+This paper focuses on Structural Causal Models and Bayesian Causal Networks.
+Bayesian Causal Networks (BCN) cast a model based on **conditional probabilities.**
+In Figure 1 we can see an example of a probabilistic model depicted  as a directed acyclic graph (DAG).
+The nodes are the white circles.
 The edges are the arrows, defined by the conditional probabilities.
-All variables in our graph are observed variables. 
 C is our **collider** variable because the affect of A on C and T on C collide [@pearl2009causality].
@@
 
-The field of causal inference introduced tools to combine various different sources of knowledge such as theoretical knowledge [@morgan_winship_2014] for more realistic models beyond the confinement of our observations.
-Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality] including unobserved variables.
-SCMs are a nonparametric modification of structural equations models (a set of functional equations) with some extensions.
-These extensions are a combination of the mathematical language from graphical models and the implementation of the potential outcome framework [@pearl2012causal].
-Structural equations models (SEM) are parametric causal models which are very popular in fields like economics, psychology and sociology (Pearl 2009).
-One should point out that there is a lot of controversy around SEMs because many scholars challenge the value of the parameters in a SEM [@pearl2012causal].
-The central idea of SCMs is to provide a model specifying an underlying data generating process, without the redundancy of creating inconsistent parameter specifications.^[Inconsistency refers to inherent nature of creating parameters based of observational data.
-One inherent feature of observational data is that it is seldom consistent and henceforth respective estimates are questionable.For further information see @hernan2008does]
-SCMs use assignment equations to specify an underlying data-generation-process.
-Note that these equations have not always had a concise mathematical notation (Pearl, 2009).
-Initially there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
-But treating an equation as a **algebraic equation**, makes no sense in the causality context because this equation has no causal information.
+Causal inference introduced tools that combine combine different sources of knowledge. 
+We can e.g. include theoretical knowledge [@morgan_winship_2014] and observational knowledge with structural causal models.
+Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality].
+SCMs are non-parametric structural equations models with added features.  
+A structural equation model (SEM) is a set of autonous equations.
+This set represents the 'state of the world'. 
+SEMs are popular in fields like economics, psychology and sociology (Pearl 2009).
+SCM are also defined by a (sub-)set of structural equations.
+Additionally, SCMs  feature mathematical components from graphical models and the potential-outcome framework [@pearl2012causal].
+There is a lot of controversy around SEMs.
+Many scholars challenge the parameter-specification in a SEM [@pearl2012causal].
+^[Inconsistency refers to inherent nature of creating parameters based of observational data.
+Observational data  is seldom consistent.
+Henceforth respective estimates are often questioned.
+For further information see @hernan2008does]
+SCMs specify an underlying data generating process without the computational effort of creating a parametric model.
+Note, that structural equations refer to assignment equations, used in computer science. 
+Equations in causal models did not always have a concise notation (Pearl, 2009).
+First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
+Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
 This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
@@
 This misconception has caused a lot of challenges which I will address further on in this paper. 
-As mentioned, we define variables as functions therefore a variable for example $A=f_A(B,U_A)$.
-A is defined by B and the latent factor $U_A$.
+As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
+B defines A and the latent factor $U_A$.
 To summarize, a SCM consists of a set of (autonomous) equations to generate (a) endogenous variables and (b) exogenous variables.
-In figure 2 we can see a plain-vanilla structural causal model. 
+In figure 2 we can see a vanilla structural causal model. 
 The square nodes represent the latent variables.
@@
 
-Beyond the different causal model specifications, Judea Pearl (2009) introduced the hierarchy of causation.
-Pearl focuses on three layers namely association, intervention and counterfactuals. 
-A higher level implies a more detailed knowledge of the relationship between the variables.
+To contextualise causal methods, Judea Pearl (2009) introduced the hierarchy of causation.
+Pearl focuses on three layers: association, intervention and counterfactuals. 
+A higher level implies more detailed knowledge of the relationship between the variables.
 The first query is association, where we examine relationships based on observations.
@@
 
-In table 1 we can see that machine learning (ML) methods, bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy.
-These methods depend on association alone.
-Associational methods neglect information on changes in our root nodes. 
-Root nodes are nodes that are not caused by any other variable.
+Vanilla machine learning (ML), bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy (see table 1).
+These methods demand the least information and depend on association alone.
+Associational methods ignore external changes outside of our data.
+The interventional distribution has information on these external changes.
+The interventional distribution is only defined in high level causal methods. 
 The second query deals with interventions.
@@
 In **policy intervention** we specify a different function for an equation.
+off-policy intervention models different intervention that is not in our historical data [@oberst2019counterfactual].
+Causal bayesian networks , Markov Decision Processes (MDP) and reinforcement learning model intervention.
 
-Additionally there is also off-policy intervention, where we model a different intervention that is not in our historical data based on the historical data [@oberst2019counterfactual].
-Causal bayesian networks , markov decision processes (MDP) and reinforcement learning frequent these intervention methods.
 The third query is counterfactual modelling. 
@@
 SCMs and potential outcome models allow for counterfactual modelling.
-The reason why these models can model counterfactuals is because they also include information on the interventional distribution [@oberst2019counterfactual].
-Given that e.g. BCN only entail conditional probabilities and no information on relationships outside of the observations in the data, they cannot create counterfactuals (Pearl 2009).
+These models can model counterfactuals because they include a interventional distribution [@oberst2019counterfactual].
+BCNs only entail conditional probabilities.
+There is no information on relationships outside of the observations in the data.
+Henceforth, they cannot create counterfactuals (Pearl 2009).
 
-Another issue in this paper is understanding the relationship of time and causal modelling.
-Causal models mostly disregard concise notions of time and make the strong assumption that relationships between variables hold beyond the confinement of time (Peters et al. 2017).
+Another issue in this paper is understanding the relationship of time and causality.
+SCMs make the assumption that causal relationships hold over time (Peters et al. 2017).
 These vague definitions of time is more prevalent in social sciences. 
 Meanwhile hard sciences deal with time in a more concise manner. 
-Some research has also looked at causal models with a more concise specification of time using differential equations (Mooij et al. 2013).
+Differential equations model time on a more mechanic manner (Mooij et al. 2013).
 
-To summarize, existing literature has provided an excellent introduction to structural causal models and bayesian causal networks. 
-Judea Pearl (2009) provided a comprehensive and accessible introduction into this topic with his book on causality.
-His work has addressed various misconceptions in social sciences, economics, causality and statistics. 
-Peters et al. (2017) added to this discourse by addressing the relationship of causal modelling to physical sciences and causal inference for observational data.
-Additionally, other scholars e.g. [@bareinboim2020pearl] have worked on the application of causal methods in machine learning. 
+Existing literature has provided an excellent introduction to SCMs and BCNs. 
+Pearl (2009) provided a comprehensive introduction into this topic with his book 'Causality'.
+His work addressed misconceptions in social sciences, causality and statistics. 
+Peters et al. (2017) address the relationship of causality, physical sciences, and causal inference.
+They also look at causal inference for observational data.
+E.g. [@bareinboim2020pearl] worked on the application of causal methods in machine learning. 
 @tarka2018overview discuss the history of SEMs and SCMs.
-This paper brings together these contributions. 
-The objective of this paper is to provide a comprehensive summary of structural causal models and respective applications in social sciences, physical sciences and machine learning.
-The insights in this paper are an abstraction of these pieces of work, offering the reading a basic understanding of the topic for future research and an overview of the current research.
-The rest of the paper is structured as follows:
+This paper combines these contributions. 
+The aim of this paper is to summarise SCMs and it's intersection with social sciences, physics, and machine learning.
+I structure the rest of the paper as follows:
 
@@
 
-Judea Pearl argues that path analysis, a method proposed by Sewall Wright in the 1920s is the foundation of modern structural causal models.
-Sewall Wright, a Statistican and Geneticist, sparked the interest in causal modelling with his path analysis which subsequently was adopted by various other disciplines such as econometrics, psychometrics and sociology.
+Path analysis is the foundation of modern structural causal models.
 This path analysis is a structural equation model with one variable per indicator.
+Sewall Wright, a Statistican and Geneticist, introduced the topic in the 1920s. (Pearl, 2012)
+Various other disciplines such as econometrics, psychometrics and sociology adopted path analysis.
+Aside from path diagrams also introduced graph rules to formalise relationships.
 
-Aside from path diagrams also introduced graph rules for writing down relationships.
+Trygve Haavelmo (1944) provided a framework for Economics in his paper 'The Probability Approach in Econometrics'.
+This paper introduced a new of level of sophistication to the discipline of Econometrics. 
+Haavelmo won the Nobel Memorial Prize in Economic Science in 1989 for his work on this issue.
 
-Thereafter the Norwegian Economist Trygve Haavelmo (1944) provided a modelling framework for the field of economics in is paper 'The Probability Approach in Econometrics.'
-One could argue that this paper provided a new of level of sophistication to the discipline of Econometrics as a distinct field. 
-Haavelmo was awarded the Nobel Memorial Prize in Economic Science in 1989 for his work on this issue.
-
-Causal assumptions are encoded in missing links.
+# TODO (other authors)
 
@@
 
+- Tests for causal relationship 
+- d-seperation
+
 # Conclusion

### 2021-05-14 draft.Rmd
order

--- 2021-04-19 draft.Rmd
@@
 ---
-title: "draft"
-author: "Daniel Saggau"
-date: "4/19/2021"
-output: pdf_document
+title: "A Gentle Introduction into Structural Causal Models"
+author: 
+- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
+- Department of Statistics, Ludwig Maximilian University Munich, Germany
+date: " April 5th, 2021"
+output: 
+  pdf_document:
+        number_sections: yes
+        fig_caption: yes
+fontsize: 12
+linestretch: 1.5
+citation_package: --biblatex
+bibliography: [ref.bib]
+nocite: '@*'
+abtract: The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling.
+To provide a comprehensive understanding of causal modelling, I introduce two prominent causal model specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM), focusing on the latter.Probabilistic specifications such as a BCN cast a model based on conditional probabilities.
+SCMs cast a model based on assignment functions and extend probabilistic models by specifying the data generating process rather than solely utilizing conditional probabilities.
+Another difference between these models is their ability to address different queries such as *predictions*, *interventions* and *counterfactuals*.
+These queries are part of Pearl's causal hierarchy (2009).
+Pearl matches these queries with their respective actions namely *observing*, *doing* and *imagining*.
+I compare the feasibility of addressing these queries and undertaking respective actions for both specifications.
+To contextualize SCMs within the field of causality, I also discuss the role of time in causal inference and provide a brief overview of the history of causal modelling.
+This paper uses various directed acyclic graphs to highlight the differences in these modelling approaches.
+The insights of this paper can be used as a baseline for subsequent research on structural causal models.
+
 ---
@@
 
-##
+# Introduction
+
+Most students have heard the phrase 'correlation does not imply causation.'
+While correlation implies co-occurrence, for many problems correlation is not enough. 
+Algorithmic decision making based on co-occurrence is insufficient  in high stake settings [@bareinboim2020pearl].
+For many problems we want to understand causal relationships between variables.
+There are different approaches on how to model causal relationships.
+This paper focuses on Structural Causal Models and Bayesian Causal Networks.
+Bayesian Causal Networks (BCN) cast a model based on **conditional probabilities.**
+In Figure 1 we can see an example of a probabilistic model depicted  as a directed acyclic graph (DAG).
+The nodes are the white circles.
+The edges are the arrows, defined by the conditional probabilities.
+C is our **collider** variable because the affect of A on C and T on C collide [@pearl2009causality].
+
+```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=7pt,
+    node distance=3cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\node[my node] (A){A};
+\node[my node,right =1 cm of A](C){C};
+\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
+\draw[->] (A) -- (T);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
+
+Causal inference introduced tools that combine combine different sources of knowledge. 
+We can e.g. include theoretical knowledge [@morgan_winship_2014] and observational knowledge with structural causal models.
+Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality].
+SCMs are non-parametric structural equations models with added features.  
+A structural equation model (SEM) is a set of autonous equations.
+This set represents the 'state of the world'. 
+SEMs are popular in fields like economics, psychology and sociology (Pearl 2009).
+SCM are also defined by a (sub-)set of structural equations.
+Additionally, SCMs  feature mathematical components from graphical models and the potential-outcome framework [@pearl2012causal].
+There is a lot of controversy around SEMs.
+Many scholars challenge the parameter-specification in a SEM [@pearl2012causal].
+^[Inconsistency refers to inherent nature of creating parameters based of observational data.
+Observational data  is seldom consistent.
+Henceforth respective estimates are often questioned.
+For further information see @hernan2008does]
+SCMs specify an underlying data generating process without the computational effort of creating a parametric model.
+Note, that structural equations refer to assignment equations, used in computer science. 
+Equations in causal models did not always have a concise notation (Pearl, 2009).
+First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
+Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
+This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
+The problem is that the equation is symmetric.
+The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
+This misconception has caused a lot of challenges which I will address further on in this paper. 
+As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
+B defines A and the latent factor $U_A$.
+To summarize, a SCM consists of a set of (autonomous) equations to generate (a) endogenous variables and (b) exogenous variables.
+In figure 2 we can see a vanilla structural causal model. 
+The square nodes represent the latent variables.
+The circle nodes represent the observed variables.
+The arrows are our structural equations and depict the relationship between variables.
+^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks.
+For further information see Pearl (2009).
+Due to the confined scope of this paper, I will not explore cyclic structures.]
 
@@
 
-```{tikz,fig.cap="Atomic Intervention", fig.align="center", echo =F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-  \begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (A){A};
-\node[latent,left =1 cm of A](UA){$U_A$};
-\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5 cm)$) (T) {t};
-\node[my node,right = 1 cm of T, fill=gray!30](C){C};
-\node[latent,right=1 cm of A](UC){$U_C$};
-\draw[->] (UA) -- (A);
-\draw[->] (UC) -- (C);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
+To contextualise causal methods, Judea Pearl (2009) introduced the hierarchy of causation.
+Pearl focuses on three layers: association, intervention and counterfactuals. 
+A higher level implies more detailed knowledge of the relationship between the variables.
+The first query is association, where we examine relationships based on observations.
 
-```{tikz,fig.cap="Policy Intervention",  fig.align= "center", echo =F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-  \begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=6pt,
-    node distance=2cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (A){A};
-\node[latent,left of=A](UA){$U_A$};
-\node[my node, fill=gray!70] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {$T_\pi$};
-\node[my node,right = 1 cm of T, fill=gray!30](C){C};
-\node[latent,left = 1cm of T](UT){$U_T$};
-\node[latent,right = 1cm of A](UC){$U_C$};
-\draw[->] (A) -- (T);
-\draw[->] (UA) -- (A);
-\draw[->] (UT) -- (T);
-\draw[->] (UC) -- (C);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
+\footnotesize
+ |Method          | Action |  Example | Usage | 
+|------------------|-------------|--------------------|-------------------|
+| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
+| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
 
-# Counterfactuals
+Table: Pearls Hierachy of Causation (2009)
 
-Process is described as follows:
+\normalsize
 
-(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
-(b) Action: Exchange $(X = x)$ 
-(c) Prediction: Compute $(Y = y)$
+Vanilla machine learning (ML), bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy (see table 1).
+These methods demand the least information and depend on association alone.
+Associational methods ignore external changes outside of our data.
+The interventional distribution has information on these external changes.
+The interventional distribution is only defined in high level causal methods. 
+The second query deals with interventions.
+Here we can use Pearls (2009) do-calculus.
+The do-calculus enables us to study the manipulation of parent nodes.
+ There are various types of intervention.
+One example is **atomic intervention,** where we set a variable to a constant.
+In **policy intervention** we specify a different function for an equation.
+off-policy intervention models different intervention that is not in our historical data [@oberst2019counterfactual].
 
-# Probabilistic Models vs SCMs
+Hypothetical interventions:
+
+Causal bayesian networks , Markov Decision Processes (MDP) and reinforcement learning model intervention.
+
+The third query is counterfactual modelling. 
+Here we deal with hypothetical settings. 
+SCMs and potential outcome models allow for counterfactual modelling.
+These models can model counterfactuals because they include a interventional distribution [@oberst2019counterfactual].
+BCNs only entail conditional probabilities.
+There is no information on relationships outside of the observations in the data.
+Henceforth, they cannot create counterfactuals (Pearl 2009).
+
+Another issue in this paper is understanding the relationship of time and causality.
+SCMs make the assumption that causal relationships hold over time (Peters et al. 2017).
+These vague definitions of time is more prevalent in social sciences. 
+Meanwhile hard sciences deal with time in a more concise manner. 
+Differential equations model time on a more mechanic manner (Mooij et al. 2013).
+
+Existing literature has provided an excellent introduction to SCMs and BCNs. 
+Pearl (2009) provided a comprehensive introduction into this topic with his book 'Causality'.
+His work addressed misconceptions in social sciences, causality and statistics. 
+Peters et al. (2017) address the relationship of causality, physical sciences, and causal inference.
+They also look at causal inference for observational data.
+E.g. [@bareinboim2020pearl] worked on the application of causal methods in machine learning. 
+@tarka2018overview discuss the history of SEMs and SCMs.
+This paper combines these contributions. 
+The aim of this paper is to summarise SCMs and it's intersection with social sciences, physics, and machine learning.
+I structure the rest of the paper as follows:
+
+# Foundations of Structural Causal Models
+
+Path analysis is the foundation of modern structural causal models.
+This path analysis is a structural equation model with one variable per indicator.
+Sewall Wright, a Statistican and Geneticist, introduced the topic in the 1920s. (Pearl, 2012)
+Various other disciplines such as econometrics, psychometrics and sociology adopted path analysis.
+Aside from path diagrams also introduced graph rules to formalise relationships.
+
+(tbc.)
+
+## Reichenbachs common causal Principle
+
+- noise independent -> called **causal suffieciency** clause.
+
+SCM fullfills via :
+
+- dependecy only via vertice connection (noise)
+- condition on noise, variables become independent
+
+## Assumptions
+
+- Independence of mechanisms 
+- conditional independence: 
+- joint distribution
+
+Density factorizes and their expectation
+
+Causal assumptions differentiate causal models from association learning methods.
+
+ | Association-based Concepts |  Causal Concepts
+|--------------------------|------------------------|
+| Correlation              | Randomization  
+| Regression               | Confounding
+| Conditional Independence | Disturbance
+| Likelihood               | Error Terms      
+| Odds Ratio               | Structural Coefficients       
+| Propensity Score         | Spurious Correlation
+
+Table: Concepts in Causality and Association concepts
+
+These causal concepts are not expressible based on distribution functions/statistical associations. (Pearl 2010)
+
+Disturbance in SCM: Correlated and causal factor ; responsible for variation
+Disturbance in Regression: Uncorrelated
+
+Causal assumption not testable (e.g. $Cov(U_a, U_b)=0$).
+d-seperation to test assumptions in totality (cannot make assumptions in isolation).
+
+exploit invariant characteristics of SEM whithout committing to shape. 
+structural if function autonomous and invariant to change in form of other functions.
+
+## Pearl Causal Hierachy
 
@@
 |                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+
+**Counterfactuals**
+
+Process is described as follows:
+
+(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
+(b) Action: Exchange $(X = x)$ 
+(c) Prediction: Compute $(Y = y)$
+
+## Graphical Models
+
+A graph in a SCM contains endogenous and exogenous variables. 
+In a directed graph all edges have arrows.
+Directed arrows are direct (causal) effects.
+No edge between variables means that there is no causal effect. 
+If we have some edges without arrows, we call that graph semi-directed.
+If we have all edges without arrows, we call that graph un-directed.
+The most famous graphical model is the acyclic graph.
+An acyclic graph has no roots that cause itself (directly and indirectly).
+Because we do not specify changes over time we typically only deal with acyclic models.
+This is partially because we assume the relationship to be constant over time.
+
+There are various tools to derive graphical models.^[The PC-Algorithm and the IC-Algorthm are two prominent examples.]
+
+- d-seperation 
+- vertices variables and arrows are direct effects.
+
+Any DAG can be written as  SCM.
+given joint disrtribution can be discover grahp? Markov condition
+
+## Causal Inference and Time
+
+$$
+\begin{array}{|l|l|l|l|l|l|}
+\hline \text { model } & \begin{array}{l}
+\text { predict in IID } \\
+\text { setting }
+\end{array} & \begin{array}{l}
+\text { predict under } \\
+\text { changing } \\
+\text { distributions / } \\
+\text { interventions }
+\end{array} & \begin{array}{l}
+\text { answer } \\
+\text { counter- } \\
+\text { factual } \\
+\text { questions }
+\end{array} & \begin{array}{l}
+\text { obtain } \\
+\text { physical } \\
+\text { insight }
+\end{array} & \begin{array}{l}
+\text { automatically } \\
+\text { learn from } \\
+\text { data }
+\end{array} \\
+\hline \begin{array}{l}
+\text { mechanistic } \\
+\text { model } \\
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & ? \\
+\hline \begin{array}{l}
+\text { structural } \\
+\text { causal model }
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{Y} ? ? \\
+\hline \begin{array}{l}
+\text { causal } \\
+\text { graphical } \\
+\text { model }
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{Y} ? \\
+\hline \begin{array}{l}
+\text { statistical } \\
+\text { model }
+\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} & \mathrm{Y} \\
+\hline
+\end{array}
+$$
+Table: Source: Peters et al. (2017)
+
+# Conclusion
+
+This paper provides a gentle introduction into structural causal models.
+SCMs entail many features, complementing research on association learning by providing depth. 
+This in turn, is of particular benefit for high stake decision settings. 
+SCMs differentiate from other methods through the specification of endogenous and exogenous variables, treating the exogenous factors as pivotal components of the actual model rather than assuming they are ommitable errors that are uncorrelated.
+As suggested by Pearls Hierachy, there are different levels to learning and each higher step can do anything the prior step can but with more detail and information. 
+Henceforth, as machin
+
+# References

### 2021-05-14 presentation_scm.Rmd
order

--- 2021-05-13 references/vanilla_presentation .Rmd
@@
 ---
-title: "Fairness in Machine Learning SS21"
-subtitle: "Beyond Static Fairness"
-date: "03/05/2021"
+title: "scm_presentation"
 author: "Daniel Saggau"
-output:
-  beamer_presentation:
-    theme: Madrid
-    colortheme: crane
+date: "5/14/2021"
+output: beamer_presentation
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
-
-# Dynamical Repercussions
-
-- Apple card discriminating women
-
-# Introduction
-
-- Fairness in dynamical system has become important because equalizing true positive rate at each step does not converge as fast in systems with e.g. population dynamics 
-- Research on dynamical systems has focused on markov decision processes
-- Prior research on causal fairness has focused in static systems
-
-# Fairness-Gym
-
-- Environment to simulate fairness
-
-# Why care about causal fairness models?
-
-- Fairness is not static (D'Amour et al. 2020) introduce shortcomings of existing fairness correction measures
-- From a modelling perspective, we can improve the specification (Markov Decision Process)
-- Regular models may express a model in conditional probabilities (probabilistic model) or may be expressed as differential equation 
-- SCMs are cast in functional form which is more stable (but may also be expressed as differential equation)
-- Additionally, we specify the model beyond conditional probabilities -> we specify the latent variables (exogenous variables not observable within our dataset)
 
@@
 
-**Application:**
+# Causal Inference and Time 
+\tiny
 
-- Off Policy Interventions
+$$
+\begin{array}{|l|l|l|l|l|l|}
+\hline \text { model } & \begin{array}{l}
+\text { predict IID } \\
+\text { setting }
+\end{array} & \begin{array}{l}
+\text { predict under } \\
+\text { changing } \\
+\text { distributions / } \\
+\text { interventions }
+\end{array} & \begin{array}{l}
+\text { answer } \\
+\text { counter- } \\
+\text { factual } \\
+\text { questions }
+\end{array} & \begin{array}{l}
+\text { obtain } \\
+\text { physical } \\
+\text { insight }
+\end{array} & \begin{array}{l}
+\text { automatically } \\
+\text { learn from } \\
+\text { data }
+\end{array} \\
+\hline \begin{array}{l}
+\text { mechanistic } \\
+\text { model } \\
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & ? \\
+\hline \begin{array}{l}
+\text { structural } \\
+\text { causal model }
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{Y} ? ? \\
+\hline \begin{array}{l}
+\text { causal } \\
+\text { graphical } \\
+\text { model }
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{Y} ? \\
+\hline \begin{array}{l}
+\text { statistical } \\
+\text { model }
+\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} & \mathrm{Y} \\
+\hline
+\end{array}
+$$

### 2021-05-14 references/appendix.Rmd
readme and stuff

--- 2021-05-11 scm.Rmd
@@
 ---
-title: "scm_presentation"
+title: "appendix"
 author: "Daniel Saggau"
-date: "5/11/2021"
-output: beamer_presentation
+date: "5/3/2021"
+output: pdf_document
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
-```
-
-# Introduction
-
-Causal assumptions differentiate causal models from association learning methods.
-
- | Association-based Concepts |  Causal Concepts
-|--------------------------|------------------------|
-| Correlation              | Randomization  
-| Regression               | Confounding
-| Conditional Independence | Disturbance
-| Likelihood               | Error Terms      
-| Odds Ratio               | Structural Coefficients       
-| Propensity Score         | Spurious Correlation
-
-# Fundamental Differences to begin
-
-SCM:
-
-- is a nonparametric SEM 
-- has functional form rather than using probabilities
-- entails features from the PO framework and graphical representation 
-- conflict whether to use graphs or not
-
-* A SEM is a parametric specification used in applied sciences (parameters contested)
-* A Bayesian causal network is another popular causal model using conditional probabilities and NO functions 
-* Differences in performance between BCN and SCM
-
-#
-
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
-
-# Graphical Illustration
-
-```{tikz,fig.align="center",fig.cap="Probabilistic Model", echo =F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\node[my node] (A){A};
-\node[my node,right =1 cm of A](C){C};
-\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
-\draw[->] (A) -- (T);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
-
-```{tikz,fig.cap="Structural Causal Model", fig.align="center", echo =F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-  \begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=6pt,
-    node distance=2cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (A){A};
-\node[latent,left of=A](UA){$U_A$};
-\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
-\node[my node,right = 1 cm of T, fill=gray!30](C){C};
-\node[latent,left = 1cm of T](UT){$U_T$};
-\node[latent,right = 1cm of A](UC){$U_C$};
-\draw[->] (A) -- (T);
-\draw[->] (UA) -- (A);
-\draw[->] (UT) -- (T);
-\draw[->] (UC) -- (C);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
-
-# Pearls Causal Hierachy
-
 \footnotesize
- |Method          | Action |  Example | Usage | 
+ |Method          | Action |  Example | Methods | 
 |------------------|-------------|--------------------|-------------------|
@@
 | Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
-| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           |SCM
 
 Table: Pearls Hierachy of Causation (2009)
-
 \normalsize

### 2021-05-14 references/intro.Rmd
readme and stuff

---
title: "A Gentle Introduction into Structural Causal Models"
author: 
- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
- Department of Statistics, Ludwig Maximilian University Munich, Germany
date: " April 5th, 2021"
output: 
  pdf_document:
        number_sections: yes
        fig_caption: yes
fontsize: 12
linestretch: 1.5
citation_package: --biblatex
bibliography: [ref.bib]
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Most undergraduate students that take a class in introductory statistics have heard the phrase 'correlation does not imply causation'.
While correlation implies co-occurrence, people are frequently interesting a causal understanding of relationships between variables.
One may argue that especially in high stake settings, algorithmic decision making based on co-occurrence is insufficient [@bareinboim2020pearl].
This paper tries to clarify structural differences between association-based models and causal models. 
There are different approaches on how to model causal relationships.
Structural Causal Models and Bayesian Causal Networks are arguably the two most prominent causal models for causal graphs.
Bayesian Causal Networks (BCN) cast a model based on conditional probabilities.
All relationships are defined in **conditional probabilities.**
BCNs ignore exogenous variables [@pearl2009causality].
In Figure 1 we can see an example of what a probabilistic model looks like in a directed acylic graph.
Our nodes are the white circles.
The edges are the arrows, defined by the conditional probabilities.
All variables in our graph are observed variables. 
We do not specify exogenous variables and C is our **collider** variable because the affect of A on C and T on C collide [@pearl2009causality].

```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=7pt,
    node distance=3cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\node[my node] (A){A};
\node[my node,right =1 cm of A](C){C};
\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
\draw[->] (A) -- (T);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

To accommodate the lack of data on exogenous variables, the field of causal inference introduced tools to combine various different sources of knowledge such as theoretical knowledge [@morgan_winship_2014].
This enables practitioners to include latent variables.
Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality] including latent variables.
SCMs are a nonparametric modification of structural equations models (a set of functional equations) with some extensions.
These extensions are some of the mathematical language from graphical models and the implementation of the potential outcome framework [@pearl2012causal].
Structural equations models (SEM) are parametric causal models which are very popular in fields like economics, psychology and sociology (Pearl 2009).
One should point out that there is a lot of controversy around SEMs because many scholars challenge the value of the parameters in a SEM [@pearl2012causal].
The central idea of SCMs is to provide a model specifying an underlying data generating process, without the redundancy of creating inconsistent parameter specifications.^[Inconsistency refers to inherent nature of creating parameters based of observational data. One inherent feature of observational data is that it is seldom consistent and henceforth respective estimates are questionable.For further information see @hernan2008does]

SCMs use assignment equations to specify an underlying data-generation-process.
Note that these equations have not always had a concise mathematical notation (Pearl, 2009).
Initially there was no sign to express the assignment equation and people used the '=' and one would e.g. write A=B.
But treating an equation as a **algebraic equation**, namely an equation that is symmetric, makes no sense in the causality context because this equation has no causal information.
This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations hence the equation is symmetric.
The initial = sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
This misconception has caused a lot of challenges which I will address further on in this paper. 
As mentioned, we define variables as functions therefore a variable could be fined as $A=f_A(B,U_A)$ where A is defined by B and the latent factor $U_A$.
To summarize, a SCM consists of a set of (autonomous) equations to generate (a) endogenous variables and (b) exogenous variables.

```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left of=A](UA){$U_A$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,left = 1cm of T](UT){$U_T$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (T);
\draw[->] (UA) -- (A);
\draw[->] (UT) -- (T);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

In figure 2 we can see a plain-vanilla structural causal model. 
The square nodes represent the latent variables.
The circle nodes represent the observed variables.
The arrows are our structural equations and depict the relationship between variables.
^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks. For further information see Pearl (2009). Due to the confined scope of this paper, I will not explore cyclic structures.]

Beyond the different causal model specifications, Judea Pearl (2009) introduced the hierarchy of causation.
Pearl focuses on three layers namely association, intervention and counterfactuals. 
A higher level implies a more detailed knowledge of the relationship between the variables.
The first query is prediction where we examine based on observations.
We can see that machine learning methods, bayesian networks and regression models are at the lowest level in the causal hierarchy.
These methods depend on association alone and lack information on what happens if we make changes in our root nodes (our nodes that are not caused by other variables) and what would happen if we had a different setting.

\footnotesize
 |Method          | Action |  Example | Usage | 
|------------------|-------------|--------------------|-------------------|
| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO

Table: Pearls Hierachy of Causation (2009)

\normalsize
The second query deals with interventions.
Here we can use Pearls (2009) do-calculus to examine what happens if root nodes are manipulated.
Note that there are various different types of intervention, such as atomic intervention, where we set a variable to a constant, or e.g. policy intervention where we specify a different function for a specific equation in our set of equations.
Additionally there is also off-policy intervention, where we model a different intervention that is not in our historical data based on the historical data [@oberst2019counterfactual].
Causal bayesian networks , markov decision processes (MDP) and reinforcement learning frequent these intervention methods.

The third query is counterfactual modelling. 
Here we deal with hypothetical settings. 
SCMs and potential outcome models allow for counterfactual modelling.
The reason why these models can model counterfactuals is because they also include information on the interventional distribution [@oberst2019counterfactual].
Given that e.g. BCN only entail conditional probabilities and no information on relationships outside of the observations in the data, they cannot create counterfactuals (Pearl 2009).

Another issue in this paper is understanding the relationship of time and causal modelling.
Causal models mostly disregard concise notions of time and make the strong assumption that relationships between variables hold beyond the confinement of time (Peters et al. 2017).
These vague definitions of time is more prevalent in social sciences. 
Meanwhile hard sciences deal with time in a more concise manner. 
Some research has also looked at causal models with a more concise specification of time using differential equations (Mooij et al. 2013).

To summarize, existing literature has provided an excellent introduction to structural causal models and bayesian causal networks. 
Judea Pearl (2009) provided a comprehensive and accessible introduction into this topic with his book on causality.
His work has addressed various misconceptions in social sciences, economics, causality and statistics. 
Peters et al. (2017) added to this discourse by addressing the relationship of causal modelling to physical sciences and causal inference for observational data.
Additionally, other scholars e.g. [@bareinboim2020pearl] have worked on the application of causal methods in machine learning. 
This paper brings together these contributions. 
The objective of this paper is to provide a comprehensive summary of structural causal models and respective applications in social sciences, physical sciences and machine learning.
The insights in this paper are an abstraction of these pieces of work, offering the reading a basic understanding of the topic for future research and an overview of the current research.

The rest of the paper is structured as follows: 
The upcoming section studies the hierarchy of causation, looking at the different queries in the SCM and the BCN and what implications the functional/conditional probability specification carry.
The third section provides further reference to the intersection of machine learning and structural causal models. 
The fourth section studies differences and similarities between SCMs and causal models (SEM,PO) that emerged in the social sciences. 
The fifth section examines the relationship of physical sciences and SCMs, addressing the issue of time and modelling based on differential equations.

# References

### 2021-05-14 draft.Rmd
paper stuff

@@
 
-Hypothetical interventions:
+(tbc.) Hypothetical interventions:
 
@@
 This paper combines these contributions. 
-The aim of this paper is to summarise SCMs and it's intersection with social sciences, physics, and machine learning.
+The aim of this paper is to summarise SCMs and it's intersection with social sciences and physics.
 I structure the rest of the paper as follows:
@@
 ## Reichenbachs common causal Principle 
-
 - noise independent -> called **causal suffieciency** clause.
@@
 Causal assumptions differentiate causal models from association learning methods.
-
- | Association-based Concepts |  Causal Concepts
-|--------------------------|------------------------|
-| Correlation              | Randomization  
-| Regression               | Confounding
-| Conditional Independence | Disturbance
-| Likelihood               | Error Terms      
-| Odds Ratio               | Structural Coefficients       
-| Propensity Score         | Spurious Correlation
-
-Table: Concepts in Causality and Association concepts
 

### 2021-05-14 draft.Rmd
structure

@@
 I compare the feasibility of addressing these queries and undertaking respective actions for both specifications.
-To contextualize SCMs within the field of causality, I also discuss the role of time in causal inference and provide a brief overview of the history of causal modelling.
+To contextualize SCMs within the field of causality, I also discuss the role of time in causality.
 This paper uses various directed acyclic graphs to highlight the differences in these modelling approaches.
 The insights of this paper can be used as a baseline for subsequent research on structural causal models.
-
 ---
@@
 
+Consists of graph and assignments: 
+Baseline:
+
+$$C:= N_c$$
+$$E:=f_E(C,N_E)$$
+
+source: Peters et al. (2017)
+
+# Brief historical introduction:
+
 Path analysis is the foundation of modern structural causal models.
@@
 
-(tbc.)
+(tbc. Cowles, Haavelmo,Aldrich, )
 
-## Reichenbachs common causal Principle 
-- noise independent -> called **causal suffieciency** clause.
+# Assumptions in Causality
 
-SCM fullfills via :
+**Reichenbachs common causal Principle** 
+**Independence**
+
+If we specify the causal structure correctly:
+
+(a) possible to undertake local intervention -> change f(x), regardless of f(y|x)
+(b) these components are autonmous objects -> set of autonmous equations
+
+Further: independence of noise -> $$N_T , N_A$$
+
+Example conditional probabilities: independence conditional probabilities/distribution and mechanism (ICM)
+
+(peters et al., 2017)
+
+- noise independent -> called **causal sufficiency** clause. 
+- Algorithmic independence
+
+SCM fulfills via :
 
@@
 - condition on noise, variables become independent  
-
-## Assumptions
-
 - Independence of mechanisms 
@@
 
-## Pearl Causal Hierachy
+# Pearl Causal Hierachy
 

### 2021-05-14 draft.Rmd
commit draft

@@
 nocite: '@*'
-abtract: The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling.
+
+---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
+
+**Abstract** The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling.
 To provide a comprehensive understanding of causal modelling, I introduce two prominent causal model specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM), focusing on the latter.Probabilistic specifications such as a BCN cast a model based on conditional probabilities.
@@
 The insights of this paper can be used as a baseline for subsequent research on structural causal models.
----
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
+\newpage
+\tableofcontents
+\newpage
 
@@
 Bayesian Causal Networks (BCN) cast a model based on **conditional probabilities.**
-In Figure 1 we can see an example of a probabilistic model depicted  as a directed acyclic graph (DAG).
-The nodes are the white circles.
-The edges are the arrows, defined by the conditional probabilities.
-C is our **collider** variable because the affect of A on C and T on C collide [@pearl2009causality].
-
-```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\node[my node] (A){A};
-\node[my node,right =1 cm of A](C){C};
-\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
-\draw[->] (A) -- (T);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
-
-Causal inference introduced tools that combine combine different sources of knowledge. 
-We can e.g. include theoretical knowledge [@morgan_winship_2014] and observational knowledge with structural causal models.
 Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality].
 SCMs are non-parametric structural equations models with added features.
-A structural equation model (SEM) is a set of autonous equations.
-This set represents the 'state of the world'. 
-SEMs are popular in fields like economics, psychology and sociology (Pearl 2009).
-SCM are also defined by a (sub-)set of structural equations.
-Additionally, SCMs  feature mathematical components from graphical models and the potential-outcome framework [@pearl2012causal].
-There is a lot of controversy around SEMs.
-Many scholars challenge the parameter-specification in a SEM [@pearl2012causal].
-^[Inconsistency refers to inherent nature of creating parameters based of observational data.
-Observational data  is seldom consistent.
-Henceforth respective estimates are often questioned.
-For further information see @hernan2008does]
-SCMs specify an underlying data generating process without the computational effort of creating a parametric model.
-Note, that structural equations refer to assignment equations, used in computer science. 
-Equations in causal models did not always have a concise notation (Pearl, 2009).
-First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
-Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
-This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
-The problem is that the equation is symmetric.
-The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
-This misconception has caused a lot of challenges which I will address further on in this paper. 
-As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
-B defines A and the latent factor $U_A$.
-To summarize, a SCM consists of a set of (autonomous) equations to generate (a) endogenous variables and (b) exogenous variables.
-In figure 2 we can see a vanilla structural causal model. 
-The square nodes represent the latent variables.
-The circle nodes represent the observed variables.
-The arrows are our structural equations and depict the relationship between variables.
-^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks.
-For further information see Pearl (2009).
-Due to the confined scope of this paper, I will not explore cyclic structures.]
-
-```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-  \begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=6pt,
-    node distance=2cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (A){A};
-\node[latent,left of=A](UA){$U_A$};
-\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
-\node[my node,right = 1 cm of T, fill=gray!30](C){C};
-\node[latent,left = 1cm of T](UT){$U_T$};
-\node[latent,right = 1cm of A](UC){$U_C$};
-\draw[->] (A) -- (T);
-\draw[->] (UA) -- (A);
-\draw[->] (UT) -- (T);
-\draw[->] (UC) -- (C);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
-
-To contextualise causal methods, Judea Pearl (2009) introduced the hierarchy of causation.
-Pearl focuses on three layers: association, intervention and counterfactuals. 
-A higher level implies more detailed knowledge of the relationship between the variables.
-The first query is association, where we examine relationships based on observations.
-
-\footnotesize
- |Method          | Action |  Example | Usage | 
-|------------------|-------------|--------------------|-------------------|
-| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
-| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
-| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
-
-Table: Pearls Hierachy of Causation (2009)
-
-\normalsize
-
-Vanilla machine learning (ML), bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy (see table 1).
-These methods demand the least information and depend on association alone.
-Associational methods ignore external changes outside of our data.
-The interventional distribution has information on these external changes.
-The interventional distribution is only defined in high level causal methods. 
-The second query deals with interventions.
-Here we can use Pearls (2009) do-calculus.
-The do-calculus enables us to study the manipulation of parent nodes.
- There are various types of intervention.
-One example is **atomic intervention,** where we set a variable to a constant.
-In **policy intervention** we specify a different function for an equation.
-off-policy intervention models different intervention that is not in our historical data [@oberst2019counterfactual].
-
-(tbc.) Hypothetical interventions:
-
-Causal bayesian networks , Markov Decision Processes (MDP) and reinforcement learning model intervention.
-
-The third query is counterfactual modelling. 
-Here we deal with hypothetical settings. 
-SCMs and potential outcome models allow for counterfactual modelling.
-These models can model counterfactuals because they include a interventional distribution [@oberst2019counterfactual].
-BCNs only entail conditional probabilities.
-There is no information on relationships outside of the observations in the data.
-Henceforth, they cannot create counterfactuals (Pearl 2009).
-
-Another issue in this paper is understanding the relationship of time and causality.
-SCMs make the assumption that causal relationships hold over time (Peters et al. 2017).
-These vague definitions of time is more prevalent in social sciences. 
-Meanwhile hard sciences deal with time in a more concise manner. 
-Differential equations model time on a more mechanic manner (Mooij et al. 2013).
-
 Existing literature has provided an excellent introduction to SCMs and BCNs. 
@@
 His work addressed misconceptions in social sciences, causality and statistics.
+His contribution on the hierarchy of causation advanced the comprehensive understanding of causality. 
 Peters et al. (2017) address the relationship of causality, physical sciences, and causal inference.
-They also look at causal inference for observational data.
+Their work highlights the importance of time in causality and inference based on observational data alone. 
 E.g. [@bareinboim2020pearl] worked on the application of causal methods in machine learning. 
 @tarka2018overview discuss the history of SEMs and SCMs.
-This paper combines these contributions. 
 The aim of this paper is to summarise SCMs and it's intersection with social sciences and physics.
 I structure the rest of the paper as follows: 
+Section 2 introduces the assumptions in causal modelling. Further, section 2 discusses different causal differencecs, fundamental differences and a brief history of the methods. 
+Section 3 addresses Pearls Causal Hierachy
 
@@
 
-# Brief historical introduction:
+## Assumptions in Causality
 
-Path analysis is the foundation of modern structural causal models.
-This path analysis is a structural equation model with one variable per indicator.
-Sewall Wright, a Statistican and Geneticist, introduced the topic in the 1920s. (Pearl, 2012)
-Various other disciplines such as econometrics, psychometrics and sociology adopted path analysis.
-Aside from path diagrams also introduced graph rules to formalise relationships.
-
-(tbc. Cowles, Haavelmo,Aldrich, )
-
-# Assumptions in Causality
-
-**Reichenbachs common causal Principle** 
-**Independence**
+*Definition:***Independence**
 
@@
 
-# Pearl Causal Hierachy
+*Definition:* **Reichenbachs common causal Principle**
 
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+## Comparing Causal Tools
 
-**Counterfactuals**
+A structural equation model (SEM) is a set of autonous equations.
+This set represents the 'state of the world'. 
+SEMs are popular in fields like economics, psychology and sociology (Pearl 2009).
+SCM are also defined by a (sub-)set of structural equations.
+Additionally, SCMs  feature mathematical components from graphical models and the potential-outcome framework [@pearl2012causal].
+There is a lot of controversy around SEMs.
+Many scholars challenge the parameter-specification in a SEM [@pearl2012causal].
+^[Inconsistency refers to inherent nature of creating parameters based of observational data.
+Observational data  is seldom consistent.
+Henceforth respective estimates are often questioned.
+For further information see @hernan2008does]
+SCMs specify an underlying data generating process without the computational effort of creating a parametric model.
+Note, that structural equations refer to assignment equations, used in computer science. 
+Equations in causal models did not always have a concise notation (Pearl, 2009).
+First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
+Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
+This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
+The problem is that the equation is symmetric.
+The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
+This misconception has caused a lot of challenges which I will address further on in this paper. 
+As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
+B defines A and the latent factor $U_A$.
+To summarize, a SCM consists of a set of (autonomous) equations to generate (a) endogenous variables and (b) exogenous variables.
+
+In Figure 1 we can see an example of a probabilistic model depicted  as a directed acyclic graph (DAG).
+The nodes are the white circles.
+The edges are the arrows, defined by the conditional probabilities.
+C is our **collider** variable because the affect of A on C and T on C collide [@pearl2009causality].
+
+```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=7pt,
+    node distance=3cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\node[my node] (A){A};
+\node[my node,right =1 cm of A](C){C};
+\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
+\draw[->] (A) -- (T);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
+
+In figure 2 we can see a vanilla structural causal model. 
+The square nodes represent the latent variables.
+
+```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+  \begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=6pt,
+    node distance=2cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
+\node[my node, fill=gray!30] (A){A};
+\node[latent,left of=A](UA){$U_A$};
+\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
+\node[my node,right = 1 cm of T, fill=gray!30](C){C};
+\node[latent,left = 1cm of T](UT){$U_T$};
+\node[latent,right = 1cm of A](UC){$U_C$};
+\draw[->] (A) -- (T);
+\draw[->] (UA) -- (A);
+\draw[->] (UT) -- (T);
+\draw[->] (UC) -- (C);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
+
+The circle nodes represent the observed variables.
+The arrows are our structural equations and depict the relationship between variables.
+^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks.
+For further information see Pearl (2009).
+Due to the confined scope of this paper, I will not explore cyclic structures.]
+
+## A Brief History of Causal Modeling
+
+Path analysis is the foundation of modern structural causal models.
+This path analysis is a structural equation model with one variable per indicator.
+Sewall Wright, a Statistican and Geneticist, introduced the topic in the 1920s. (Pearl, 2012)
+Various other disciplines such as econometrics, psychometrics and sociology adopted path analysis.
+Aside from path diagrams also introduced graph rules to formalise relationships.
+(tbc. Cowles, Haavelmo,Aldrich, )
+
+# Pearl's Causal Hierachy
+
+To contextualise causal methods, Judea Pearl (2009) introduced the hierarchy of causation.
+Pearl focuses on three layers: association, intervention and counterfactuals (Table 1).
+
+ |Method          | Action | | Usage | 
+|------------------|-------------|-------------------|
+| Association $P(a|b)$               | Co-occurrence                           |(Un-)Supervised ML, BN, Reg.  
+| Intervention $P(a|do(b),c)$       | Do-manipulation                       |CBN,MDP,RL    
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals            | SCM ,PO
+
+A higher level implies more detailed knowledge of the relationship between the variables.
+The first query is association, where we examine relationships based on observations.
+
+Table: Pearls Hierachy of Causation (2009)
+
+## Prediction
+
+Vanilla machine learning (ML), bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy (see table 1).
+These methods demand the least information and depend on association alone.
+Associational methods ignore external changes outside of our data.
+The interventional distribution has information on these external changes.
+The interventional distribution is only defined in high level causal methods.
+
+## Intervention
+
+The second query deals with interventions.
+Here we can use Pearls (2009) do-calculus.
+The do-calculus enables us to study the manipulation of parent nodes.
+ There are various types of intervention.
+One example is **atomic intervention,** where we set a variable to a constant.
+In **policy intervention** we specify a different function for an equation.
+off-policy intervention models different intervention that is not in our historical data [@oberst2019counterfactual].
+Causal bayesian networks , Markov Decision Processes (MDP) and reinforcement learning model intervention.
+(tbc.) Hypothetical interventions:
+
+## Counterfactuals
+
+The third query is counterfactual modelling. 
+Here we deal with hypothetical settings. 
+SCMs and potential outcome models allow for counterfactual modelling.
+These models can model counterfactuals because they include a interventional distribution [@oberst2019counterfactual].
+BCNs only entail conditional probabilities.
+There is no information on relationships outside of the observations in the data.
+Henceforth, they cannot create counterfactuals (Pearl 2009).
 
@@
 
-## Graphical Models
+## Implications
+
+BCN:
+
+SCM:
+
+# Graphical Models
+
+## Features
 
@@
 
+## Implications
+
 Any DAG can be written as  SCM.
@@
 
-## Causal Inference and Time
+# Causal Inference and Time
+
+Another issue in this paper is understanding the relationship of time and causality.
+SCMs make the assumption that causal relationships hold over time (Peters et al. 2017).
+These vague definitions of time is more prevalent in social sciences. 
+Meanwhile hard sciences deal with time in a more concise manner. 
+Differential equations model time on a more mechanic manner (Mooij et al. 2013).
 
@@
 $$
+
 Table: Source: Peters et al. (2017)
@@
 
+\newpage
+
 # References

### 2021-05-14 references/appendix.Rmd
commit draft

@@
 \normalsize
+
+ |Method         | CBN |  SCM
+|----------------|---------------------------|------------------------|
+| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
+|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
+|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
+|                |                                                                                                                      | 
+| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
+|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
+|                |                                                                                                                      | 
+|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
+|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+
+Table: Source: Pearl (2009)

### 2021-05-14 draft.Rmd
reference color adjustment

@@
         fig_caption: yes
+link-citations: yes
+linkcolor: blue
 fontsize: 12
@@
 \newpage
+\hypersetup{linkcolor=black}
 \tableofcontents
 \newpage
+
+\hypersetup{linkcolor=blue}
 

### 2021-05-14 draft.Rmd
peters

@@
 His contribution on the hierarchy of causation advanced the comprehensive understanding of causality. 
-Peters et al. (2017) address the relationship of causality, physical sciences, and causal inference.
+@peters_elements_2017 address the relationship of causality, physical sciences, and causal inference.
 Their work highlights the importance of time in causality and inference based on observational data alone. 
@@
 
-source: Peters et al. (2017)
+source: @peters_elements_2017
 
@@
 
-(peters et al., 2017)
+[@peters_elements_2017]
 
@@
 Another issue in this paper is understanding the relationship of time and causality.
-SCMs make the assumption that causal relationships hold over time (Peters et al. 2017).
+SCMs make the assumption that causal relationships hold over time[@peters_elements_2017].
 These vague definitions of time is more prevalent in social sciences. 
@@
 
-Table: Source: Peters et al. (2017)
+Table: Source: @peters_elements_2017
 

### 2021-05-14 draft.Rmd
citations

@@
 Existing literature has provided an excellent introduction to SCMs and BCNs. 
-Pearl (2009) provided a comprehensive introduction into this topic with his book 'Causality'.
+@pearl2009causality provided a comprehensive introduction into this topic with his book 'Causality'.
 His work addressed misconceptions in social sciences, causality and statistics.
@@
 This set represents the 'state of the world'. 
-SEMs are popular in fields like economics, psychology and sociology (Pearl 2009).
+SEMs are popular in fields like economics, psychology and sociology [@pearl2009causality].
 SCM are also defined by a (sub-)set of structural equations.
@@
 The second query deals with interventions.
-Here we can use Pearls (2009) do-calculus.
+Here we can use @pearl2009causality do-calculus.
 The do-calculus enables us to study the manipulation of parent nodes.
@@
 There is no information on relationships outside of the observations in the data.
-Henceforth, they cannot create counterfactuals (Pearl 2009).
+Henceforth, they cannot create counterfactuals [@pearl2009causality].
 
@@
 Meanwhile hard sciences deal with time in a more concise manner. 
-Differential equations model time on a more mechanic manner (Mooij et al. 2013).
+Differential equations model time on a more mechanic manner (@mooij2013ordinary).
 

### 2021-05-15 draft.Rmd
xyt
mathpazo font

@@
 fontsize: 12
+fontfamily: mathpazo 
 linestretch: 1.5
@@
 nocite: '@*'
-
 ---
@@
 Meanwhile hard sciences deal with time in a more concise manner. 
-Differential equations model time on a more mechanic manner (@mooij2013ordinary).
+Differential equations model time on a more mechanic manner [@mooij2013ordinary].
 

### 2021-05-15 draft.Rmd
appendix

@@
 linestretch: 1.5
-citation_package: --biblatex
+#citation_package: --biblatex
 bibliography: [ref.bib]
+citation_package: natbib
+biblio-style: alphadin
 nocite: '@*'
@@
 # References
+
+<div id="refs"></div>
+\newpage
+
+# Appendix

### 2021-05-16 scm.Rmd
scm stuff

@@
 
+# Motivation
+
+```{r, out.width= "70%", out.height = "70%", fig.align='center'}
+library(knitr)
+setwd("~/Desktop/SOSE2021/causal_ml")
+include_graphics('./images/correlation.jpg')
+```
+
 # Introduction (1)
 
-- Machine learning has provided many insights into different research problems 
+- Machine learning has provided many insights into different problems 
 - One issue is the consideration of 'What are we actually predicting'?
@@
 | Propensity Score         | Spurious Correlation
+
+# Table of Contents
+
+* Foundations of SCMs 
+  * Assumptions 
+  * Comparative Causal Tools 
+  * Historical Development 
+* Pearl's Causal Hierachy 
+  * prediction 
+  * intervention 
+  * Counterfactuals 
+  * Implications 
+* Graphical Models 
+  * Features 
+  * Implications
+* Causality and Time
 

### 2021-05-16 scm.Rmd
scm structure

@@
 
-# History
-
-* Path Analysis -> SEM -> SCM
-*
-
-# How to build a SCM
+# Assumptions
 
@@
 
+# Assumptions
+
+**Error terms**
+
+Regression: Omittable outside factor 
+SCM/SEM; Latent influencial factor that is pivotal for the model specification but not observable
+
+Consists of graph and assignments: 
+Baseline:
+
+$$C:= N_c$$
+$$E:=f_E(C,N_E)$$
+
+source: @peters_elements_2017
+
 # SCM Applications:
@@
 - Graphical visualization via directed acyclic graph
--
+
+# Comparative Causal Tools
+
+# Historical Development
+
+* Path Analysis -> SEM -> SCM
+
+# Fundamental Differences (1)
+
+- conflict whether to use graphs or not 
+* A SEM is a parametric specification used in applied sciences (parameters contested)
+* A Bayesian causal network is another popular causal model using conditional probabilities and NO functions 
+* Differences in performance between BCN and SCM# Performance Evaluation
+
+# Implications
+
+ |Method         | CBN |  SCM
+|----------------|---------------------------|------------------------|
+| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
+|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
+|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
+|                |                                                                                                                      | 
+| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
+|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
+|                |                                                                                                                      | 
+|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
+|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+
+# Pearls Causal Hierachy
+
+\footnotesize
+ |Method          | Action |  Example | Usage | 
+|------------------|-------------|--------------------|-------------------|
+| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
+| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
+
+Table: Pearls Hierachy of Causation (2009)
+
+\normalsize
 
@@
 
-# Fundamental Differences (1)
-
-- conflict whether to use graphs or not 
-* A SEM is a parametric specification used in applied sciences (parameters contested)
-* A Bayesian causal network is another popular causal model using conditional probabilities and NO functions 
-* Differences in performance between BCN and SCM
-
-# Performance Evaluation
-
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
-
-# Assumptions
-
-**Error terms**
-
-Regression: Omittable outside factor 
-SCM/SEM; Latent influencial factor that is pivotal for the model specification but not observable
-
-# Pearls Causal Hierachy
-
-\footnotesize
- |Method          | Action |  Example | Usage | 
-|------------------|-------------|--------------------|-------------------|
-| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
-| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
-| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
-
-Table: Pearls Hierachy of Causation (2009)
-
-\normalsize
-
-# Causality -In Between Physical and Social Sciences
-
 # Causality and Time

### 2021-05-16 scm.Rmd
draft

@@
 output: beamer_presentation
+citation_package: --biblatex
+bibliography: [ref.bib]
 ---
@@
 Regression: Omittable outside factor 
-SCM/SEM; Latent influencial factor that is pivotal for the model specification but not observable
+SCM/SEM; Latent influential factor that is pivotal for the model specification but not observable
 
@@
 - To accomodate that issue, research on differential equation based SCMs started
+
+# References

### 2021-05-16 scm.Rmd
struc

@@
 
-```{r, out.width= "70%", out.height = "70%", fig.align='center'}
+```{r, out.width= "80%", out.height = "80%", fig.align='center'}
 library(knitr)
@@
 
-# Assumptions
+# Assumptions (1)
 
@@
 
-# Assumptions
+# Assumptions (2)
 
@@
 
+# Prediction
+
+Vanilla machine learning (ML), bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy (see table 1).
+These methods demand the least information and depend on association alone.
+Associational methods ignore external changes outside of our data.
+The interventional distribution has information on these external changes.
+The interventional distribution is only defined in high level causal methods.
+
+# Intervention
+
+The second query deals with interventions.
+Here we can use @pearl2009causality do-calculus.
+The do-calculus enables us to study the manipulation of parent nodes.
+ There are various types of intervention.
+One example is **atomic intervention,** where we set a variable to a constant.
+In **policy intervention** we specify a different function for an equation.
+off-policy intervention models different intervention that is not in our historical data [@oberst2019counterfactual].
+Causal bayesian networks , Markov Decision Processes (MDP) and reinforcement learning model intervention.
+
+# Counterfactuals
+
+Process is described as follows:
+
+(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
+(b) Action: Exchange $(X = x)$ 
+(c) Prediction: Compute $(Y = y)$
+
 # Graphical Illustration

### 2021-05-16 draft.Rmd
random bib exploration with citr
scm

--- 2021-05-15 draft.Rmd
@@
 
-A structural equation model (SEM) is a set of autonous equations.
+A structural equation model (SEM) is a set of autonomous equations.
 This set represents the 'state of the world'. 
@@
 The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
-This misconception has caused a lot of challenges which I will address further on in this paper. 
+This misconception has caused a lot of challenges.^[For more information see @pearl2009causality]
 As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.

### 2021-05-16 scm.Rmd
scm

@@
 source: @peters_elements_2017
+
+# Assumptions (3)
+
+First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
+Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
+This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
+The problem is that the equation is symmetric.
+The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
+This misconception has caused a lot of challenges.^[For more information see @pearl2009causality]
+As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
 

### 2021-05-17 draft.Rmd
hardt chapter

@@
 Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality].
+Summarized, *'they are simulators that generate a distribution based noise variables (independent) and a unique joint distribution...'* [@hardtrecht].
 SCMs are non-parametric structural equations models with added features.
@@
 ## Assumptions in Causality
+
+**Established Conditions:**
+
+Definition: **SUTVA**
+
+'The treatment that one unit receives does not change the effect of treatment for any other unit.'[@hardtrecht]
+
+Definition: **Consistency**
+
+The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' [@hardtrecht]
+
+Definition: **Ignorability**
+
+The potential outcomes are conditionally independent of treatment given some set of deconfounding variables. [@hardtrecht]
+(perfect RCT)
+
+- First tow hold for SCM counterfactuals 
+- third not testable but can check via backdoor criterion in SCM
 
@@
 ^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks.
-For further information see Pearl (2009).
-Due to the confined scope of this paper, I will not explore cyclic structures.]
+For further information see  @pearl2009causality.
+Due to the confined scope of this paper, I will not explore cyclic structures.
+
+Potential outcomes treat outcomes merely as potential outcomes rather rather than true outcomes. 
+- Only observe outcomes for units that are treated
+- ATE is a typical query
+- Every SCM can generate PO 
+- Cannot derive SCM from PO alone 
+- SCM has more concepts : causal graphs, mediating paths, counterfactuals for every variable etc. 
+- PO Easier to apply because apply to observational data [@hardtrecht]
 
@@
 
-To contextualise causal methods, Judea Pearl (2009) introduced the hierarchy of causation.
+To contextualise causal methods,  @pearl2009causality introduced the hierarchy of causation.
 Pearl focuses on three layers: association, intervention and counterfactuals (Table 1).
@@
 A higher level implies more detailed knowledge of the relationship between the variables.
+ @pearl2009causality introduced the do-calculus.
+ The do-calculus is a mathematical framework allowing for active substitution (E.g. $X:=10$) [@hardtrecht].
 The first query is association, where we examine relationships based on observations.
@@
 
-## Prediction
+## Association
 

### 2021-05-17 draft.Rmd
causal
ref ml fairness

@@
 Summarized, *'they are simulators that generate a distribution based noise variables (independent) and a unique joint distribution...'* [@hardtrecht].
+- noise variables helo generate joint distribution [@barocas].
 SCMs are non-parametric structural equations models with added features.
@@
 
+Also called causal diagram
+- collider, mediator, fork [@barocas]
+
 ## Features
 
-A graph in a SCM contains endogenous and exogenous variables. 
+A graph in a SCM contains endogenous and exogenous variables 
 In a directed graph all edges have arrows.
@@
 Because we do not specify changes over time we typically only deal with acyclic models.
-This is partially because we assume the relationship to be constant over time.
+This is partially because we assume the relationship to be constant over time. [@creager]
 
@@
 ## Implications
+
+- backdoor criterion non experimental way remove confounding [@barocas]
 
@@
 # Appendix
+
+ |Method         | CBN |  SCM
+|----------------|---------------------------|------------------------|
+| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
+|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
+|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
+|                |                                                                                                                      | 
+| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
+|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
+|                |                                                                                                                      | 
+|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
+|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+
+Table: Source: Pearl (2009)

### 2021-05-20 presentation_video.Rmd
independence assumption
toc

--- 2021-05-16 scm.Rmd
@@
 * Pearl's Causal Hierachy 
-  * prediction 
-  * intervention 
+  * Prediction 
+  * Intervention 
   * Counterfactuals 
@@
 - Assignment equation ':=' rather than regular equation '=' 
-- is a nonparametric SEM 
-- has functional form rather than using probabilities
+- Nonparametric SEM 
+- Functional form rather than using  conditional probabilities
 - entails features from the PO framework and graphical representation 
@@
 As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
+
+# Assumptions (4)
+
+**Independence**
+
+*Definition:*
 

### 2021-05-20 presentation_video.Rmd
shorter

@@
 date: "5/11/2021"
-output: beamer_presentation
 citation_package: --biblatex
 bibliography: [ref.bib]
+output: 
+  beamer_presentation:
+    theme: Madrid
+
 ---
@@
 
-- Consists of system of equations 
+- System of equations 
 - Assignment equation ':=' rather than regular equation '=' 
@@
 - Functional form rather than using conditional probabilities
-- entails features from the PO framework and graphical representation 
+- SCM = PO + Graphical Tools
+
 - Exogenous factors are part of the model specification
@@
 
-Vanilla machine learning (ML), bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy (see table 1).
-These methods demand the least information and depend on association alone.
-Associational methods ignore external changes outside of our data.
-The interventional distribution has information on these external changes.
-The interventional distribution is only defined in high level causal methods.
+ -ML, BN and regression are at the lowest level in the causal hierarchy
+- Prediction methods demand the least information and depend on association alone
+- Associational methods ignore external changes outside of our data.
+- intervention distribution has information on these external changes.
+- intervention distribution is only defined in high level causal methods.
 

### 2021-05-20 draft.Rmd
draft
presentation

@@
 
-The potential outcomes are conditionally independent of treatment given some set of deconfounding variables. [@hardtrecht]
+The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. [@hardtrecht]
 (perfect RCT)
@@
 (a) possible to undertake local intervention -> change f(x), regardless of f(y|x)
-(b) these components are autonmous objects -> set of autonmous equations
+(b) these components are autonomous objects -> set of autonomous equations
 
@@
 
-- dependecy only via vertice connection (noise)
+- dependency only via vertice connection (noise)
 - condition on noise, variables become independent  
@@
 
+- no need functional dependency -> less assumptions about form
+- show collider variables
+- iv backdoor, estimation by mechanism
+[@morgan_winship_2014]
+
 - backdoor criterion non experimental way remove confounding [@barocas]

### 2021-05-20 presentation_video.Rmd
presentation

@@
 
-# Assumptions (1)
+# Foundations of SCMs (1)
 
@@
 
-# Assumptions (2)
+# SCM vs SEM
 
@@
 
-# Assumptions (3)
+# Assumptions
 
-First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
-Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
-This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
-The problem is that the equation is symmetric.
-The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
-This misconception has caused a lot of challenges.^[For more information see @pearl2009causality]
-As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
+- First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
+- Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
+- This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
+- The problem is that the equation is symmetric.
+- The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
+- This misconception has caused a lot of challenges.^[For more information see @pearl2009causality]
+- As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
 
-# Assumptions (4)
+# Assumptions in Causality
 
-**Independence**
+**Established Conditions:**
 
-*Definition:*
+Definition: **SUTVA**
+
+'The treatment that one unit receives does not change the effect of treatment for any other unit.'[@hardtrecht]
+
+Definition: **Consistency**
+
+The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' [@hardtrecht]
+
+Definition: **Ignorability**
+
+The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. [@hardtrecht]
+(perfect RCT)
+
+- First tow hold for SCM counterfactuals 
+- third not testable but can check via backdoor criterion in SCM
 
@@
 - Graphical visualization via directed acyclic graph
-
-# Comparative Causal Tools
-
-# Historical Development
-
-* Path Analysis -> SEM -> SCM
 
@@
 
+\footnotesize
+
  |Method         | CBN |  SCM
@@
 | Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
+|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific 
 |                |                                                                                                                      | 
@@
 |                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+
+\normalsize
 
@@
 
- -ML, BN and regression are at the lowest level in the causal hierarchy
+- ML, BN and regression are at the lowest level in the causal hierarchy
 - Prediction methods demand the least information and depend on association alone
@@
 
-The second query deals with interventions.
-Here we can use @pearl2009causality do-calculus.
-The do-calculus enables us to study the manipulation of parent nodes.
- There are various types of intervention.
-One example is **atomic intervention,** where we set a variable to a constant.
-In **policy intervention** we specify a different function for an equation.
-off-policy intervention models different intervention that is not in our historical data [@oberst2019counterfactual].
-Causal bayesian networks , Markov Decision Processes (MDP) and reinforcement learning model intervention.
+- The second query deals with interventions
+- Here we can use do-calculus
+- The do-calculus enables us to study the manipulation of parent nodes
+- There are various types of intervention
+- One example is **atomic intervention,** where we set a variable to a constant
+- In **policy intervention** we specify a different function for an equation
+- Off-policy intervention models different intervention that is not in our historical data
+- CBN , MDP and reinforcement learning model intervention.
 
@@
 
-# Graphical Illustration
+# Graphical Tools
+
+- Nodes 
+- Edges
+- Parents/Ancestors
+-
+
+# Graphical Illustration - Probabilisitic Model
 
@@
 ```
+
+# Graphical Illustration - Structural Causal Model
 
@@
 - Regular Time Specification is also more vague 
-- To accomodate that issue, research on differential equation based SCMs started
+- To accommodate that issue, research on differential equation based SCMs started
+
+# Graphical Overview
+
+\tiny
+$$
+\begin{array}{|l|l|l|l|l|l|}
+\hline \text { model } & \begin{array}{l}
+\text { predict in IID } \\
+\text { setting }
+\end{array} & \begin{array}{l}
+\text { predict under } \\
+\text { changing } \\
+\text { distributions / } \\
+\text { interventions }
+\end{array} & \begin{array}{l}
+\text { answer } \\
+\text { counter- } \\
+\text { factual } \\
+\text { questions }
+\end{array} & \begin{array}{l}
+\text { obtain } \\
+\text { physical } \\
+\text { insight }
+\end{array} & \begin{array}{l}
+\text { automatically } \\
+\text { learn from } \\
+\text { data }
+\end{array} \\
+\hline \begin{array}{l}
+\text { mechanistic } \\
+\text { model } \\
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & ? \\
+\hline \begin{array}{l}
+\text { structural } \\
+\text { causal model }
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{Y} ? ? \\
+\hline \begin{array}{l}
+\text { causal } \\
+\text { graphical } \\
+\text { model }
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{Y} ? \\
+\hline \begin{array}{l}
+\text { statistical } \\
+\text { model }
+\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} & \mathrm{Y} \\
+\hline
+\end{array}
+$$
+\normalsize
 

### 2021-05-21 draft.Rmd
s

--- 2021-05-17 draft.Rmd
@@
 bibliography: [ref.bib]
+csl: ieee.csl
 nocite: '@*'

### 2021-05-21 presentation_video.Rmd

@@
 - Associations are not enough for high stake settings 
-- In disciplines like psychology or economics people are less interested in associational learning
 - We want causation and not correlation
@@
   * Assumptions 
-  * Comparative Causal Tools 
-  * Historical Development 
+  * Comparing Causal Tools 
 * Pearl's Causal Hierachy 
@@
 - The second query deals with interventions
-- Here we can use do-calculus
+- Mathematical Tool: do-calculus
 - The do-calculus enables us to study the manipulation of parent nodes

### 2021-05-21 presentation_video.Rmd
xyz

@@
 ---
-title: "A Gentle Introduction to Structural Causal Models"
+title: "Introduction to Structural Causal Models"
 author: "Daniel Saggau"
@@
     theme: Madrid
-
+header-includes:
+- \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
+- \usecolortheme[named=UBCblue]{structure}
 ---
@@
 
+# Example Intevention
+
+```{r}
+
+```
+
 # Counterfactuals
@@
 - Edges
-- Parents/Ancestors
--
+- Parents/Ancestors/Descendents
+- (Missing) Arrows
 

### 2021-05-21 presentation_video.Rmd
presentation

@@
 
-# Example Intevention
+# Example Intevention (1)
+**Atomic Intervention**:
 
-```{r}
+- by replacing function with a constant
 
-```
+$$C_1:= f_{c_1}(p,q) \rightarrow C_1:=600$$
+$$C_2:= f_{c_2}(a,b)$$
+$$E:=f_E(C_1,C_2,N_E)$$
+
+# Example Intevention (2)
+
+**Policy Intervention:**
+
+- Intervention by replacing function with a different conditional probability
+
+$$C_1:= f_{c_1}(p,q) \rightarrow C_1:=f(\pi)$$
+$$C_2:= f_{c_2}(a,b)$$
+$$E:=f_E(C_1,C_2,N_E)$$
 
@@
 
+- No need to specify exact parametric shape
+- highlight colliders 
+- estimation back door criterion 
+-
 - Nodes 
@@
 - (Missing) Arrows
+
+# Difference Path Diagramm and Pearls Causal Diagrams
 

### 2021-05-21 presentation_video.Rmd
More stuff
presentation

@@
 
-# Introduction (1)
-
-- Machine learning has provided many insights into different problems 
-- One issue is the consideration of 'What are we actually predicting'?
-- Mainstream tools are build on association-based learning 
-- Associations are not enough for high stake settings 
-- We want causation and not correlation
-
-# Introduction (2)
-
-Causal assumptions differentiate causal models from association learning methods.
+# Introduction
 
@@
 | Propensity Score         | Spurious Correlation
+
+Table: [@pearl2012causal]
 
@@
 
-- System of equations 
+- System of equations with functions
 - Assignment equation ':=' rather than regular equation '=' 
 - Nonparametric SEM 
-- Functional form rather than using conditional probabilities
-- SCM = PO + Graphical Tools
+- This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations (symmetry)
+- The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
+- As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
 
-- Exogenous factors are part of the model specification
-
-# SCM vs SEM
-
-**Error terms**
-
+**Error:* **
 Regression: Omittable outside factor 
 SCM/SEM; Latent influential factor that is pivotal for the model specification but not observable
-
-Consists of graph and assignments: 
-Baseline:
 
@@
 
-# Assumptions
+# Assumptions in Causality
 
-- First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
-- Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
-- This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
-- The problem is that the equation is symmetric.
-- The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
-- This misconception has caused a lot of challenges.^[For more information see @pearl2009causality]
-- As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
+Independence:
 
-# Assumptions in Causality
+- noise terms independent
+- mechanisms independent (other variables invariant)(local changes)
+- change in distribution stems from change in mechanism
+
+- Causal Markov Condition allows to factorize 
+- Each Conditional is a causal mechanism
+
+Source: [@peters_elements_2017]
+
+# Assumptions in Causality (2)
 
@@
 * Differences in performance between BCN and SCM# Performance Evaluation
-
-# Implications
-
-\footnotesize
-
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific 
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
-
-\normalsize
 
@@
 
-$$C_1:= f_{c_1}(p,q) \rightarrow C_1:=600$$
-$$C_2:= f_{c_2}(a,b)$$
+$$C_1:= f_{c_1}(p,N_1) \rightarrow C_1:=600$$
+$$C_2:= f_{c_2}(a,N_2)$$
 $$E:=f_E(C_1,C_2,N_E)$$
@@
 
-$$C_1:= f_{c_1}(p,q) \rightarrow C_1:=f(\pi)$$
-$$C_2:= f_{c_2}(a,b)$$
+$$C_1:= f_{c_1}(p,N_1) \rightarrow C_1:=f(\pi)$$
+$$C_2:= f_{c_2}(a,N_2)$$
 $$E:=f_E(C_1,C_2,N_E)$$
@@
 # Counterfactuals
+
+- missing data problem in PO framework
 
@@
 
+# Causal Modelling with Differential Equations
+
+Intial Value:
+
+$$
+\mathbf{x}\left(t_{0}\right)=\mathbf{x}_{0}
+$$
+
+$$
+\frac{d \mathbf{x}}{d t}=f(\mathbf{x}), \mathbf{x} \in \mathbb{R}^{d}
+$$
+
+$$
+\mathbf{x}(t+d t)=\mathbf{x}(t)+d t \cdot f(\mathbf{x}(t))
+$$
+
 # Graphical Overview
 
-\tiny
+\scriptsize
+
 $$
-\begin{array}{|l|l|l|l|l|l|}
+\begin{array}{|l|l|l|l|l|}
 \hline \text { model } & \begin{array}{l}
@@
 \text { interventions }
-\end{array} & \begin{array}{l}
+\end{array} & \begin{array}{c}
 \text { answer } \\
@@
 \text { insight }
-\end{array} & \begin{array}{l}
-\text { automatically } \\
-\text { learn from } \\
-\text { data }
 \end{array} \\
@@
 \text { mechanistic } \\
-\text { model } \\
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & ? \\
+\text { model }
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} \\
 \hline \begin{array}{l}
@@
 \text { causal model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{Y} ? ? \\
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} \\
 \hline \begin{array}{l}
@@
 \text { model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{Y} ? \\
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} \\
 \hline \begin{array}{l}
@@
 \text { model }
-\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} & \mathrm{Y} \\
+\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} \\
 \hline
@@
 $$
+Source: [@peters_elements_2017]
 \normalsize
 
+# Extensions
+
+- Mediation Analysis 
+- PO-Framework
+- Causal Algorithms
+- Causal Constraints Model
+
+# Conclusions
+
 # References
+
+# Appendix

### 2021-05-21 presentation_video.Rmd
graphical models

@@
 - Graphical visualization via directed acyclic graph
+- Example: SCM for fairness in dynamical system [@creager] : credit loan approval, time allocation, college admission
 
@@
 
+- Nodes -> Variables (endogenous/exogenous)
+- Edges -> relationship (equations)
+- Parents/Ancestors/Descendents
 - No need to specify exact parametric shape
@@
 - estimation back door criterion 
--
-- Nodes 
-- Edges
-- Parents/Ancestors/Descendents
-- (Missing) Arrows
-
-# Difference Path Diagramm and Pearls Causal Diagrams
+- Test theoretical model structure via causal algorithms to detect structure in data (IC/PC Algo.)
 
@@
 
+- Assosicational learning is easy to model because of lower information neccessary
+- but not always appropiate in high stake settings 
+- SCM as expressive simulator for causal modelling outperforms many other models
+- computational advantage casting causal model as system of assignment equations 
+- Enables modelling of higher order concepts like counterfactuals 
+- Extensions through differential equations for concise modelling of time
+
+\newpage 
 # References
 
+<div id="refs"></div>
+\newpage
+
 # Appendix
+
+\footnotesize
+
+ |Method         | CBN |  SCM
+|----------------|---------------------------|------------------------|
+| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
+|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
+|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
+|                |                                                                                                                      | 
+| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
+|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific 
+|                |                                                                                                                      | 
+|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
+|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+
+\normalsize

### 2021-05-21 presentation_video.Rmd
video

@@
 bibliography: [ref.bib]
+csl: ieee.csl
 output: 
@@
   * Counterfactuals 
-  * Implications 
 * Graphical Models 
-  * Features 
-  * Implications
 * Causality and Time
 
-# Foundations of SCMs (1)
+# Foundations of SCMs
 
+- Nonparametric SEM 
 - System of equations with functions
-- Assignment equation ':=' rather than regular equation '=' 
-- Nonparametric SEM 
-- This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations (symmetry)
-- The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
+- Assignments ':=' (non-symmetric) instead of algebraic equation '=' (symmetry)
 - As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
-
-**Error:* **
-Regression: Omittable outside factor 
-SCM/SEM; Latent influential factor that is pivotal for the model specification but not observable
+**Error in Regression:** Omittable outside factor 
+**Error in SCM/SEM:** Latent (influential) factor that is pivotal for the model
 
@@
 
-# Assumptions in Causality
+# Assumptions in Causality (1)
 
@@
 
-- noise terms independent
-- mechanisms independent (other variables invariant)(local changes)
-- change in distribution stems from change in mechanism
-
-- Causal Markov Condition allows to factorize 
-- Each Conditional is a causal mechanism
-
-Source: [@peters_elements_2017]
+- Noise terms independent $(N_c,N_e)$
+- Mechanisms independent (other variables invariant)(local changes)
+- Change in distribution stems from change in mechanism 
+- Causal Markov Condition
 
@@
 
-**Established Conditions:**
+ **SUTVA**; 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
 
-Definition: **SUTVA**
+**Consistency**: The outcome Y agrees with the potential outcome corresponding to the treatment indicator.'
 
-'The treatment that one unit receives does not change the effect of treatment for any other unit.'[@hardtrecht]
-
-Definition: **Consistency**
-
-The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' [@hardtrecht]
-
-Definition: **Ignorability**
-
-The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. [@hardtrecht]
-(perfect RCT)
+**Ignorability**: The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. (perfect RCT)
 
 - First tow hold for SCM counterfactuals 
-- third not testable but can check via backdoor criterion in SCM
+- third not testable but can check via backdoor criterion in SCM [@hardtrecht]
 
@@
 * Differences in performance between BCN and SCM# Performance Evaluation
+
+# Table of Contents
+
+* Foundations of SCMs 
+  * Assumptions 
+  * Comparing Causal Tools 
+* **Pearl's Causal Hierachy**
+  * Prediction 
+  * Intervention 
+  * Counterfactuals 
+* Graphical Models 
+* Causality and Time
 
@@
 - Prediction methods demand the least information and depend on association alone
-- Associational methods ignore external changes outside of our data.
-- intervention distribution has information on these external changes.
-- intervention distribution is only defined in high level causal methods.
+- Association-based methods ignore external changes outside of our data
+- Intervention distribution needed for higher level information
 
@@
 
-- The second query deals with interventions
 - Mathematical Tool: do-calculus
 - The do-calculus enables us to study the manipulation of parent nodes
-- There are various types of intervention
-- One example is **atomic intervention,** where we set a variable to a constant
-- In **policy intervention** we specify a different function for an equation
-- Off-policy intervention models different intervention that is not in our historical data
+- **Atomic intervention:** where we set a variable to a constant
+- **Policy intervention:** we specify a different function for an equation
 - CBN , MDP and reinforcement learning model intervention.
@@
 
-- by replacing function with a constant
+- replacing function with a constant
 
-$$C_1:= f_{c_1}(p,N_1) \rightarrow C_1:=600$$
-$$C_2:= f_{c_2}(a,N_2)$$
+$$C_1:= f_{c_1}(p,N_{c_1}) \rightarrow C_1:=600$$
+$$C_2:= f_{c_2}(a,N_{c_2})$$
 $$E:=f_E(C_1,C_2,N_E)$$
@@
 
-- Intervention by replacing function with a different conditional probability
+- replacing function with a different conditional probability
 
-$$C_1:= f_{c_1}(p,N_1) \rightarrow C_1:=f(\pi)$$
-$$C_2:= f_{c_2}(a,N_2)$$
+$$C_1:= f_{c_1}(p,N_{c_1}) \rightarrow C_1:=f(\pi)$$
+$$C_2:= f_{c_2}(a,N_{c_2})$$
 $$E:=f_E(C_1,C_2,N_E)$$
@@
 
+# Table of Contents
+
+* Foundations of SCMs 
+  * Assumptions 
+  * Comparing Causal Tools 
+* Pearl's Causal Hierachy
+  * Prediction 
+  * Intervention 
+  * Counterfactuals 
+* **Graphical Models**
+* Causality and Time
+
 # Graphical Tools
@@
 - No need to specify exact parametric shape
-- highlight colliders 
-- estimation back door criterion 
+- Highlight colliders 
+- Estimation back door criterion 
 - Test theoretical model structure via causal algorithms to detect structure in data (IC/PC Algo.)
@@
 
+# Table of Contents
+
+* Foundations of SCMs 
+  * Assumptions 
+  * Comparing Causal Tools 
+* Pearl's Causal Hierachy
+  * Prediction 
+  * Intervention 
+  * Counterfactuals 
+* Graphical Models
+* **Causality and Time**
+
 # Causality and Time
@@
 
+Derivative of function x with respect to time t:
+
 $$
@@
 $$
+Value of Function at time t + dt:
 
@@
 $$
+
+- dependence on prior time point and change in time contribute to the value at time point
 
@@
 - Causal Algorithms
+- IV-Estimation
 - Causal Constraints Model 
+- Causal Trees
 
-# Conclusions
+# Concluding Remarks
 
@@
 - but not always appropiate in high stake settings 
-- SCM as expressive simulator for causal modelling outperforms many other models
-- computational advantage casting causal model as system of assignment equations 
+- SCM as simulator for causal modelling, entailing a lot of information (DGP, intervention distribution)
+- Note: *"Garbage in, Garbage out"*
+- computational advantage casting causal model as system of assignments 
 - Enables modelling of higher order concepts like counterfactuals 
@@
 # References
+\tiny
 
 <div id="refs"></div>
+\normalsize
 \newpage

### 2021-05-21 presentation_video.Rmd
SCM

@@
 bibliography: [ref.bib]
-csl: ieee.csl
+#csl: ieee.csl
 output: 
@@
 - System of equations with functions
-- Assignments ':=' (non-symmetric) instead of algebraic equation '=' (symmetry)
-- As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
+- **Assignments** ':=' (non-symmetric) 
+- algebraic equation '=' (symmetry)
 **Error in Regression:** Omittable outside factor 
-**Error in SCM/SEM:** Latent (influential) factor that is pivotal for the model
+**Error in SCM:** Latent (influential) factor that is pivotal for the model
 
@@
 
- **SUTVA**; 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
+*  **SUTVA** 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
+* **Consistency** The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' 
+* **Ignorability** The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. (perfect RCT)
 
-**Consistency**: The outcome Y agrees with the potential outcome corresponding to the treatment indicator.'
-
-**Ignorability**: The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. (perfect RCT)
-
-- First tow hold for SCM counterfactuals 
+- First tow hold for Counterfactuals in SCM
 - third not testable but can check via backdoor criterion in SCM [@hardtrecht]
@@
 
-- conflict whether to use graphs or not 
-* A SEM is a parametric specification used in applied sciences (parameters contested)
-* A Bayesian causal network is another popular causal model using conditional probabilities and NO functions 
-* Differences in performance between BCN and SCM# Performance Evaluation
+- Conflict whether to use **graphs** or not and to use **parameters** in SEM
+* SEM is a parametric model used in applied sciences 
+* A Bayesian causal network (BCN) use conditional probabilities instead of functions 
+* PO is mostly equivalent but has not graphical framework
+* SCM and PO are only frameworks that allow for counterfactuals
 
@@
 
-# Causality and Time
-
-- Time in Physical Sciences: Mechanical and exact 
-- Time in Social Sciences: Often Vague 
-- Regular Time Specification is also more vague 
-- To accommodate that issue, research on differential equation based SCMs started
-
 # Causal Modelling with Differential Equations
 
-Intial Value:
+- Time in Social Sciences: Often Vague 
+- Time in Physical Sciences: Mechanical via **Differential equations** 
+- dependence on prior time point and change in time contribute to the value at time point
 
+Initial Value: 
 $$
@@
 Derivative of function x with respect to time t: 
-
 $$
@@
 $$
+
 Value of Function at time t + dt: 
-
 $$
@@
 $$
-
-- dependence on prior time point and change in time contribute to the value at time point
 
@@
 \hline \text { model } & \begin{array}{l}
-\text { predict in IID } \\
-\text { setting }
+\text { IID setting }
 \end{array} & \begin{array}{l}
-\text { predict under } \\
 \text { changing } \\
-\text { distributions / } \\
-\text { interventions }
+\text { distributions } 
 \end{array} & \begin{array}{c}
-\text { answer } \\
 \text { counter- } \\
@@
 \end{array} & \begin{array}{l}
-\text { obtain } \\
 \text { physical } \\

### 2021-05-21 draft.Rmd
graphical models
pres

@@
 Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality].
-Summarized, *'they are simulators that generate a distribution based noise variables (independent) and a unique joint distribution...'* [@hardtrecht].
-- noise variables helo generate joint distribution [@barocas].
+Summarized, *'they are simulators that generate a distribution based on noise variables (independent) and a unique joint distribution...'* [@hardtrecht].
+- noise variables help generate joint distribution [@barocas].
 SCMs are non-parametric structural equations models with added features.

### 2021-05-21 presentation_video.Rmd
pres

@@
 
-# Assumptions in Causality (2)
+# Assumptions under which causal inference is possible (2)
 
@@
 
-- Flexible simulations for higher order problems (intervention, counterfactual)
+- Flexible simulations for higher order problems (intervention, counterfactual) based on elementary noise variables
 - Graphical visualization via directed acyclic graph
@@
 * SCM and PO are only frameworks that allow for counterfactuals
+- Every SCM has the same information as respective probabilisitc model + more (DGP)
+- SCMs can derive potential outcomes but PO cannot derive SCMs
 
@@
 - Mathematical Tool: do-calculus
-- The do-calculus enables us to study the manipulation of parent nodes
+- The do-calculus enables us to study the manipulation of parent nodes (direct effect)
 - **Atomic intervention:** where we set a variable to a constant
@@
 - Edges -> relationship (equations)
-- Parents/Ancestors/Descendents
+- Parents/Ancestors/Descendants
 - No need to specify exact parametric shape
@@
 - Test theoretical model structure via causal algorithms to detect structure in data (IC/PC Algo.) 
+- More information: [@morgan_winship_2014]
 
@@
 
-- Assosicational learning is easy to model because of lower information neccessary
-- but not always appropiate in high stake settings 
-- SCM as simulator for causal modelling, entailing a lot of information (DGP, intervention distribution)
+- Association-based learning is easy to model because of lower information neccessary
+- but not always appropriate in high stake settings 
+- SCM as simulator for causal modelling based on noise variables
+- SCMS entail a lot of information (DGP, intervention distribution)
 - Note: *"Garbage in, Garbage out"*
-- computational advantage casting causal model as system of assignments 
+- computational advantage casting causal model as assignments 
 - Enables modelling of higher order concepts like counterfactuals 

### 2021-05-21 presentation_video.Rmd
stuff

@@
 - algebraic equation '=' (symmetry)
-**Error in Regression:** Omittable outside factor 
-**Error in SCM:** Latent (influential) factor that is pivotal for the model
+- **Error in Regression:** Omittable outside factor 
+- **Error in SCM:** Latent (influential) factor that is pivotal for the model
 

### 2021-05-21 presentation_video.Rmd
lecture

@@
 author: "Daniel Saggau"
-date: "5/11/2021"
+date: "May 2021"
 citation_package: --biblatex
@@
 * SCM and PO are only frameworks that allow for counterfactuals
-- Every SCM has the same information as respective probabilisitc model + more (DGP)
+- Every SCM has the same information as respective probabilistic model + more (DGP)
 - SCMs can derive potential outcomes but PO cannot derive SCMs
@@
   * Comparing Causal Tools 
-* Pearl's Causal Hierachy
+* Pearl's Causal Hierarchy
   * Prediction 

### 2021-05-21 presentation_video.Rmd
minoor
presentation

@@
 
+# Seminar Topics
+
 # Table of Contents
@@
 
+- Flexible simulations for higher order problems (intervention, counterfactual) based on elementary noise variables
+- Graphical visualization via directed acyclic graph
 - Nonparametric SEM 
@@
 
-source: @peters_elements_2017
+Source: @peters_elements_2017
 
-# Assumptions in Causality (1)
+# Assumptions (1)
 
@@
 - Causal Markov Condition
-
-# Assumptions under which causal inference is possible (2)
-
-*  **SUTVA** 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
-* **Consistency** The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' 
-* **Ignorability** The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. (perfect RCT)
-
-- First tow hold for Counterfactuals in SCM
-- third not testable but can check via backdoor criterion in SCM [@hardtrecht]
-
-# SCM Applications:
-
-- Flexible simulations for higher order problems (intervention, counterfactual) based on elementary noise variables
-- Graphical visualization via directed acyclic graph
-- Example: SCM for fairness in dynamical system [@creager] : credit loan approval, time allocation, college admission
 
@@
 
-# Prediction
-
-- ML, BN and regression are at the lowest level in the causal hierarchy
-- Prediction methods demand the least information and depend on association alone
-- Association-based methods ignore external changes outside of our data
-- Intervention distribution needed for higher level information
-
 # Intervention
@@
 
-- missing data problem in PO framework
-
-Process is described as follows:
+- Missing data problem in PO framework 
+- SCM Process is described as follows:
 
@@
 (c) Prediction: Compute $(Y = y)$
+
+# Assumptions
+
+*  **SUTVA** 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
+* **Consistency** The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' 
+* **Ignorability** The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. (perfect RCT)
+
+- First tow hold for Counterfactuals in SCM
+- third not testable but can check via backdoor criterion in SCM [@hardtrecht]
 
@@
 - No need to specify exact parametric shape
-- Highlight colliders 
+- Focus on acyclical structures (DAG)
 - Estimation back door criterion 
@@
 - More information: [@morgan_winship_2014]
+
+# Undirected/Semidirected Graph
+
+```{tikz, well, fig.cap ="Undirected graph", fig.align="center", echo=F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+  \begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=7pt,
+    node distance=3cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\tikzstyle{line} = [thick,-,>=stealth, line width=0.5pt]
+\node[my node] (A){A};
+\node[my node,right of=A](C){C};
+\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (B) {B};
+\draw[line] (A) -- (B);
+\draw[line] (C) -- (A);
+\draw[line] (B) -- (C);
+\end{tikzpicture}
+```
+
+```{tikz, wela, fig.cap ="Semi-directed graph", fig.align="center", echo=F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=7pt,
+    node distance=3cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\tikzstyle{line} = [thick,-,>=stealth, line width=0.5pt]
+\node[my node] (A){A};
+\node[my node,right of=A](C){C};
+\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (B) {B};
+\draw[line] (A) -- (B);
+\draw[line] (C) -- (A);
+\draw[->] (B) -- (C);
+\end{tikzpicture}
+```
 
@@
 \node[my node,right =1 cm of A](C){C};
-\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
-\draw[->] (A) -- (T);
+\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (E) {E};
+\draw[->] (A) -- (E);
 \draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
+\draw[->] (E) -- (C);
 \end{tikzpicture}
@@
 \node[latent,left of=A](UA){$U_A$};
-\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
-\node[my node,right = 1 cm of T, fill=gray!30](C){C};
-\node[latent,left = 1cm of T](UT){$U_T$};
+\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (E) {E};
+\node[my node,right = 1 cm of E, fill=gray!30](C){C};
+\node[latent,left = 1cm of E](UE){$U_E$};
 \node[latent,right = 1cm of A](UC){$U_C$};
-\draw[->] (A) -- (T);
+\draw[->] (A) -- (E);
 \draw[->] (UA) -- (A);
-\draw[->] (UT) -- (T);
+\draw[->] (UE) -- (E);
 \draw[->] (UC) -- (C);
 \draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
+\draw[->] (E) -- (C);
 \end{tikzpicture}
@@
 
-# Extensions
+# Does and Do nots
 
-- Mediation Analysis 
-- PO-Framework
-- Causal Algorithms
-- IV-Estimation
-- Causal Constraints Model 
-- Causal Trees
+**If you...**
+
+* ...only have conditional probabilities 
+* ...are only interested in association 
+* ...have very limited knowledge of your actual underlying causal structure
+
+$\rightarrow$ you should probably go for a probabilistic model, association based model or a model based on conditional probabilities
+
+**But if you**
+
+* ... have a good understanding of underlying latent factors 
+* ... have a lot of information on your underlying relationship in your data 
+* ... are dealing with high stake settings 
+* ... want to simulate different interventions before implementing them
+
+$\rightarrow$ you should probably go for a SCM
 

### 2021-05-21 presentation_video.Rmd
n

@@
 
-# Seminar Topics
-
 # Table of Contents
@@
 
-# Foundations of SCMs
+# Seminar Topics
 
-- Flexible simulations for higher order problems (intervention, counterfactual) based on elementary noise variables
+* Topic 1: **Structural Causal Models**
+* Topic 2: Linking d-separation with (conditional) independence & Causal Structure Learning
+* Topic 3: do-calculus: estimating the effect of interventions & counterfactuals 
+* Topic 4: Rubin’s Potential Outcome Framework & Causal Trees
+* Topic 5: Algorithmic Recourse 
+* Topic 6: Causal Interpretation of Black-Box Models
+
+# Foundations of SCMs (1)
+
+- Flexible simulations for higher order problems (intervention, counterfactual) 
+- based on (elementary) 'noise'/latent variables
 - Graphical visualization via directed acyclic graph
@@
 - System of equations with functions
+
+# Foundations of SCMs (2)
+
 - **Assignments** ':=' (non-symmetric) 
@@
 
-# Assumptions (1)
-
-Independence:
+# Assumptions - Independence
 
@@
 
-# Fundamental Differences (1)
+# Fundamental Differences
 
@@
 - First tow hold for Counterfactuals in SCM
-- third not testable but can check via backdoor criterion in SCM [@hardtrecht]
+- third not testable but can check via backdoor criterion in SCM
+- Source: [@hardtrecht]
 
@@
 
-# Does and Do nots
-
-**If you...**
-
-* ...only have conditional probabilities 
-* ...are only interested in association 
-* ...have very limited knowledge of your actual underlying causal structure
-
-$\rightarrow$ you should probably go for a probabilistic model, association based model or a model based on conditional probabilities
-
-**But if you**
-
-* ... have a good understanding of underlying latent factors 
-* ... have a lot of information on your underlying relationship in your data 
-* ... are dealing with high stake settings 
-* ... want to simulate different interventions before implementing them
-
-$\rightarrow$ you should probably go for a SCM
-
 # Concluding Remarks
@@
 - Extensions through differential equations for concise modelling of time
+
+# Dos and Do nots
+
+**If you...**
+
+* ...only have conditional probabilities 
+* ...are only interested in association 
+* ...have very limited knowledge of your actual underlying structure
+
+$\rightarrow$ you should probably go for a probabilistic model, association based model or a model based on conditional probabilities
+
+**But if you**
+
+* ... have an understanding of underlying relationships in the data
+* ... are dealing with high stake settings 
+* ... want to simulate different interventions
+
+$\rightarrow$ you should probably go for a SCM
 
@@
 \normalsize
+
+# Questions:
+
+Thank you so much for your time!
+
+If you have any questions contact me via:
+
+* Mattermost: daniel.saggau
+* Email: daniel.saggau@campus.lmu.de

### 2021-05-21 presentation_video.Rmd
more

@@
 
-# Fundamental Differences
+# Differences
+
+**SEM:**
 
 - Conflict whether to use **graphs** or not and to use **parameters** in SEM
-* SEM is a parametric model used in applied sciences 
-* A Bayesian causal network (BCN) use conditional probabilities instead of functions 
-* PO is mostly equivalent but has not graphical framework
-* SCM and PO are only frameworks that allow for counterfactuals
+- SEM is a parametric model used in applied sciences
+
+**BCN**:
+
+- A Bayesian causal network (BCN) use conditional probabilities instead of functions 
 - Every SCM has the same information as respective probabilistic model + more (DGP)
+- Computationally less efficient and cannot cast counterfactuals
+
+**Potential Outcome Framework:**
+
+- PO is mostly equivalent but has not graphical framework
 - SCMs can derive potential outcomes but PO cannot derive SCMs
@@
 - SCM as simulator for causal modelling based on noise variables
-- SCMS entail a lot of information (DGP, intervention distribution)
-- Note: *"Garbage in, Garbage out"*
-- computational advantage casting causal model as assignments 
+- SCMs entail a lot of information (DGP, intervention distribution)
+- BUT: **"Garbage in, Garbage out"**
+- Computational advantage casting causal model as assignments 
 - Enables modelling of higher order concepts like counterfactuals 
@@
 
-* ...only have conditional probabilities 
-* ...are only interested in association 
-* ...have very limited knowledge of your actual underlying structure
+* ..only have conditional probabilities 
+* ..are only interested in association 
+* ..have very limited knowledge of your actual underlying structure
 
@@
 
-* ... have an understanding of underlying relationships in the data
-* ... are dealing with high stake settings 
-* ... want to simulate different interventions
+* ..have an understanding of underlying relationships in the data
+* ..are dealing with high stake settings 
+* ..want to simulate different interventions
 
-$\rightarrow$ you should probably go for a SCM
+$\rightarrow$ you could probably go for a SCM
 

### 2021-05-24 questions.Rmd
Q&A

--- 2021-05-21 presentation_video.Rmd
@@
 ---
-title: "Introduction to Structural Causal Models"
+title: "Questions Presentation - Structural Causal Models"
 author: "Daniel Saggau"
-date: "May 2021"
-citation_package: --biblatex
-bibliography: [ref.bib]
-#csl: ieee.csl
-output: 
-  beamer_presentation:
-    theme: Madrid
-header-includes:
-- \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
-- \usecolortheme[named=UBCblue]{structure}
+output: pdf_document
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = TRUE)
 ```
 
-# Motivation
-
-```{r, out.width= "80%", out.height = "80%", fig.align='center'}
-library(knitr)
-setwd("~/Desktop/SOSE2021/causal_ml")
-include_graphics('./images/correlation.jpg')
-```
-
-# Introduction
-
- | Association-based Concepts |  Causal Concepts
-|--------------------------|------------------------|
-| Correlation              | Randomization  
-| Regression               | Confounding
-| Conditional Independence | Disturbance
-| Likelihood               | Error Terms      
-| Odds Ratio               | Structural Coefficients       
-| Propensity Score         | Spurious Correlation
-
-Table: [@pearl2012causal]
-
-# Seminar Topics
-
-# Table of Contents
-
-* Foundations of SCMs 
-  * Assumptions 
-  * Comparing Causal Tools 
-* Pearl's Causal Hierachy 
-  * Prediction 
-  * Intervention 
-  * Counterfactuals 
-* Graphical Models 
-* Causality and Time
-
-# Foundations of SCMs
-
-- Flexible simulations for higher order problems (intervention, counterfactual) based on elementary noise variables
-- Graphical visualization via directed acyclic graph
-- Nonparametric SEM 
-- System of equations with functions
-- **Assignments** ':=' (non-symmetric) 
-- algebraic equation '=' (symmetry)
-- **Error in Regression:** Omittable outside factor 
-- **Error in SCM:** Latent (influential) factor that is pivotal for the model
-
-$$C:= N_c$$
-$$E:=f_E(C,N_E)$$
-
-Source: @peters_elements_2017
-
-# Assumptions (1)
-
-Independence:
-
-- Noise terms independent $(N_c,N_e)$
-- Mechanisms independent (other variables invariant)(local changes)
-- Change in distribution stems from change in mechanism 
-- Causal Markov Condition
-
-# Fundamental Differences (1)
-
-- Conflict whether to use **graphs** or not and to use **parameters** in SEM
-* SEM is a parametric model used in applied sciences 
-* A Bayesian causal network (BCN) use conditional probabilities instead of functions 
-* PO is mostly equivalent but has not graphical framework
-* SCM and PO are only frameworks that allow for counterfactuals
-- Every SCM has the same information as respective probabilistic model + more (DGP)
-- SCMs can derive potential outcomes but PO cannot derive SCMs
-
-# Table of Contents
-
-* Foundations of SCMs 
-  * Assumptions 
-  * Comparing Causal Tools 
-* **Pearl's Causal Hierachy**
-  * Prediction 
-  * Intervention 
-  * Counterfactuals 
-* Graphical Models 
-* Causality and Time
-
-# Pearls Causal Hierachy
+1) What kind of models could the following DAGs be? Briefly justify your choices.
 
 \footnotesize
- |Method          | Action |  Example | Usage | 
-|------------------|-------------|--------------------|-------------------|
-| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
-| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
-| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
-
-Table: Pearls Hierachy of Causation (2009)
-
-\normalsize
-
-# Intervention
-
-- Mathematical Tool: do-calculus
-- The do-calculus enables us to study the manipulation of parent nodes (direct effect)
-- **Atomic intervention:** where we set a variable to a constant
-- **Policy intervention:** we specify a different function for an equation
-- CBN , MDP and reinforcement learning model intervention.
-
-# Example Intevention (1)
-**Atomic Intervention**:
-
-- replacing function with a constant
-
-$$C_1:= f_{c_1}(p,N_{c_1}) \rightarrow C_1:=600$$
-$$C_2:= f_{c_2}(a,N_{c_2})$$
-$$E:=f_E(C_1,C_2,N_E)$$
-
-# Example Intevention (2)
-
-**Policy Intervention:**
-
-- replacing function with a different conditional probability
-
-$$C_1:= f_{c_1}(p,N_{c_1}) \rightarrow C_1:=f(\pi)$$
-$$C_2:= f_{c_2}(a,N_{c_2})$$
-$$E:=f_E(C_1,C_2,N_E)$$
-
-# Counterfactuals
-
-- Missing data problem in PO framework 
-- SCM Process is described as follows:
-
-(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
-(b) Action: Exchange $(X = x)$ 
-(c) Prediction: Compute $(Y = y)$
-
-# Assumptions
-
-*  **SUTVA** 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
-* **Consistency** The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' 
-* **Ignorability** The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. (perfect RCT)
-
-- First tow hold for Counterfactuals in SCM
-- third not testable but can check via backdoor criterion in SCM [@hardtrecht]
-
-# Table of Contents
-
-* Foundations of SCMs 
-  * Assumptions 
-  * Comparing Causal Tools 
-* Pearl's Causal Hierachy
-  * Prediction 
-  * Intervention 
-  * Counterfactuals 
-* **Graphical Models**
-* Causality and Time
-
-# Graphical Tools
-
-- Nodes -> Variables (endogenous/exogenous)
-- Edges -> relationship (equations)
-- Parents/Ancestors/Descendants
-- No need to specify exact parametric shape
-- Focus on acyclical structures (DAG)
-- Estimation back door criterion 
-- Test theoretical model structure via causal algorithms to detect structure in data (IC/PC Algo.) 
-- More information: [@morgan_winship_2014]
-
-# Undirected/Semidirected Graph
-
-```{tikz, well, fig.cap ="Undirected graph", fig.align="center", echo=F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-  \begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{line} = [thick,-,>=stealth, line width=0.5pt]
-\node[my node] (A){A};
-\node[my node,right of=A](C){C};
-\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (B) {B};
-\draw[line] (A) -- (B);
-\draw[line] (C) -- (A);
-\draw[line] (B) -- (C);
-\end{tikzpicture}
-```
-
-```{tikz, wela, fig.cap ="Semi-directed graph", fig.align="center", echo=F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{line} = [thick,-,>=stealth, line width=0.5pt]
-\node[my node] (A){A};
-\node[my node,right of=A](C){C};
-\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (B) {B};
-\draw[line] (A) -- (B);
-\draw[line] (C) -- (A);
-\draw[->] (B) -- (C);
-\end{tikzpicture}
-```
-
-# Graphical Illustration - Probabilisitic Model
-
-```{tikz,fig.align="center",fig.cap="Probabilistic Model", echo =F}
+```{tikz,fig.align="center",fig.cap="Model 1", echo =F}
 \usetikzlibrary{calc}
@@
 
-# Graphical Illustration - Structural Causal Model
-
-```{tikz,fig.cap="Structural Causal Model", fig.align="center", echo =F}
+```{tikz,fig.cap="Model 2", fig.align="center", echo =F}
 \usetikzlibrary{calc}
@@
 ```
-
-# Table of Contents
-
-* Foundations of SCMs 
-  * Assumptions 
-  * Comparing Causal Tools 
-* Pearl's Causal Hierarchy
-  * Prediction 
-  * Intervention 
-  * Counterfactuals 
-* Graphical Models
-* **Causality and Time**
-
-# Causal Modelling with Differential Equations
-
-- Time in Social Sciences: Often Vague 
-- Time in Physical Sciences: Mechanical via **Differential equations** 
-- dependence on prior time point and change in time contribute to the value at time point
-
-Initial Value: 
-$$
-\mathbf{x}\left(t_{0}\right)=\mathbf{x}_{0}
-$$
-
-Derivative of function x with respect to time t: 
-$$
-\frac{d \mathbf{x}}{d t}=f(\mathbf{x}), \mathbf{x} \in \mathbb{R}^{d}
-$$
-
-Value of Function at time t + dt: 
-$$
-\mathbf{x}(t+d t)=\mathbf{x}(t)+d t \cdot f(\mathbf{x}(t))
-$$
-
-# Graphical Overview
-
-\scriptsize
-
-$$
-\begin{array}{|l|l|l|l|l|}
-\hline \text { model } & \begin{array}{l}
-\text { IID setting }
-\end{array} & \begin{array}{l}
-\text { changing } \\
-\text { distributions } 
-\end{array} & \begin{array}{c}
-\text { counter- } \\
-\text { factual } \\
-\text { questions }
-\end{array} & \begin{array}{l}
-\text { physical } \\
-\text { insight }
-\end{array} \\
-\hline \begin{array}{l}
-\text { mechanistic } \\
-\text { model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} \\
-\hline \begin{array}{l}
-\text { structural } \\
-\text { causal model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} \\
-\hline \begin{array}{l}
-\text { causal } \\
-\text { graphical } \\
-\text { model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} \\
-\hline \begin{array}{l}
-\text { statistical } \\
-\text { model }
-\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} \\
-\hline
-\end{array}
-$$
-Source: [@peters_elements_2017]
 \normalsize
 
-# Does and Do nots
+2) True or False Questions: Circle the correct answer. Only one answer is correct.
 
-**If you...**
+a) The SCM is the same thing as the SEM. There are no differences. 
+b) We can derive any potential outcome model based on a structural causal model.
+c) The SCM model specification will always outperform other models by design.
+d) A Bayesian Causal Network can do anything a Structural Causal Network can do.
 
-* ...only have conditional probabilities 
-* ...are only interested in association 
-* ...have very limited knowledge of your actual underlying causal structure
+3) Write down the respective system of equations for model 2 in question 1.
 
-$\rightarrow$ you should probably go for a probabilistic model, association based model or a model based on conditional probabilities
-
-**But if you**
-
-* ... have a good understanding of underlying latent factors 
-* ... have a lot of information on your underlying relationship in your data 
-* ... are dealing with high stake settings 
-* ... want to simulate different interventions before implementing them
-
-$\rightarrow$ you should probably go for a SCM
-
-# Concluding Remarks
-
-- Association-based learning is easy to model because of lower information neccessary
-- but not always appropriate in high stake settings 
-- SCM as simulator for causal modelling based on noise variables
-- SCMS entail a lot of information (DGP, intervention distribution)
-- Note: *"Garbage in, Garbage out"*
-- computational advantage casting causal model as assignments 
-- Enables modelling of higher order concepts like counterfactuals 
-- Extensions through differential equations for concise modelling of time
-
-\newpage 
-# References
-\tiny
-
-<div id="refs"></div>
-\normalsize
-\newpage
-
-# Appendix
-
-\footnotesize
-
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific 
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
-
-\normalsize
+4) Name two arguments on why we care about SCMs. Also, give two arguments why SCMs might not always be the most suitable model choice.

### 2021-05-27 scm_short_presentation.Rmd
slides

--- 2021-05-21 presentation_video.Rmd
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
-```
-
-# Motivation
-
-```{r, out.width= "80%", out.height = "80%", fig.align='center'}
-library(knitr)
-setwd("~/Desktop/SOSE2021/causal_ml")
-include_graphics('./images/correlation.jpg')
-```
-
-# Introduction
-
- | Association-based Concepts |  Causal Concepts
-|--------------------------|------------------------|
-| Correlation              | Randomization  
-| Regression               | Confounding
-| Conditional Independence | Disturbance
-| Likelihood               | Error Terms      
-| Odds Ratio               | Structural Coefficients       
-| Propensity Score         | Spurious Correlation
-
-Table: [@pearl2012causal]
-
-# Seminar Topics
-
-# Table of Contents
+# Overview
 
@@
 
-# Foundations of SCMs
-
-- Flexible simulations for higher order problems (intervention, counterfactual) based on elementary noise variables
-- Graphical visualization via directed acyclic graph
-- Nonparametric SEM 
-- System of equations with functions
-- **Assignments** ':=' (non-symmetric) 
-- algebraic equation '=' (symmetry)
-- **Error in Regression:** Omittable outside factor 
-- **Error in SCM:** Latent (influential) factor that is pivotal for the model
-
-$$C:= N_c$$
-$$E:=f_E(C,N_E)$$
-
-Source: @peters_elements_2017
-
-# Assumptions (1)
-
-Independence:
-
-- Noise terms independent $(N_c,N_e)$
-- Mechanisms independent (other variables invariant)(local changes)
-- Change in distribution stems from change in mechanism 
-- Causal Markov Condition
-
-# Fundamental Differences (1)
-
-- Conflict whether to use **graphs** or not and to use **parameters** in SEM
-* SEM is a parametric model used in applied sciences 
-* A Bayesian causal network (BCN) use conditional probabilities instead of functions 
-* PO is mostly equivalent but has not graphical framework
-* SCM and PO are only frameworks that allow for counterfactuals
-- Every SCM has the same information as respective probabilistic model + more (DGP)
-- SCMs can derive potential outcomes but PO cannot derive SCMs
-
-# Table of Contents
-
-* Foundations of SCMs 
-  * Assumptions 
-  * Comparing Causal Tools 
-* **Pearl's Causal Hierachy**
-  * Prediction 
-  * Intervention 
-  * Counterfactuals 
-* Graphical Models 
-* Causality and Time
-
-# Pearls Causal Hierachy
-
-\footnotesize
- |Method          | Action |  Example | Usage | 
-|------------------|-------------|--------------------|-------------------|
-| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
-| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
-| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
-
-Table: Pearls Hierachy of Causation (2009)
-
-\normalsize
-
-# Intervention
-
-- Mathematical Tool: do-calculus
-- The do-calculus enables us to study the manipulation of parent nodes (direct effect)
-- **Atomic intervention:** where we set a variable to a constant
-- **Policy intervention:** we specify a different function for an equation
-- CBN , MDP and reinforcement learning model intervention.
-
-# Example Intevention (1)
-**Atomic Intervention**:
-
-- replacing function with a constant
-
-$$C_1:= f_{c_1}(p,N_{c_1}) \rightarrow C_1:=600$$
-$$C_2:= f_{c_2}(a,N_{c_2})$$
-$$E:=f_E(C_1,C_2,N_E)$$
-
-# Example Intevention (2)
-
-**Policy Intervention:**
-
-- replacing function with a different conditional probability
-
-$$C_1:= f_{c_1}(p,N_{c_1}) \rightarrow C_1:=f(\pi)$$
-$$C_2:= f_{c_2}(a,N_{c_2})$$
-$$E:=f_E(C_1,C_2,N_E)$$
-
-# Counterfactuals
-
-- Missing data problem in PO framework 
-- SCM Process is described as follows:
-
-(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
-(b) Action: Exchange $(X = x)$ 
-(c) Prediction: Compute $(Y = y)$
-
-# Assumptions
-
-*  **SUTVA** 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
-* **Consistency** The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' 
-* **Ignorability** The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. (perfect RCT)
-
-- First tow hold for Counterfactuals in SCM
-- third not testable but can check via backdoor criterion in SCM [@hardtrecht]
-
-# Table of Contents
-
-* Foundations of SCMs 
-  * Assumptions 
-  * Comparing Causal Tools 
-* Pearl's Causal Hierachy
-  * Prediction 
-  * Intervention 
-  * Counterfactuals 
-* **Graphical Models**
-* Causality and Time
-
-# Graphical Tools
-
-- Nodes -> Variables (endogenous/exogenous)
-- Edges -> relationship (equations)
-- Parents/Ancestors/Descendants
-- No need to specify exact parametric shape
-- Focus on acyclical structures (DAG)
-- Estimation back door criterion 
-- Test theoretical model structure via causal algorithms to detect structure in data (IC/PC Algo.) 
-- More information: [@morgan_winship_2014]
-
-# Undirected/Semidirected Graph
-
-```{tikz, well, fig.cap ="Undirected graph", fig.align="center", echo=F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-  \begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{line} = [thick,-,>=stealth, line width=0.5pt]
-\node[my node] (A){A};
-\node[my node,right of=A](C){C};
-\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (B) {B};
-\draw[line] (A) -- (B);
-\draw[line] (C) -- (A);
-\draw[line] (B) -- (C);
-\end{tikzpicture}
-```
-
-```{tikz, wela, fig.cap ="Semi-directed graph", fig.align="center", echo=F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{line} = [thick,-,>=stealth, line width=0.5pt]
-\node[my node] (A){A};
-\node[my node,right of=A](C){C};
-\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (B) {B};
-\draw[line] (A) -- (B);
-\draw[line] (C) -- (A);
-\draw[->] (B) -- (C);
-\end{tikzpicture}
-```
-
-# Graphical Illustration - Probabilisitic Model
-
-```{tikz,fig.align="center",fig.cap="Probabilistic Model", echo =F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\node[my node] (A){A};
-\node[my node,right =1 cm of A](C){C};
-\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (E) {E};
-\draw[->] (A) -- (E);
-\draw[->] (A) -- (C);
-\draw[->] (E) -- (C);
-\end{tikzpicture}
-```
-
-# Graphical Illustration - Structural Causal Model
+# Example
 
@@
 
-# Table of Contents
+# Pearl's Causal Hierachy
 
-* Foundations of SCMs 
-  * Assumptions 
-  * Comparing Causal Tools 
-* Pearl's Causal Hierarchy
-  * Prediction 
-  * Intervention 
-  * Counterfactuals 
-* Graphical Models
-* **Causality and Time**
+\footnotesize
+ |Method          | Action |  Example | Usage | 
+|------------------|-------------|--------------------|-------------------|
+| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
+| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
 
-# Causal Modelling with Differential Equations
+Table: Pearls Hierachy of Causation (2009)
 
-- Time in Social Sciences: Often Vague 
-- Time in Physical Sciences: Mechanical via **Differential equations** 
-- dependence on prior time point and change in time contribute to the value at time point
-
-Initial Value: 
-$$
-\mathbf{x}\left(t_{0}\right)=\mathbf{x}_{0}
-$$
-
-Derivative of function x with respect to time t: 
-$$
-\frac{d \mathbf{x}}{d t}=f(\mathbf{x}), \mathbf{x} \in \mathbb{R}^{d}
-$$
-
-Value of Function at time t + dt: 
-$$
-\mathbf{x}(t+d t)=\mathbf{x}(t)+d t \cdot f(\mathbf{x}(t))
-$$
-
-# Graphical Overview
-
-\scriptsize
-
-$$
-\begin{array}{|l|l|l|l|l|}
-\hline \text { model } & \begin{array}{l}
-\text { IID setting }
-\end{array} & \begin{array}{l}
-\text { changing } \\
-\text { distributions } 
-\end{array} & \begin{array}{c}
-\text { counter- } \\
-\text { factual } \\
-\text { questions }
-\end{array} & \begin{array}{l}
-\text { physical } \\
-\text { insight }
-\end{array} \\
-\hline \begin{array}{l}
-\text { mechanistic } \\
-\text { model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} \\
-\hline \begin{array}{l}
-\text { structural } \\
-\text { causal model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} \\
-\hline \begin{array}{l}
-\text { causal } \\
-\text { graphical } \\
-\text { model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} \\
-\hline \begin{array}{l}
-\text { statistical } \\
-\text { model }
-\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} \\
-\hline
-\end{array}
-$$
-Source: [@peters_elements_2017]
 \normalsize
 
-# Does and Do nots
-
-**If you...**
-
-* ...only have conditional probabilities 
-* ...are only interested in association 
-* ...have very limited knowledge of your actual underlying causal structure
-
-$\rightarrow$ you should probably go for a probabilistic model, association based model or a model based on conditional probabilities
-
-**But if you**
-
-* ... have a good understanding of underlying latent factors 
-* ... have a lot of information on your underlying relationship in your data 
-* ... are dealing with high stake settings 
-* ... want to simulate different interventions before implementing them
-
-$\rightarrow$ you should probably go for a SCM
-
-# Concluding Remarks
-
-- Association-based learning is easy to model because of lower information neccessary
-- but not always appropriate in high stake settings 
-- SCM as simulator for causal modelling based on noise variables
-- SCMS entail a lot of information (DGP, intervention distribution)
-- Note: *"Garbage in, Garbage out"*
-- computational advantage casting causal model as assignments 
-- Enables modelling of higher order concepts like counterfactuals 
-- Extensions through differential equations for concise modelling of time
-
-\newpage 
-# References
-\tiny
-
-<div id="refs"></div>
-\normalsize
-\newpage
-
-# Appendix
-
-\footnotesize
-
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific 
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
-
-\normalsize
+#

### 2021-05-27 scm_short_presentation.Rmd
intermediate

@@
 
+# Definition & Assumptions SCM
+
+- Noise terms independent $(N_c,N_e)$
+- Mechanisms independent (other variables invariant)(local changes)
+
+Definition 1: A structural causal model M is given by a set of variables $X1, ..., Xd$ and corresponding assignments of the form $X_i := f_i(Pi, Ui), i = 1, ..., d .$
+Here, $P_i \subseteq f_{X1}, ..., X_{dg}$ is a subset of the variables that we call the parents of $Xi$.
+The random variables $U1, ..., Ud$ are called noise variables, which we require to be jointly independent.
+The directed graph corresponding to the model has one node for each variable Xi, which has incoming edges from all the parents Pi. 
+We will call such a graph the causal graph corresponding to the structural causal model.
+
 # Example
+
+\begin{columns}
+\begin{column}{0.48\textwidth}
 
@@
 \node[latent,right = 1cm of A](UC){$U_C$};
-\draw[->] (A) -- (E);
 \draw[->] (UA) -- (A);
@@
 \draw[->] (A) -- (C);
+\draw[->] (A) -- (E);
 \draw[->] (E) -- (C);
@@
 ```
+
+\end{column}
+\begin{column}{0.48\textwidth}
+$\{U_A\}$ - Numeric Ability
+
+$\{U_E\}$ - Literary Ability
+
+$\{U_C\}$ - Quality of the Student
+
+$\{A\}$ - Math Test
+
+$\{E\}$ - English-Test
+
+$\{C\}$ - GPA
+\end{column}
+\end{columns}
 
@@
 
-#
+# Questions for Discussion
+
+1. How do you think causal model will impact machine learning in the upcoming future? 
+2.

### 2021-05-28 scm_short_presentation.Rmd
scm short presentation

--- 2021-05-27 scm_short_presentation.Rmd
@@
 citation_package: --biblatex
-bibliography: [ref.bib]
-#csl: ieee.csl
 output: 
@@
 
-# Example
+# Assumptions SCM
+
+- Noise terms independent 
+- Mechanisms independent/autonomous 
+- Set of equations 
+- Endogenous and Exogenous Variables
+
+# Graphical Example
+
+\begin{columns}
+\begin{column}{0.48\textwidth}
 
@@
 \tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (A){A};
-\node[latent,left of=A](UA){$U_A$};
-\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (E) {E};
-\node[my node,right = 1 cm of E, fill=gray!30](C){C};
-\node[latent,left = 1cm of E](UE){$U_E$};
-\node[latent,right = 1cm of A](UC){$U_C$};
-\draw[->] (A) -- (E);
-\draw[->] (UA) -- (A);
+\node[my node, fill=gray!30] (N){N};
+\node[latent,left of=N](UN){$U_N$};
+\node[my node, fill=gray!30] at ($(N)!0.5!(N)-(0pt,1.5cm)$) (L) {L};
+\node[my node,right = 1 cm of L, fill=gray!30](E){E};
+\node[latent,left = 1cm of L](UL){$U_L$};
+\node[latent,right = 1cm of N](UE){$U_E$};
+\draw[->] (UN) -- (N);
+\draw[->] (UL) -- (L);
 \draw[->] (UE) -- (E);
-\draw[->] (UC) -- (C);
-\draw[->] (A) -- (C);
-\draw[->] (E) -- (C);
+\draw[->] (N) -- (E);
+\draw[->] (L) -- (E);
 \end{tikzpicture}
 ```
+
+\end{column}
+\begin{column}{0.48\textwidth}
+$\{U_N\}$ - Numeric Ability
+
+$\{U_L\}$ - Literary Ability
+
+$\{U_C\}$ - Quality of the Student
+
+$\{A\}$ - Math Test
+
+$\{E\}$ - English-Test
+
+$\{C\}$ - GPA
+
+\end{column}
+\end{columns}
+
+# Mathematical Example
+
+\begin{columns}
+\begin{column}{0.48\textwidth}
+
+$N:= f_N(U_N)$\newline
+$L:=f_L(U_L)$\newline
+$E:= f_E(N,L,U_E)$\newline
+
+\end{column}
+\begin{column}{0.48\textwidth}
+
+$\{U_N\}$ - Numeric Ability
+
+$\{U_L\}$ - Literary Ability
+
+$\{U_E\}$ - Quality of the Student
+
+$\{N\}$ - Math Test
+
+$\{L\}$ - English-Test
+
+$\{E\}$ - GPA
+
+\end{column}
+\end{columns}
 
@@
 
-#
+# Questions for Discussion

### 2021-05-28 scm_short_presentation.Rmd
cxyt

@@
 
+- Mechanisms independent/autonomous 
+- Set of equations (assignments ':=')
+- Observational Data 
+- Endogenous and Exogenous Variables
+- Latent noise variables 
 - Noise terms independent
-- Mechanisms independent/autonomous 
-- Set of equations 
-- Endogenous and Exogenous Variables
 
@@
 # Questions for Discussion
+
+Do you have any concepts that are still unclear to you?

### 2021-06-01 draft.Rmd
draft

@@
 - noise variables help generate joint distribution [@barocas].
+
+- latent factors great [@creager][@oberst2019counterfactual].
+- untestable assumptions [@pearl2009causality] frowned upon by statistics for long time
+- graphical framework better for modeling [@morgan_winship_2014] and reason why this appraoch is better than others
+
 SCMs are non-parametric structural equations models with added features.
@@
 Disturbance in SCM: Correlated and causal factor ; responsible for variation
-Disturbance in Regression: Uncorrelated
+Disturbance in Regression: Uncorrelated [@pearl2009causality]
 
@@
 
-exploit invariant characteristics of SEM whithout committing to shape. 
+exploit invariant characteristics of SEM without committing to shape. 
 structural if function autonomous and invariant to change in form of other functions.
+
+Example:
+
+(#TODO change to own example)
+Joint factorization sprinker
+
+$$
+P_{X_{3}=\mathrm{On}}\left(x_{1}, x_{2}, x_{4}, x_{5}\right)=P\left(x_{1}\right) P\left(x_{2} \mid x_{1}\right) P\left(x_{4} \mid x_{2}, X_{3}=\mathrm{On}\right) P\left(x_{5} \mid x_{4}\right)
+$$
+
+$$
+\begin{array}{l}
+P_{x}(v)=\prod_{\left\{i \mid V_{i} \notin X\right\}} P\left(v_{i} \mid p a_{i}\right)\\
+\text { for all } \boldsymbol{v} \text { consistent with } x
+\end{array}
+$$
 
@@
 Differential equations model time on a more mechanic manner [@mooij2013ordinary].
+scriptsize
 
 $$
-\begin{array}{|l|l|l|l|l|l|}
+\begin{array}{|l|l|l|l|l|}
 \hline \text { model } & \begin{array}{l}
-\text { predict in IID } \\
-\text { setting }
+\text { IID setting }
 \end{array} & \begin{array}{l}
-\text { predict under } \\
 \text { changing } \\
-\text { distributions / } \\
-\text { interventions }
-\end{array} & \begin{array}{l}
-\text { answer } \\
+\text { distributions } 
+\end{array} & \begin{array}{c}
 \text { counter- } \\
@@
 \end{array} & \begin{array}{l}
-\text { obtain } \\
 \text { physical } \\
 \text { insight }
-\end{array} & \begin{array}{l}
-\text { automatically } \\
-\text { learn from } \\
-\text { data }
 \end{array} \\

### 2021-06-02 draft.Rmd
draft

@@
 
-**Established Conditions:**
+Definition:Independence
 
-Definition: **SUTVA**
-
-'The treatment that one unit receives does not change the effect of treatment for any other unit.'[@hardtrecht]
-
-Definition: **Consistency**
-
-The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' [@hardtrecht]
-
-Definition: **Ignorability**
-
-The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. [@hardtrecht]
-(perfect RCT)
-
-- First tow hold for SCM counterfactuals 
-- third not testable but can check via backdoor criterion in SCM
-
-*Definition:***Independence**
+*The causal generative process of a system’s variables is composed of autonomous modules that do not inform or influence each other.*
+*In the probabilistic case, this means that the conditional distribution of each variable given its causes (i.e., its mechanism) does not inform or influence the other conditional distributions*,
+*In case we have only two variables, this reduces to an independence between the cause distribution and the mechanism producing the effect distribution* [@peters_elements_2017].
 
@@
 
+**Established Conditions for counterfactuals:**
+
+Definition: **SUTVA**
+
+'The treatment that one unit receives does not change the effect of treatment for any other unit.'[@hardtrecht]
+
+Definition: **Consistency**
+
+The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' [@hardtrecht]
+
+Definition: **Ignorability**
+
+The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. [@hardtrecht]
+(perfect RCT)
+
+- First tow hold for SCM counterfactuals 
+- third not testable but can check via backdoor criterion in SCM
+
 ## Implications

### 2021-06-09 paper_scm.Rmd
paper

--- 2021-06-01 draft.Rmd
@@
 - Department of Statistics, Ludwig Maximilian University Munich, Germany
-date: " April 5th, 2021"
+date: "June 2021"
 output: 
@@
 
-Most students have heard the phrase 'correlation does not imply causation.'
-While correlation implies co-occurrence, for many problems correlation is not enough. 
-Algorithmic decision making based on co-occurrence is insufficient  in high stake settings [@bareinboim2020pearl].
-For many problems we want to understand causal relationships between variables.
-There are different approaches on how to model causal relationships.
-This paper focuses on Structural Causal Models and Bayesian Causal Networks.
-Bayesian Causal Networks (BCN) cast a model based on **conditional probabilities.**
-Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality].
-Summarized, *'they are simulators that generate a distribution based on noise variables (independent) and a unique joint distribution...'* [@hardtrecht].
-- noise variables help generate joint distribution [@barocas].
+For many research problems, we want to understand the relationship between variables beyond co-occurrence.
+The most popular causal model is the structural causal model or in short SCM.
+The SCM is an expressive simulator to estimate causal relationships, accounting latent factors. 
+The basis of these SCMs is a set of equations, or more precisely assignments, providing functions to derive the conditional probabilities for our model [@hardtrecht].
+These assignments describe our variables in our model.
+Compared to probabilistic models, where we only specify conditional probabilities, the SCM actually enables the combination of theoretical and observational data.
+Conditional probabilities cannot represent latent variables because there is no conditional probability in our observational data for unobserved variables [@pearl2009causality].
 
-- latent factors great [@creager][@oberst2019counterfactual].
-- untestable assumptions [@pearl2009causality] frowned upon by statistics for long time
-- graphical framework better for modeling [@morgan_winship_2014] and reason why this appraoch is better than others
+The aim of this paper is to summarise SCMs and it's intersection with social sciences and physics.
 
-SCMs are non-parametric structural equations models with added features.
-Existing literature has provided an excellent introduction to SCMs and BCNs. 
-@pearl2009causality provided a comprehensive introduction into this topic with his book 'Causality'.
-His work addressed misconceptions in social sciences, causality and statistics.
-His contribution on the hierarchy of causation advanced the comprehensive understanding of causality. 
-@peters_elements_2017 address the relationship of causality, physical sciences, and causal inference.
-Their work highlights the importance of time in causality and inference based on observational data alone. 
-E.g. [@bareinboim2020pearl] worked on the application of causal methods in machine learning. 
-@tarka2018overview discuss the history of SEMs and SCMs.
-The aim of this paper is to summarise SCMs and it's intersection with social sciences and physics.
-I structure the rest of the paper as follows: 
-Section 2 introduces the assumptions in causal modelling. Further, section 2 discusses different causal differencecs, fundamental differences and a brief history of the methods. 
-Section 3 addresses Pearls Causal Hierachy
+Section 2 introduces the assumptions in causal modelling. Further, section 2 discusses different assumptions in causal modelling. Section 3
+Section 4 addresses Pearls Causal Hierachy.
+Section 5 provides a brief introduction into graphical models.
+Section 6 focuses on the intersection of SCMs and the perception of time.
 
-# Foundations of Structural Causal Models
+# Assumptions
 
-Consists of graph and assignments: 
-Baseline:
+This section discusses the independence assumption in SCMs.
 
-$$C:= N_c$$
-$$E:=f_E(C,N_E)$$
+**Independence:**
 
-source: @peters_elements_2017
+**Independent Mechanisms:**
 
-## Assumptions in Causality
+**Markov Condition**
 
-**Established Conditions:**
-
-Definition: **SUTVA**
-
-'The treatment that one unit receives does not change the effect of treatment for any other unit.'[@hardtrecht]
-
-Definition: **Consistency**
-
-The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' [@hardtrecht]
-
-Definition: **Ignorability**
-
-The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. [@hardtrecht]
-(perfect RCT)
-
-- First tow hold for SCM counterfactuals 
-- third not testable but can check via backdoor criterion in SCM
-
-*Definition:***Independence**
-
-If we specify the causal structure correctly:
-
-(a) possible to undertake local intervention -> change f(x), regardless of f(y|x)
-(b) these components are autonomous objects -> set of autonomous equations
-
-Further: independence of noise -> $$N_T , N_A$$
-
-Example conditional probabilities: independence conditional probabilities/distribution and mechanism (ICM)
-
-[@peters_elements_2017]
-
-- noise independent -> called **causal sufficiency** clause. 
-- Algorithmic independence
-
-SCM fulfills via :
-
-- dependency only via vertice connection (noise)
-- condition on noise, variables become independent  
-- Independence of mechanisms 
-- conditional independence: 
-- joint distribution
-
-Density factorizes and their expectation
-
-Causal assumptions differentiate causal models from association learning methods.
-
-These causal concepts are not expressible based on distribution functions/statistical associations. (Pearl 2010)
-
-Disturbance in SCM: Correlated and causal factor ; responsible for variation
-Disturbance in Regression: Uncorrelated [@pearl2009causality]
-
-Causal assumption not testable (e.g. $Cov(U_a, U_b)=0$).
-d-seperation to test assumptions in totality (cannot make assumptions in isolation).
-
-exploit invariant characteristics of SEM without committing to shape. 
-structural if function autonomous and invariant to change in form of other functions.
-
-Example:
-
-(#TODO change to own example)
-Joint factorization sprinker
+**Truncated Factorization**
 
@@
 
-*Definition:* **Reichenbachs common causal Principle**
+# SCM Example
 
-## Comparing Causal Tools
+$N:= f_N(U_N)$\newline
+$L:=f_L(U_L)$\newline
+$E:= f_E(N,L,U_E)$\newline
 
-A structural equation model (SEM) is a set of autonomous equations.
-This set represents the 'state of the world'. 
-SEMs are popular in fields like economics, psychology and sociology [@pearl2009causality].
-SCM are also defined by a (sub-)set of structural equations.
-Additionally, SCMs  feature mathematical components from graphical models and the potential-outcome framework [@pearl2012causal].
-There is a lot of controversy around SEMs.
-Many scholars challenge the parameter-specification in a SEM [@pearl2012causal].
-^[Inconsistency refers to inherent nature of creating parameters based of observational data.
-Observational data  is seldom consistent.
-Henceforth respective estimates are often questioned.
-For further information see @hernan2008does]
-SCMs specify an underlying data generating process without the computational effort of creating a parametric model.
-Note, that structural equations refer to assignment equations, used in computer science. 
-Equations in causal models did not always have a concise notation (Pearl, 2009).
-First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
-Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
-This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
-The problem is that the equation is symmetric.
-The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
-This misconception has caused a lot of challenges.^[For more information see @pearl2009causality]
-As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
-B defines A and the latent factor $U_A$.
-To summarize, a SCM consists of a set of (autonomous) equations to generate (a) endogenous variables and (b) exogenous variables.
+$\{U_N\}$ - Numeric Ability
 
-In Figure 1 we can see an example of a probabilistic model depicted  as a directed acyclic graph (DAG).
-The nodes are the white circles.
-The edges are the arrows, defined by the conditional probabilities.
-C is our **collider** variable because the affect of A on C and T on C collide [@pearl2009causality].
+$\{U_L\}$ - Literary Ability
 
-```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\node[my node] (A){A};
-\node[my node,right =1 cm of A](C){C};
-\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
-\draw[->] (A) -- (T);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
+$\{U_E\}$ - Quality of the Student
 
-In figure 2 we can see a vanilla structural causal model. 
-The square nodes represent the latent variables.
+$\{N\}$ - Math Test
+
+$\{L\}$ - English-Test
+
+$\{E\}$ - GPA
+
+# Pearl's Causal Hierachy
+
+ |Method          | Action | | Usage | 
+|------------------|-------------|-------------------|
+| Association $P(a|b)$               | Co-occurrence                           |(Un-)Supervised ML, BN, Reg.  
+| Intervention $P(a|do(b),c)$       | Do-manipulation                       |CBN,MDP,RL    
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals            | SCM ,PO
+
+**Association:**
+
+**Intervention:**
+
+**Counterfactuals:**
+
+Process is described as follows:
+
+(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
+(b) Action: Exchange $(X = x)$ 
+(c) Prediction: Compute $(Y = y)$
+
+# Graphical Models
+
+A structural causal model contains an underlying graphical model [@hardtrecht]. 
+The most popular graphical model is the directed acyclic graph or in short DAG.
+A DAG entails nodes and edges. 
+Nodes represent our different variables. 
+Edges depict the assignment equations. 
+All edges are directed in the DAG.
+The graph is called acyclic if no variable causes (directly or indirectly) itself [@morgan_winship_2014].
+There is a hierachy within DAGs.
 
@@
 
-The circle nodes represent the observed variables.
-The arrows are our structural equations and depict the relationship between variables.
-^[Note that there are also cyclic structural causal models but no cyclic bayesian causal networks.
-For further information see  @pearl2009causality.
-Due to the confined scope of this paper, I will not explore cyclic structures.
-
-Potential outcomes treat outcomes merely as potential outcomes rather rather than true outcomes. 
-- Only observe outcomes for units that are treated
-- ATE is a typical query
-- Every SCM can generate PO 
-- Cannot derive SCM from PO alone 
-- SCM has more concepts : causal graphs, mediating paths, counterfactuals for every variable etc. 
-- PO Easier to apply because apply to observational data [@hardtrecht]
-
-## A Brief History of Causal Modeling
-
-Path analysis is the foundation of modern structural causal models.
-This path analysis is a structural equation model with one variable per indicator.
-Sewall Wright, a Statistican and Geneticist, introduced the topic in the 1920s. (Pearl, 2012)
-Various other disciplines such as econometrics, psychometrics and sociology adopted path analysis.
-Aside from path diagrams also introduced graph rules to formalise relationships.
-(tbc. Cowles, Haavelmo,Aldrich, )
-
-# Pearl's Causal Hierachy
-
-To contextualise causal methods,  @pearl2009causality introduced the hierarchy of causation.
-Pearl focuses on three layers: association, intervention and counterfactuals (Table 1).
-
- |Method          | Action | | Usage | 
-|------------------|-------------|-------------------|
-| Association $P(a|b)$               | Co-occurrence                           |(Un-)Supervised ML, BN, Reg.  
-| Intervention $P(a|do(b),c)$       | Do-manipulation                       |CBN,MDP,RL    
-| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals            | SCM ,PO
-
-A higher level implies more detailed knowledge of the relationship between the variables.
- @pearl2009causality introduced the do-calculus.
- The do-calculus is a mathematical framework allowing for active substitution (E.g. $X:=10$) [@hardtrecht].
-The first query is association, where we examine relationships based on observations.
-
-Table: Pearls Hierachy of Causation (2009)
-
-## Association
-
-Vanilla machine learning (ML), bayesian networks (BN) and regression models (Reg) are at the lowest level in the causal hierarchy (see table 1).
-These methods demand the least information and depend on association alone.
-Associational methods ignore external changes outside of our data.
-The interventional distribution has information on these external changes.
-The interventional distribution is only defined in high level causal methods.
-
-## Intervention
-
-The second query deals with interventions.
-Here we can use @pearl2009causality do-calculus.
-The do-calculus enables us to study the manipulation of parent nodes.
- There are various types of intervention.
-One example is **atomic intervention,** where we set a variable to a constant.
-In **policy intervention** we specify a different function for an equation.
-off-policy intervention models different intervention that is not in our historical data [@oberst2019counterfactual].
-Causal bayesian networks , Markov Decision Processes (MDP) and reinforcement learning model intervention.
-(tbc.) Hypothetical interventions:
-
-## Counterfactuals
-
-The third query is counterfactual modelling. 
-Here we deal with hypothetical settings. 
-SCMs and potential outcome models allow for counterfactual modelling.
-These models can model counterfactuals because they include a interventional distribution [@oberst2019counterfactual].
-BCNs only entail conditional probabilities.
-There is no information on relationships outside of the observations in the data.
-Henceforth, they cannot create counterfactuals [@pearl2009causality].
-
-Process is described as follows:
-
-(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
-(b) Action: Exchange $(X = x)$ 
-(c) Prediction: Compute $(Y = y)$
-
-## Implications
-
-BCN:
-
-SCM:
-
-# Graphical Models
-
-Also called causal diagram
-- collider, mediator, fork [@barocas]
-
-## Features
-
-A graph in a SCM contains endogenous and exogenous variables 
-In a directed graph all edges have arrows.
-Directed arrows are direct (causal) effects.
-No edge between variables means that there is no causal effect. 
-If we have some edges without arrows, we call that graph semi-directed.
-If we have all edges without arrows, we call that graph un-directed.
-The most famous graphical model is the acyclic graph.
-An acyclic graph has no roots that cause itself (directly and indirectly).
-Because we do not specify changes over time we typically only deal with acyclic models.
-This is partially because we assume the relationship to be constant over time. [@creager]
-
-There are various tools to derive graphical models.^[The PC-Algorithm and the IC-Algorthm are two prominent examples.]
-
-- d-seperation 
-- vertices variables and arrows are direct effects.
-
-## Implications
-
-- no need functional dependency -> less assumptions about form
-- show collider variables
-- iv backdoor, estimation by mechanism
-[@morgan_winship_2014]
-
-- backdoor criterion non experimental way remove confounding [@barocas]
-
-Any DAG can be written as SCM.
-given joint disrtribution can be discover grahp? Markov condition
-
-# Causal Inference and Time
-
-Another issue in this paper is understanding the relationship of time and causality.
-SCMs make the assumption that causal relationships hold over time[@peters_elements_2017].
-These vague definitions of time is more prevalent in social sciences. 
-Meanwhile hard sciences deal with time in a more concise manner. 
-Differential equations model time on a more mechanic manner [@mooij2013ordinary].
-scriptsize
+# SCMs and Time
 
@@
 
-This paper provides a gentle introduction into structural causal models.
-SCMs entail many features, complementing research on association learning by providing depth. 
-This in turn, is of particular benefit for high stake decision settings. 
-SCMs differentiate from other methods through the specification of endogenous and exogenous variables, treating the exogenous factors as pivotal components of the actual model rather than assuming they are ommitable errors that are uncorrelated.
-As suggested by Pearls Hierachy, there are different levels to learning and each higher step can do anything the prior step can but with more detail and information. 
-Henceforth, as machin
-
 \newpage
@@
 \newpage
-
-# Appendix
-
- |Method         | CBN |  SCM
-|----------------|---------------------------|------------------------|
-| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
-|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
-|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
-|                |                                                                                                                      | 
-| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
-|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
-|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
-|                |                                                                                                                      | 
-|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
-|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
-
-Table: Source: Pearl (2009)

### 2021-06-10 paper_scm.Rmd
sq

@@
 
+History
+
 The aim of this paper is to summarise SCMs and it's intersection with social sciences and physics.
@@
 
+# SCM Example
+
+$S:= f_S(U_S)$
+$E:= f_E(U_E)$
+$C:= f_C(S,E,U_C)$
+
+$\{S\}$ - Smoking
+$\{C\}$ - Cancer
+$\{E\}$ - Eating
+
 # Assumptions
@@
 
+Independence of Noise, Independence of Mechanism, Causal Sufficiency, and why time is ignored.
+The question should be really, what are we commiting to when we use SCMs.
+
+**Acyclicity**
+
+Assume acyclic because conditional probabilities with variables causing itself very hard .
+
 **Independence:**
+
+Noise indepedent so noise smoking and noise eating should be independent.
 
@@
 
-# SCM Example
-
-$N:= f_N(U_N)$\newline
-$L:=f_L(U_L)$\newline
-$E:= f_E(N,L,U_E)$\newline
-
-$\{U_N\}$ - Numeric Ability
-
-$\{U_L\}$ - Literary Ability
-
-$\{U_E\}$ - Quality of the Student
-
-$\{N\}$ - Math Test
-
-$\{L\}$ - English-Test
-
-$\{E\}$ - GPA
-
 # Pearl's Causal Hierachy
@@
 |------------------|-------------|-------------------|
-| Association $P(a|b)$               | Co-occurrence                           |(Un-)Supervised ML, BN, Reg.  
-| Intervention $P(a|do(b),c)$       | Do-manipulation                       |CBN,MDP,RL    
-| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals            | SCM ,PO
+| Association $P(a|t)$               | Co-occurrence                           |(Un-)Supervised ML, BN, Reg.  
+| Intervention $P(a|do(t),c)$       | Do-manipulation                       |CBN,MDP,RL    
+| Counterfactual $P(a_t|a`,t`)$     | Hypotheticals            | SCM ,PO
 
@@
 
+-> example probabilistic terms
+
+$P(C|s)$
+
 **Intervention:**
+
+$P(a|do(t),c)$
+
+$do(t)$ -> replace function $T:=f_T(\pi)$ with different conditional probability (or constant) [@pearl2009causality].
 
@@
 There is a hierachy within DAGs.
+
+-> how to obtain scm
 
@@
 # SCMs and Time
+
+largely ingore time
+
+time in mechanical modeling crucial [@peters_elements_2017].
+
+- Time in Social Sciences: Often Vague 
+- Time in Physical Sciences: Mechanical via **Differential equations** 
+- dependence on prior time point and change in time contribute to the value at time point
+
+Initial Value: 
+$$
+\mathbf{x}\left(t_{0}\right)=\mathbf{x}_{0}
+$$
+
+Derivative of function x with respect to time t: 
+$$
+\frac{d \mathbf{x}}{d t}=f(\mathbf{x}), \mathbf{x} \in \mathbb{R}^{d}
+$$
+
+Value of Function at time t + dt: 
+$$
+\mathbf{x}(t+d t)=\mathbf{x}(t)+d t \cdot f(\mathbf{x}(t))
+$$
 

### 2021-06-10 paper_scm.Rmd
stuff

@@
 For many research problems, we want to understand the relationship between variables beyond co-occurrence.
+
+Research dealing with these question moved from traditional statistics, to ,
+
 The most popular causal model is the structural causal model or in short SCM.
@@
 Conditional probabilities cannot represent latent variables because there is no conditional probability in our observational data for unobserved variables [@pearl2009causality].
-
-History
 
@@
 
+The structural causal model consists of a set of equations.
+These equations are asymmetric, because they are not bi-directional.
+We cannot change the sides of the equation as one can do with regular equations [@pearl2012causal].
+
 $S:= f_S(U_S)$
@@
 $\{C\}$ - Cancer
-$\{E\}$ - Eating
+$\{E\}$ - Exercise
+
+A structural causal model contains an underlying graphical model [@hardtrecht]. 
+The most popular graphical model is the directed acyclic graph or in short DAG.
+A DAG entails nodes and edges. 
+Nodes represent our different variables. 
+Edges depict the assignment equations. 
+All edges are directed in the DAG.
+The graph is called acyclic if no variable causes (directly or indirectly) itself [@morgan_winship_2014].
+There is a hierachy within DAGs.
+
+```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+  \begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=6pt,
+    node distance=2cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
+\node[my node, fill=gray!30] (A){A};
+\node[latent,left of=A](UA){$U_A$};
+\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
+\node[my node,right = 1 cm of T, fill=gray!30](C){C};
+\node[latent,left = 1cm of T](UT){$U_T$};
+\node[latent,right = 1cm of A](UC){$U_C$};
+\draw[->] (A) -- (T);
+\draw[->] (UA) -- (A);
+\draw[->] (UT) -- (T);
+\draw[->] (UC) -- (C);
+\draw[->] (A) -- (C);
+\draw[->] (T) -- (C);
+\end{tikzpicture}
+```
+
+Probabilisitic Representation
+
+```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=7pt,
+    node distance=3cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\node[my node] (S){S};
+\node[my node,right =1 cm of S](E){E};
+\node[my node] at ($(S)!0.5!(E)-(0pt,1.5cm)$) (C) {C};
+\draw[->] (S) -- (C);
+\draw[->] (E) -- (C);
+\end{tikzpicture}
+```
 
@@
 (c) Prediction: Compute $(Y = y)$
-
-# Graphical Models
-
-A structural causal model contains an underlying graphical model [@hardtrecht]. 
-The most popular graphical model is the directed acyclic graph or in short DAG.
-A DAG entails nodes and edges. 
-Nodes represent our different variables. 
-Edges depict the assignment equations. 
-All edges are directed in the DAG.
-The graph is called acyclic if no variable causes (directly or indirectly) itself [@morgan_winship_2014].
-There is a hierachy within DAGs.
-
--> how to obtain scm
-
-```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-\usetikzlibrary{arrows}
-  \begin{tikzpicture}[
-    sharp corners=2pt,
-    inner sep=6pt,
-    node distance=2cm,
-    >=latex]
-\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (A){A};
-\node[latent,left of=A](UA){$U_A$};
-\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
-\node[my node,right = 1 cm of T, fill=gray!30](C){C};
-\node[latent,left = 1cm of T](UT){$U_T$};
-\node[latent,right = 1cm of A](UC){$U_C$};
-\draw[->] (A) -- (T);
-\draw[->] (UA) -- (A);
-\draw[->] (UT) -- (T);
-\draw[->] (UC) -- (C);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
-\end{tikzpicture}
-```
 

### 2021-06-10 presentation/presentation_video.Rmd
stuff

--- 2021-05-21 presentation_video.Rmd
@@
 
-- First tow hold for Counterfactuals in SCM
+- First two hold for Counterfactuals in SCM
 - third not testable but can check via backdoor criterion in SCM
@@
 - No need to specify exact parametric shape
-- Focus on acyclical structures (DAG)
+- Focus on acyclic structures (DAG)
 - Estimation back door criterion 

### 2021-06-10 presentation/scm_short_presentation.Rmd
stuff

--- 2021-05-28 scm_short_presentation.Rmd
@@
 
-- Noise terms independent 
 - Mechanisms independent/autonomous 
-- Set of equations 
+- Set of equations (assignments $':='$)
+- Observational Data 
 - Endogenous and Exogenous Variables
+- Latent noise variables to simulate joint distribution
+- Noise terms $U_1..U_i$ (joint) independence
 

### 2021-06-10 paper_scm.Rmd
paper

@@
 bibliography: [ref.bib]
-csl: ieee.csl
 nocite: '@*'
@@
 
-Section 2 introduces the assumptions in causal modelling. Further, section 2 discusses different assumptions in causal modelling. Section 3
-Section 4 addresses Pearls Causal Hierachy.
-Section 5 provides a brief introduction into graphical models.
-Section 6 focuses on the intersection of SCMs and the perception of time.
+Section 2 introduces the assumptions in causal modelling
+Section 3 addresses Pearls Causal Hierachy.
+Section 4 provides a brief introduction into graphical models.
+Section 5 focuses on the intersection of SCMs and the perception of time.
 
-# SCM Example
+# SCM Foundation
 
 The structural causal model consists of a set of equations.
-These equations are asymmetric, because they are not bi-directional.
+These equations are asymmetric assignments, because they are not bi-directional.
 We cannot change the sides of the equation as one can do with regular equations [@pearl2012causal].
 
-$S:= f_S(U_S)$
-$E:= f_E(U_E)$
-$C:= f_C(S,E,U_C)$
+We can set up a generic example with tree variables. 
+One common example from epidemiology is looking at the impact of problem behaviour and health enhancing behaviour.
+Problem behaviour is a latent variable, but we can use observational data to characterize problem behaviour.
+One example of problem behaviour is smoking 
+To estimate smoking, we can for instance ask participants how frequently they smoke and then describe p.
 
-$\{S\}$ - Smoking
-$\{C\}$ - Cancer
-$\{E\}$ - Exercise
+As we can see, these functions are driven by underling latent variables.
+These latent factors are the foundation of the structural causal model [@hardtrecht; @pearl2009causality; @pearl2012causal].
 
-A structural causal model contains an underlying graphical model [@hardtrecht]. 
+\begin{equation}
+S:= f_S(U_S)
+\end{equation}
+\begin{equation}
+E:= f_E(U_E)
+\end{equation}
+\begin{equation}
+C:= f_C(S,E,U_C)
+\end{equation}
+
+where: 
+$\{S\}$ - Frequency of Smoking
+$\{E\}$ - Frequency of Exercise
+$\{C\}$ - Lung Cancer
+
+Every structural causal model contains an underlying graphical model [@hardtrecht]. 
 The most popular graphical model is the directed acyclic graph or in short DAG.
@@
 \tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (A){A};
-\node[latent,left of=A](UA){$U_A$};
+\node[my node, fill=gray!30] (S){S};
+\node[latent,left of=S](US){$U_S$};
 \node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
@@
 
-Probabilisitic Representation
+For reference, we can also display this model based on other models. 
+One example is the causal bayesian network, which uses conditional probabilities instead of functions to describe the relationship between variables. 
+In the probabilistic representation, we ignore latent factors.
 

### 2021-06-10 paper_scm.Rmd
paper sq

@@
 For many research problems, we want to understand the relationship between variables beyond co-occurrence.
+The most popular causal model is the structural causal model or in short SCM.
+The SCM is the non-parametric counterpart to structural equation models.
+Researchers on structural causal models tried to distance themselves from common practices for structural equation models [@pearl2009causality; @peters_elements_2017].
+The geneticist and statistican Sewall Wright introduced the first ancestor of the SCM, the path analysis [@pearl2009causality].
+Path analysis now falls into the broader class of structural equation models.
+Path analysis is a structural equation model with one variable per indicator.
 
-Research dealing with these question moved from traditional statistics, to ,
-
-The most popular causal model is the structural causal model or in short SCM.
-The SCM is an expressive simulator to estimate causal relationships, accounting latent factors. 
+The SCM is an expressive simulator to estimate causal relationships, accounting for latent factors. 
+Latent factors are unobserved factors [@pearl2009causality].
+SCMs entail endogenous and exogenous variables. 
+Exogenous variables should not be confused with latent variables.
+@peters_elements_2017 define endogenous and exogenous as follows: *Endogeneous variables are those that the modeler tries to understand, while exogenous ones are determined by factors outside the model, and are taken as given.*
 The basis of these SCMs is a set of equations, or more precisely assignments, providing functions to derive the conditional probabilities for our model [@hardtrecht].
 These assignments describe our variables in our model.
-Compared to probabilistic models, where we only specify conditional probabilities, the SCM actually enables the combination of theoretical and observational data.
+Compared to probabilistic models, where we only specify conditional probabilities, the SCM actually enables the combination of latent variables and observational data.
 Conditional probabilities cannot represent latent variables because there is no conditional probability in our observational data for unobserved variables [@pearl2009causality].
 
-The aim of this paper is to summarise SCMs and it's intersection with social sciences and physics.
+The aim of this paper is to provide a brief and intuitive introduction to SCMs.
 
@@
 We cannot change the sides of the equation as one can do with regular equations [@pearl2012causal].
+[@peters_elements_2017] define a SCM as follows:
 
-We can set up a generic example with tree variables. 
-One common example from epidemiology is looking at the impact of problem behaviour and health enhancing behaviour.
-Problem behaviour is a latent variable, but we can use observational data to characterize problem behaviour.
-One example of problem behaviour is smoking 
-To estimate smoking, we can for instance ask participants how frequently they smoke and then describe p.
+**Definition 1:** Structural Causal Model:
 
-As we can see, these functions are driven by underling latent variables.
+*An SCM* $\mathbb{C}$ *with graph* $C \rightarrow E$ *consists of two assignments*
+
+$$
+\begin{array}{l}
+C:=N_{C} \\
+E:=f_{E}\left(C, N_{E}\right)
+\end{array}
+$$
+*where* $N_{F} \perp \!\!\! \perp N_{C}$ *that is* $N_{F}$ *is independent of* $N_{C}$
+
+In their definition, C is the cause and E is the effect.
+
+We can set up an epidemiological example with three variables, looking at the impact of problem behavior and genetic code on lung cancer. 
+Problem behavior is a latent variable, but we can use observational data to characterize problem behavior.
+One example of problem behavior is smoking.
+Henceforth,to estimate problem behaviour, we can for instance ask participants how frequently they smoke.
+For genetic code, we could look for specific genes and and examine whether the presence of specific genes has an impact on getting cancer. 
+As we can see, these functions are driven by underlying latent variables.
 These latent factors are the foundation of the structural causal model [@hardtrecht; @pearl2009causality; @pearl2012causal].
@@
 \begin{equation}
-E:= f_E(U_E)
+G:= f_G(U_G)
 \end{equation}
 \begin{equation}
-C:= f_C(S,E,U_C)
+C:= f_C(S,G,U_C)
 \end{equation}
@@
 where: 
-$\{S\}$ - Frequency of Smoking
-$\{E\}$ - Frequency of Exercise
-$\{C\}$ - Lung Cancer
+$\{S\}$ - Frequency of smoking
+$\{G\}$ - Presence of specific genes
+$\{C\}$ - Lung cancer
 
 Every structural causal model contains an underlying graphical model [@hardtrecht]. 
+This is one important feature that differentiates SCMs from other frameworks^[e.g. the Potential Outcome framework[@pearl2009causality]].
+
+[@pearl2009causality] describes the SCM a process based tool, because it enables researchers to reflect on their underlying assumptions. 
+The SCM requires more assumptions and thought because we actually need to define an admissible set of variables, ensure they are independent and ensure that the underlying factors do not correlate.
+By being forced to think about all these steps, SCMs help to avoid poorly specified probabilistic specifications.
+Various research has pointed outexamples where modeling without DAGs lead to severe mistakes: 
+@hirano2001estimation suggest a method for covariate selection that according to @pearl2009myth favours bias-enhancing features in the propensity score.
+Further @bollen2013eight (2013) argue that @rosenbaum2002overt and @rubin2007design falsely declared that 'there is no reason to avoid adjustment for a variable describing subjects before treatment'.
+
+We can craete graphical models based on various different algorithms.
+
 The most popular graphical model is the directed acyclic graph or in short DAG.
+A causal graph for a SCM contains endogenous and exogenous variables. 
 A DAG entails nodes and edges.
+
 Nodes represent our different variables. 
@@
 All edges are directed in the DAG.
-The graph is called acyclic if no variable causes (directly or indirectly) itself [@morgan_winship_2014].
-There is a hierachy within DAGs.
+An acyclic graph has no roots that cause itself (directly and indirectly) [@morgan_winship_2014].
+
+This acyclic structure is important for the conditional probabilities [@forre2020causal].
+
+may not find unique solution if cyclic in equilbrium [@peters_elements_2017]
+
+Variables have incoming paths from their parent nodes.
+
+These DAGs can be built on theory.
+Another way to determine the DAG structure is using observational data. 
+One of the more prominent algorithm to estimate the underlying dag structure is the pc-algorithm [@kalisch2012causal] .
 
@@
 \node[latent,left of=S](US){$U_S$};
-\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
-\node[my node,right = 1 cm of T, fill=gray!30](C){C};
-\node[latent,left = 1cm of T](UT){$U_T$};
-\node[latent,right = 1cm of A](UC){$U_C$};
-\draw[->] (A) -- (T);
-\draw[->] (UA) -- (A);
-\draw[->] (UT) -- (T);
+\node[my node, fill=gray!30] at ($(S)!0.5!(S)-(0pt,1.5cm)$) (G) {G};
+\node[my node,right = 1 cm of G, fill=gray!30](C){C};
+\node[latent,left = 1cm of G](UG){$U_G$};
+\node[latent,right = 1cm of S](UC){$U_C$};
+\draw[->] (US) -- (S);
+\draw[->] (UG) -- (G);
 \draw[->] (UC) -- (C);
-\draw[->] (A) -- (C);
-\draw[->] (T) -- (C);
+\draw[->] (S) -- (C);
+\draw[->] (G) -- (C);
 \end{tikzpicture}
@@
 For reference, we can also display this model based on other models. 
-One example is the causal bayesian network, which uses conditional probabilities instead of functions to describe the relationship between variables. 
-In the probabilistic representation, we ignore latent factors.
+One example is the causal bayesian network, which uses conditional probabilities instead of functions to describe the relationship between variables [@pearl2009causality]
+In the probabilistic representation, we ignore latent factors [@creager; @pearl2009causality].
 
@@
 \node[my node] (S){S};
-\node[my node,right =1 cm of S](E){E};
-\node[my node] at ($(S)!0.5!(E)-(0pt,1.5cm)$) (C) {C};
+\node[my node,right =1 cm of S](G){G};
+\node[my node] at ($(S)!0.5!(G)-(0pt,1.5cm)$) (C) {C};
 \draw[->] (S) -- (C);
-\draw[->] (E) -- (C);
+\draw[->] (G) -- (C);
 \end{tikzpicture}
@@
 
-# Assumptions
+Graphical models developed a mathematical langauge to assess certain conditions. 
+One pivotal feature is the d-separation [@judea2010introduction]:
 
-This section discusses the independence assumption in SCMs.
+**Definition 2** ( $d$ -separation) A set $S$ of nodes is said to block a path $p$ if either (i) $p$ contains at least one arrow-emitting node that is in $S$, or $($ ii $)$ p contains at least one collision node that is outside $S$ and has no descendant in $S .$
+If $S$ blocks all paths from $X$ to $Y$, it is said to " $d$ -separate $X$ and $Y, "$ and then, $X$ and $Y$ are independent given $S$, written $X \perp \!\!\! \perp Y \mid S .$
 
-Independence of Noise, Independence of Mechanism, Causal Sufficiency, and why time is ignored.
-The question should be really, what are we commiting to when we use SCMs.
+Select subset of variables with backdoor criterion:
 
-**Acyclicity**
+**Definition:** Independence
 
-Assume acyclic because conditional probabilities with variables causing itself very hard .
-
-**Independence:**
+Independence of Noise,
 
@@
 **Independent Mechanisms:**
+
+\begin{equation}
+\begin{aligned}
+p(a, t) &=p(a \mid t) p(t) \\
+&=p(t \mid a) p(a)
+\end{aligned}
+\end{equation}
+
+**Definition:** Causal Sufficiency
+
+A set of variables X is usually said to be causally sufficient if there is no hidden
+common cause $C \notin X$ that is causing more than one variable in X [@peters_elements_2017; @spirtes2010introduction]
+
+ and why time is ignored.
 
@@
 
+Associational methods ignore external changes outside of our data.
+The interventional distribution has information on these external changes.
+Note, that the interventional distribution is only defined in high level causal methods.
+
 **Intervention:**
+
+Here we can use @pearl2009causality do-calculus.
+The do-calculus enables us to study the manipulation of parent nodes.
+ There are various types of intervention.
+One example is **atomic intervention,** where we set a variable to a constant.
+
+```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+  \begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=6pt,
+    node distance=2cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
+\node[my node, fill=gray!30] (p){s};
+\node[my node, fill=gray!30] at ($(p)!0.5!(p)-(0pt,1.5cm)$) (G) {G};
+\node[my node,right = 1 cm of G, fill=gray!30](C){C};
+\node[latent,left = 1cm of G](UG){$U_G$};
+\node[latent,right = 1cm of p](UC){$U_C$};
+\draw[->] (UG) -- (G);
+\draw[->] (UC) -- (C);
+\draw[->] (p) -- (C);
+\draw[->] (G) -- (C);
+\end{tikzpicture}
+```
+
+In **policy intervention** we specify a different function for an equation.
+off-policy intervention models different intervention that is not in our historical data [@oberst2019counterfactual].
 
@@
 $do(t)$ -> replace function $T:=f_T(\pi)$ with different conditional probability (or constant) [@pearl2009causality].
+
+Suppose the government wants to examine, whether increasing fines for smoking prohibited areas will led to less smoking and henceforth less lung cancer.
+The government could suggest to increase fines similar to the charges in Singapore by fining people up to a $1000 for smoking in these areas.
+Alternative, the government could also undertake treatment by imposing higher tabacco taxes.
+
+```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+  \begin{tikzpicture}[
+    sharp corners=2pt,
+    inner sep=6pt,
+    node distance=2cm,
+    >=latex]
+\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
+\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
+\node[my node, fill=gray!30] (p){$\pi$};
+\node[latent,left of=p](US){$U_S$};
+\node[my node, fill=gray!30] at ($(p)!0.5!(p)-(0pt,1.5cm)$) (G) {G};
+\node[my node,right = 1 cm of G, fill=gray!30](C){C};
+\node[latent,left = 1cm of G](UG){$U_G$};
+\node[latent,right = 1cm of p](UC){$U_C$};
+\draw[->] (US) -- (p);
+\draw[->] (UG) -- (G);
+\draw[->] (UC) -- (C);
+\draw[->] (p) -- (C);
+\draw[->] (G) -- (C);
+\end{tikzpicture}
+```
 
@@
 (c) Prediction: Compute $(Y = y)$
+
+**Stable Unit Treatment Value Assumption {SUTVA}** 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
+**Consistency** The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' 
+**Ignorability** The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. 
+As suggested by [@hardtrecht], this condition ensures that we are dealing with a perfect randomized controlled trial.
+
+- First two hold for Counterfactuals in SCM
+- third not testable but can check via backdoor criterion in SCM
+- Source: [@hardtrecht]
 
@@
 
+Structural causal models are flexible simulators to disentangle causality for manifold different queries.
+There are many advantages of structural causal models:
+(1) We are able to model latent fators forcing us to reconsider existing assumptions about the relationship in our data.
+(2) Further, we get a underlying graphical representation including a mathematical language for this graphical systems to test causal assumptions that are otherwise untestable.
+(3) Additionally, one is able to model queries beyond mere association going as far as dealing with hyptoethical situations. 
+Simultaneously,
+
 \newpage

### 2021-06-10 paper_scm.Rmd
essay

@@
 **Abstract** The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling.
-To provide a comprehensive understanding of causal modelling, I introduce two prominent causal model specifications namely (1) Bayesian Causal Networks (BCN) and (2) Structural Causal Models (SCM), focusing on the latter.Probabilistic specifications such as a BCN cast a model based on conditional probabilities.
-SCMs cast a model based on assignment functions and extend probabilistic models by specifying the data generating process rather than solely utilizing conditional probabilities.
+Probabilistic specifications cast a model based on conditional probabilities.
+SCMs cast a model based on  assymetric assignments and extend probabilistic models by specifying the entire data generating process rather than solely utilizing conditional probabilities.
 Another difference between these models is their ability to address different queries such as *predictions*, *interventions* and *counterfactuals*. 
@@
 Exogenous variables should not be confused with latent variables.
-@peters_elements_2017 define endogenous and exogenous as follows: *Endogeneous variables are those that the modeler tries to understand, while exogenous ones are determined by factors outside the model, and are taken as given.*
+For reference, @peters_elements_2017 define endogenous and exogenous as follows: *Endogeneous variables are those that the modeler tries to understand, while exogenous ones are determined by factors outside the model, and are taken as given.*
 The basis of these SCMs is a set of equations, or more precisely assignments, providing functions to derive the conditional probabilities for our model [@hardtrecht].
@@
 
-# SCM Foundation
+# Assumptions in Structural Causal Models
 
@@
 
-For reference, we can also display this model based on other models. 
-One example is the causal bayesian network, which uses conditional probabilities instead of functions to describe the relationship between variables [@pearl2009causality]
+Graphical models developed a mathematical langauge to assess certain conditions. 
+One pivotal feature is the d-separation [@judea2010introduction]:
+
+**Definition 2** ( $d$ -separation) A set $S$ of nodes is said to block a path $p$ if either (i) $p$ contains at least one arrow-emitting node that is in $S$, or $($ ii $)$ p contains at least one collision node that is outside $S$ and has no descendant in $S .$
+If $S$ blocks all paths from $X$ to $Y$, it is said to " $d$ -separate $X$ and $Y, "$ and then, $X$ and $Y$ are independent given $S$, written $X \perp \!\!\! \perp Y \mid S .$
+
+Select subset of variables with backdoor criterion:
+
+## Independence
+
+**Definition:** Independence 
+*The causal generative process of a system’s variables is composed of autonomous modules that do not inform or influence each other.*
+*In the probabilistic case, this means that the conditional distribution of each variable given its causes (i.e., its mechanism) does not inform or influence the other conditional distributions*,
+*In case we have only two variables, this reduces to an independence between the cause distribution and the mechanism producing the effect distribution* [@peters_elements_2017].
+
+If we specify the causal structure correctly:
+
+(a) possible to undertake local intervention -> change f(x), regardless of f(y|x)
+(b) these components are autonomous objects -> set of autonomous equations
+
+Independence of Noise,
+
+Noise independent so problematic behaviour and genetic code should be independent.
+
+**Mechanism**
+
+\begin{equation}
+\begin{aligned}
+p(a, t) &=p(a \mid t) p(t) \\
+&=p(t \mid a) p(a)
+\end{aligned}
+\end{equation}
+
+\begin{equation}
+p^{\mathrm{o}}(a, t)=p(t \mid a) p^{\mathrm{o}}(a) \text { and } p^{\mathrm{s}}(a, t)=p(t \mid a) p^{\mathrm{s}}(a) .
+\end{equation}
+
+**Definition:** Causal Sufficiency
+
+*A set of variables X is usually said to be causally sufficient if there is no hidden common cause $C \notin X$ that is causing more than one variable in X* [@peters_elements_2017; @spirtes2010introduction]
+
+$$
+P(Cancer|Smoking)=P(Smoking|Cancer) \times P(Cancer)
+$$
+
+**Markov Condition**
+
+@pearl2009causality defines the markov condition as follows:
+
+**Definition:** Causal Markov Condition
+
+*Every Markovian causal model M induces a distribution P(x1,…, xn) that satisfies the parental Markov condition relative the causal diagram G associated with M; that is, each variable Xi is independent of all its nondescendants, given its parents PAi in G.*
+
+$$
+\begin{array}{l}
+P_{x}(v)=\prod_{\left\{i \mid V_{i} \notin X\right\}} P\left(v_{i} \mid p a_{i}\right)\\
+\text { for all } \boldsymbol{v} \text { consistent with } x
+\end{array}
+$$
+
+## Time
+
+ why time is ignored.
+
+# Pearl's Causal Hierachy
+
+\footnotesize
+ |Method          | Action |  Example | Usage | 
+|------------------|-------------|--------------------|-------------------|
+| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
+| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
+| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
+
+Table: Pearls Hierachy of Causation (2009)
+
+\normalsize
+
+## Association:
+
+Associational methods ignore external changes outside of our data.
+The interventional distribution has information on these external changes.
+Note, that the intervention distribution is only defined in high level causal methods.
+
+Joint distribution
+
+$P(c,s,g)= P(s)\times P(c|s,g) \times P(g)$
+
+**Graph**
+
+One example is the bayesian network, which uses conditional probabilities instead of functions to describe the relationship between variables [@pearl2009causality]
 In the probabilistic representation, we ignore latent factors [@creager; @pearl2009causality].
@@
 
-Graphical models developed a mathematical langauge to assess certain conditions. 
-One pivotal feature is the d-separation [@judea2010introduction]:
-
-**Definition 2** ( $d$ -separation) A set $S$ of nodes is said to block a path $p$ if either (i) $p$ contains at least one arrow-emitting node that is in $S$, or $($ ii $)$ p contains at least one collision node that is outside $S$ and has no descendant in $S .$
-If $S$ blocks all paths from $X$ to $Y$, it is said to " $d$ -separate $X$ and $Y, "$ and then, $X$ and $Y$ are independent given $S$, written $X \perp \!\!\! \perp Y \mid S .$
-
-Select subset of variables with backdoor criterion:
-
-**Definition:** Independence
-
-Independence of Noise,
-
-Noise indepedent so noise smoking and noise eating should be independent.
-
-**Independent Mechanisms:**
-
-\begin{equation}
-\begin{aligned}
-p(a, t) &=p(a \mid t) p(t) \\
-&=p(t \mid a) p(a)
-\end{aligned}
-\end{equation}
-
-**Definition:** Causal Sufficiency
-
-A set of variables X is usually said to be causally sufficient if there is no hidden
-common cause $C \notin X$ that is causing more than one variable in X [@peters_elements_2017; @spirtes2010introduction]
-
- and why time is ignored.
-
-**Markov Condition**
-
-**Truncated Factorization**
-
-$$
-P_{X_{3}=\mathrm{On}}\left(x_{1}, x_{2}, x_{4}, x_{5}\right)=P\left(x_{1}\right) P\left(x_{2} \mid x_{1}\right) P\left(x_{4} \mid x_{2}, X_{3}=\mathrm{On}\right) P\left(x_{5} \mid x_{4}\right)
-$$
-
-$$
-\begin{array}{l}
-P_{x}(v)=\prod_{\left\{i \mid V_{i} \notin X\right\}} P\left(v_{i} \mid p a_{i}\right)\\
-\text { for all } \boldsymbol{v} \text { consistent with } x
-\end{array}
-$$
-
-# Pearl's Causal Hierachy
-
- |Method          | Action | | Usage | 
-|------------------|-------------|-------------------|
-| Association $P(a|t)$               | Co-occurrence                           |(Un-)Supervised ML, BN, Reg.  
-| Intervention $P(a|do(t),c)$       | Do-manipulation                       |CBN,MDP,RL    
-| Counterfactual $P(a_t|a`,t`)$     | Hypotheticals            | SCM ,PO
-
-**Association:**
-
--> example probabilistic terms
-
-$P(C|s)$
-
-Associational methods ignore external changes outside of our data.
-The interventional distribution has information on these external changes.
-Note, that the interventional distribution is only defined in high level causal methods.
-
-**Intervention:**
+## Intervention:
 
 Here we can use @pearl2009causality do-calculus.
+
 The do-calculus enables us to study the manipulation of parent nodes.
+
+Instead of merely seeing the co-occurence of variables, we can actively manipulate the conditional distribution of one variable.
+Commonly, one could use this method to evaluate different policies [@creager].
+In our smoking example we can actively set the smoking behavior to a fixed value or a different conditional probability.
+
+Joint distribution post intervention: $P_{S=s}(c,g)= P(c|S=s,g) \times P(g)$
+
 There are various types of intervention.
-One example is **atomic intervention,** where we set a variable to a constant.
 
-```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
+I will illustrate atomic intervention and policy intervention based on our smoking example.
+
+Suppose the government wants to examine, a different smoking behavior distribution will lower the number of lung cancer cases and medical expenses.
+This different smoking behaviour could be the result of increased taxes, or e.g. finesup to a $1000 (as in the case of Singapore) for smoking in prohibited areas.
+Alternative, the government could also undertake treatment by imposing higher tobacco taxes.
+In this case we solely focus on the effect of a different smoking behavior on lung cancer for the sake of simplicity.
+
+In **atomic intervention,** we set a variable to a constant value.
+As one can see in figure 3, c is constant that is not dependent on the latent factor, because in atomic intervention we do not derive the value of c based on the function S.
+
+```{tikz,fig.cap="Atomic Intervention" ,fig.align="center", echo =F}
 \usetikzlibrary{calc}
@@
 \tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (p){s};
+\node[my node, fill=gray!10] (p){c};
 \node[my node, fill=gray!30] at ($(p)!0.5!(p)-(0pt,1.5cm)$) (G) {G};
@@
 
-In **policy intervention** we specify a different function for an equation.
-off-policy intervention models different intervention that is not in our historical data [@oberst2019counterfactual].
+Further, in mathematical notation the model would change as follows:
 
-$P(a|do(t),c)$
+\begin{equation}
+S:= s
+\end{equation}
+\begin{equation}
+G:= f_G(U_G)
+\end{equation}
+\begin{equation}
+C:= f_C(S,G,U_C)
+\end{equation}
 
-$do(t)$ -> replace function $T:=f_T(\pi)$ with different conditional probability (or constant) [@pearl2009causality].
+In **policy intervention** we specify a different conditional probability $(S=s)$ for an equation.
+We can derive s from S, because we include information on the intervention distribution. 
+This information cannot be obtained, if we directly specify our model as conditional probabilities.
 
-Suppose the government wants to examine, whether increasing fines for smoking prohibited areas will led to less smoking and henceforth less lung cancer.
-The government could suggest to increase fines similar to the charges in Singapore by fining people up to a $1000 for smoking in these areas.
-Alternative, the government could also undertake treatment by imposing higher tabacco taxes.
-
-```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
+```{tikz,fig.cap="Policy Model" ,fig.align="center", echo =F}
 \usetikzlibrary{calc}
@@
 \tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
-\node[my node, fill=gray!30] (p){$\pi$};
+\node[my node, fill=gray!10] (p){s};
 \node[latent,left of=p](US){$U_S$};
@@
 
+where $\pi$ is our new conditional probability.
+
+One extension of policy intervention is off-policy intervention. 
+**Off-policy intervention** models different intervention that is not in our historical data.
+To accommodate the lack of data, we can estimate the intervention either based on model-based (e.g. regression),model-free estimates (propensity scores), or a mixture of these methods. 
+For further information, see e.g. @oberst2019counterfactual or [@creager].
+
+$do(t)$ -> replace function $T:=f_T(\pi)$ with different conditional probability (or constant) [@pearl2009causality].
+
 **Counterfactuals:**
@@
 **Stable Unit Treatment Value Assumption {SUTVA}** 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
+
 **Consistency** The outcome Y agrees with the potential outcome corresponding to the treatment indicator.'
+
 **Ignorability** The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. 
@@
 Derivative of function x with respect to time t:
+
 $$
@@
 Value of Function at time t + dt:
+
 $$

### 2021-06-11 paper_scm.Rmd
20 more lines

@@
 The SCM is the non-parametric counterpart to structural equation models.
-Researchers on structural causal models tried to distance themselves from common practices for structural equation models [@pearl2009causality; @peters_elements_2017].
+Researchers on SCMs tried to distance themselves from common practices for structural equation models [@pearl2009causality; @peters_elements_2017].
 The geneticist and statistican Sewall Wright introduced the first ancestor of the SCM, the path analysis [@pearl2009causality].
@@
 
-The aim of this paper is to provide a brief and intuitive introduction to SCMs.
+To accommodate existing literature, this paper provides a gentle introduction to SCMs, focusing on the underlying assumptions we make when building a structural causal model. 
+The aim of this paper is to provide an example based, inutitive understanding of different concepts within causality using manifold elements from core literature on causality.
 
-Section 2 introduces the assumptions in causal modelling
+The rest of the paper is structured as follows:
+Section 2 introduces the assumptions in causal modelling.
 Section 3 addresses Pearls Causal Hierachy.
-Section 4 provides a brief introduction into graphical models.
-Section 5 focuses on the intersection of SCMs and the perception of time.
+Section 4 focuses on the intersection of SCMs and the perception of time. 
+Section 5 concludes.
 
@@
 [@pearl2009causality] describes the SCM a process based tool, because it enables researchers to reflect on their underlying assumptions. 
-The SCM requires more assumptions and thought because we actually need to define an admissible set of variables, ensure they are independent and ensure that the underlying factors do not correlate.
+The SCM requires more assumptions and thought.
+Even for a very minimalisitic SCM, we need to define an admissible set of variables, ensure the random noise terms are independent and corroborate that the underlying mechanisms are autonomous. 
 By being forced to think about all these steps, SCMs help to avoid poorly specified probabilistic specifications. 
-Various research has pointed outexamples where modeling without DAGs lead to severe mistakes: 
+Various research has pointed out examples where modeling without DAGs lead to severe mistakes: 
 @hirano2001estimation suggest a method for covariate selection that according to @pearl2009myth favours bias-enhancing features in the propensity score.
@@
 
+Pearl [@bareinboim2020pearl; @pearl2009causality] introduced the hierachy of causation to categorize different statistical and causal tools. 
+The hierarchy of causation contains (see table 1) three levels. 
+The higher the method on the hierarchy, the more information the method requires.
+This section discusses each of these methods and their respective advantages.
+
 \footnotesize
@@
 
-Associational methods ignore external changes outside of our data.
+The first level, association, requires the least information.
+Association based methods are most prevalent and contain the largest class of methods.
+Standard statistical tools such as regression analysis, supervised and unsupervsied learning and bayesian networks all fall into this category.[@bareinboim2020pearl]
+The underlying action for association is co-occurence.
+As prominently criticized by @bender2021dangers, this reduces the number of questions we can answer, because methods are heavily dependent on the observational data. 
+In the context of deep learning and the advancement of natural language processing, @bender2021dangers suggest that many association-based methods results in stochastic parrots as opposed to natural langauge understanding.
+
+Association-based methods ignore external changes outside of our data.
 The interventional distribution has information on these external changes.
-Note, that the intervention distribution is only defined in high level causal methods.
+Note, that the intervention distribution is only defined in high order methods.
 
-Joint distribution
+Reiterating our lung cancer example, the full joint distribution looks as follows:
 
-$P(c,s,g)= P(s)\times P(c|s,g) \times P(g)$
+$P(c,s,g)= P(s) \times P(c|s,g) \times P(g)$
+
+If there are any changes in our distribution, e.g. smoking changes its distribution from $s$ to $s_{new}$ we are unable to accomodate these changes [@peters_elements_2017].
+Association-based methods are not equipped to accodate such modifications.
 
@@
 
-```{tikz,fig.cap="Policy Model" ,fig.align="center", echo =F}
+```{tikz,fig.cap="Policy Intervention" ,fig.align="center", echo =F}
 \usetikzlibrary{calc}
@@
 
-where $\pi$ is our new conditional probability.
+where $s$ is our new conditional probability.
 

### 2021-06-11 paper_scm.Rmd
recent v

@@
 Probabilistic specifications cast a model based on conditional probabilities. 
-SCMs cast a model based on  assymetric assignments and extend probabilistic models by specifying the entire data generating process rather than solely utilizing conditional probabilities.
-Another difference between these models is their ability to address different queries such as *predictions*, *interventions* and *counterfactuals*. 
+SCMs cast a model based on  asymmetric assignments and extend probabilistic models by specifying the entire data generating process rather than solely utilizing conditional probabilities.
+A SCM is build on a number of different assumptions namely the independence of noise, autonomy of mechanisms, causal sufficiency and the causal markov condition.
+I also discuss the role of time in causality and how differential equations enable us to model SCMs similar to a mechanical mdoel.
+Another difference between thes SCM and other model specifications is the ability  of SCMs to address different queries such as *predictions*, *interventions* and *counterfactuals*. 
 These queries are part of Pearl's causal hierarchy (2009).
 Pearl matches these queries with their respective actions namely *observing*, *doing* and *imagining*. 
-I compare the feasibility of addressing these queries and undertaking respective actions for both specifications.
-To contextualize SCMs within the field of causality, I also discuss the role of time in causality.
-This paper uses various directed acyclic graphs to highlight the differences in these modelling approaches.
+I compare the feasibility of addressing these queries.
 The insights of this paper can be used as a baseline for subsequent research on structural causal models.
@@
 To accommodate existing literature, this paper provides a gentle introduction to SCMs, focusing on the underlying assumptions we make when building a structural causal model.
-The aim of this paper is to provide an example based, inutitive understanding of different concepts within causality using manifold elements from core literature on causality.
+Specifically, this paper focuses on the independence of noise, independence of mechanisms, causal sufficiency, the role of time and the causal markov condition.
+Additionally, this paper provides an overview of various other statistical and causal methods within the grand scheme of pearls causal hierachy.
+I accomodate this outlook with various considerations for the usage of SCMs. 
+The aim of this paper is to provide an example based, intuitive understanding of different concepts within causality using manifold elements from core literature on causality.
 
 The rest of the paper is structured as follows:
-Section 2 introduces the assumptions in causal modelling.
-Section 3 addresses Pearls Causal Hierachy.
-Section 4 focuses on the intersection of SCMs and the perception of time. 
-Section 5 concludes.
 
 # Assumptions in Structural Causal Models
+
+## Mathematical Components
 
@@
 
-We can craete graphical models based on various different algorithms.
+## Independence of Noise and Mechanism
+
+**Definition:** Independence 
+*The causal generative process of a system’s variables is composed of autonomous modules that do not inform or influence each other.*
+*In the probabilistic case, this means that the conditional distribution of each variable given its causes (i.e., its mechanism) does not inform or influence the other conditional distributions*,
+*In case we have only two variables, this reduces to an independence between the cause distribution and the mechanism producing the effect distribution* [@peters_elements_2017].
+
+If we specify the causal structure correctly:
+
+- possible to undertake local intervention -> change f(smoking), regardless of f(cancer|smoking)
+- these components are autonomous objects -> set of autonomous assignments
+
+**Independence of Noise:**
+
+The way we view noise in SCMs differs from the classical view in regression analysis.
+Therefore, I will briefly clarify essential differences. 
+Noise variables are our latent variables which are the parent nodes of our child nodes of interest [@hardtrecht].
+In classical regression analysis, such as an ordinary least squares regression, we employ the Gauss Markov Assumptions. 
+The exogeneity assumption, one of the five Gauss Markov Assumptions, suggests that the error terms are uncorrelated with our features [@wooldridge2010econometric].
+For the SCM, the error terms are driving our features and a pivotal component of the model specification [@hardtrecht].
+To ensure that we are correctly specifying mechanisms and our underlying model structure, one needs to ensure independence of these noise terms. 
+As mentioned in the definition for SCMs, we would mathematical notate this independence as follows:
+
+$N_{F} \perp \!\!\! \perp N_{C}$
+
+Scholarship in causality established the causal sufficiency condition:
+
+**Definition:** Causal Sufficiency
+
+*A set of variables X is usually said to be causally sufficient if there is no hidden common cause $C \notin X$ that is causing more than one variable in X* [@peters_elements_2017; @spirtes2010introduction]
+
+By ensuring that this condition holds, we can easier disentangle underlying common causes, avoiding a violation of the independence of our noise terms.
+Independence of noise is one form of causal sufficiency [@peters_elements_2017].
+
+There are two related dimensions to consider namely the informational aspect of independence  and implications for  modularity. 
+If the independence conditions holds, and the cause and effect variables are independent, the cause and effect variable do not contain information about each other. 
+Modularity [@pearl2009causality] describes the advantage of being able to treat these variables as distinct modules.
+This means, even if we have a change in one variable (cause or effect), we can disentangle our variable as unique modular components.
+Especially looking at machine learning, domain shift and covariate shift are problematic for classical tools. [@peters_elements_2017]
+By using these independence assumptions in SCMs, we can accommodate these settings.
+
+Looking at our smoking example:
+
+$$
+P(Cancer|Smoking)=P(Smoking|Cancer) \times P(Cancer)
+$$
+
+**Markov Condition**
+
+@pearl2009causality defines the markov condition as follows:
+
+**Definition:** Causal Markov Condition
+
+*Every Markovian causal model M induces a distribution P(x1,…, xn) that satisfies the parental Markov condition relative the causal diagram G associated with M; that is, each variable Xi is independent of all its nondescendants, given its parents PAi in G.*
+
+**Mechanism**
+
+\begin{equation}
+\begin{aligned}
+p(a, t) &=p(a \mid t) p(t) \\
+&=p(t \mid a) p(a)
+\end{aligned}
+\end{equation}
+
+\begin{equation}
+p^{\mathrm{o}}(a, t)=p(t \mid a) p^{\mathrm{o}}(a) \text { and } p^{\mathrm{s}}(a, t)=p(t \mid a) p^{\mathrm{s}}(a) .
+\end{equation}
+
+$$
+\begin{array}{l}
+P_{x}(v)=\prod_{\left\{i \mid V_{i} \notin X\right\}} P\left(v_{i} \mid p a_{i}\right)\\
+\text { for all } \boldsymbol{v} \text { consistent with } x
+\end{array}
+$$
+
+# Directed Acyclic Graphs
 
@@
 Graphical models developed a mathematical langauge to assess certain conditions.
-One pivotal feature is the d-separation [@judea2010introduction]:
 
-**Definition 2** ( $d$ -separation) A set $S$ of nodes is said to block a path $p$ if either (i) $p$ contains at least one arrow-emitting node that is in $S$, or $($ ii $)$ p contains at least one collision node that is outside $S$ and has no descendant in $S .$
-If $S$ blocks all paths from $X$ to $Y$, it is said to " $d$ -separate $X$ and $Y, "$ and then, $X$ and $Y$ are independent given $S$, written $X \perp \!\!\! \perp Y \mid S .$
+We can create graphical models based on various algorithms.
 
-Select subset of variables with backdoor criterion:
+We can obtain a graphical model based on (a single) observational dataset through conditional independence testing.
+The graphical approach to recover a graph from an observaational dataset requires two assumptions namely jointly independent noise terms and faithfulness. 
+Intuitively speaking, this method examines how noise spreads [@peters_elements_2017]. 
+Two other methods are ICM^[The intuition here is that the noise terms cultivate the footprint of the assignments.
+For more information see: @shajarisales2015telling] and using an additive noise model^[This method uses regression and conditional independence testing to disentangle graphical structures.
+For further information see @mooij2016distinguishing].
 
-## Independence
+# SCMs and Time
 
-**Definition:** Independence 
-*The causal generative process of a system’s variables is composed of autonomous modules that do not inform or influence each other.*
-*In the probabilistic case, this means that the conditional distribution of each variable given its causes (i.e., its mechanism) does not inform or influence the other conditional distributions*,
-*In case we have only two variables, this reduces to an independence between the cause distribution and the mechanism producing the effect distribution* [@peters_elements_2017].
+largely ingore time
 
-If we specify the causal structure correctly:
+time in mechanical modeling crucial [@peters_elements_2017].
 
-(a) possible to undertake local intervention -> change f(x), regardless of f(y|x)
-(b) these components are autonomous objects -> set of autonomous equations
+- Time in Social Sciences: Often Vague 
+- Time in Physical Sciences: Mechanical via **Differential equations** 
+- dependence on prior time point and change in time contribute to the value at time point
 
-Independence of Noise,
+Initial Value: 
+$$
+\mathbf{x}\left(t_{0}\right)=\mathbf{x}_{0}
+$$
 
-Noise independent so problematic behaviour and genetic code should be independent.
-
-**Mechanism**
-
-\begin{equation}
-\begin{aligned}
-p(a, t) &=p(a \mid t) p(t) \\
-&=p(t \mid a) p(a)
-\end{aligned}
-\end{equation}
-
-\begin{equation}
-p^{\mathrm{o}}(a, t)=p(t \mid a) p^{\mathrm{o}}(a) \text { and } p^{\mathrm{s}}(a, t)=p(t \mid a) p^{\mathrm{s}}(a) .
-\end{equation}
-
-**Definition:** Causal Sufficiency
-
-*A set of variables X is usually said to be causally sufficient if there is no hidden common cause $C \notin X$ that is causing more than one variable in X* [@peters_elements_2017; @spirtes2010introduction]
+Derivative of function x with respect to time t:
 
 $$
-P(Cancer|Smoking)=P(Smoking|Cancer) \times P(Cancer)
+\frac{d \mathbf{x}}{d t}=f(\mathbf{x}), \mathbf{x} \in \mathbb{R}^{d}
 $$
 
-**Markov Condition**
-
-@pearl2009causality defines the markov condition as follows:
-
-**Definition:** Causal Markov Condition
-
-*Every Markovian causal model M induces a distribution P(x1,…, xn) that satisfies the parental Markov condition relative the causal diagram G associated with M; that is, each variable Xi is independent of all its nondescendants, given its parents PAi in G.*
+Value of Function at time t + dt:
 
 $$
-\begin{array}{l}
-P_{x}(v)=\prod_{\left\{i \mid V_{i} \notin X\right\}} P\left(v_{i} \mid p a_{i}\right)\\
-\text { for all } \boldsymbol{v} \text { consistent with } x
+\mathbf{x}(t+d t)=\mathbf{x}(t)+d t \cdot f(\mathbf{x}(t))
+$$
+
+[@peters_elements_2017] provide a taxonomy of different methods (see table 1)
+
+$$\text{Table 1 Source: Peters et al. (2017) , Modelling Taxonomy}$$
+$$
+\begin{array}{|l|l|l|l|l|}
+\hline \text { model } & \begin{array}{l}
+\text { IID setting }
+\end{array} & \begin{array}{l}
+\text { changing } \\
+\text { distributions } 
+\end{array} & \begin{array}{c}
+\text { counter- } \\
+\text { factual } \\
+\text { questions }
+\end{array} & \begin{array}{l}
+\text { physical } \\
+\text { insight }
+\end{array} \\
+\hline \begin{array}{l}
+\text { mechanistic } \\
+\text { model }
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} \\
+\hline \begin{array}{l}
+\text { structural } \\
+\text { causal model }
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} \\
+\hline \begin{array}{l}
+\text { causal } \\
+\text { graphical } \\
+\text { model }
+\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} \\
+\hline \begin{array}{l}
+\text { statistical } \\
+\text { model }
+\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} \\
+\hline
 \end{array}
@@
 
-## Time
-
- why time is ignored.
+In the table we can see that traditional statistical models fair well in IID settings, where our observations are independent and identically distributed.
+But once we encounter a changing distribution such as an covariate shift or a domain shift, these statistical models are not suitable anymore.
+Statistical models are also unable to answer any counterfactual questions about hypothetical settings outside of our observational data and are unable to provide physical insights because we do not model time as an explicity.
+Causal graphical models are able to deal with changing distributions and fair well in the IID setting. 
+Nevertheless, they are also unable to answer counterfactual questions on its own and cannot provide physical insights. 
+SCMs are able to deal with the first three queries, but largely ignore time.
+By commonly making assumptions such as that the effect is acyclic, SCMs simplify time and treat focus on equilbrium states.
+The mechanistic view, where we model time as an explicit factor, enables us to answer all those queries. 
+Note that the growing complexity makes it very challenging to implement but relevant for natural sciences where time is pivotal to disentangle relationships [@mooij2013ordinary; @peters_elements_2017].
 
@@
 Pearl [@bareinboim2020pearl; @pearl2009causality] introduced the hierachy of causation to categorize different statistical and causal tools. 
-The hierarchy of causation contains (see table 1) three levels. 
-The higher the method on the hierarchy, the more information the method requires.
+The hierarchy of causation contains (see table 1) three levels, where the high methods on the hierarchy, the more information the method requires.
 This section discusses each of these methods and their respective advantages.
 
+$$\text{Table 2 Source: Pearl (2009) , Hierarchy of Causation}$$ 
 \footnotesize
@@
 | Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO            
-
-Table: Pearls Hierachy of Causation (2009)
-
 \normalsize
@@
 
-```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
+```{tikz,fig.cap="Probabilistic Model" ,fig.align="center", echo =F}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
 \usetikzlibrary{calc}
@@
     sharp corners=2pt,
-    inner sep=7pt,
-    node distance=3cm,
+    inner sep=6pt,
+    node distance=2cm,
     >=latex]
 \tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
-\node[my node] (S){S};
-\node[my node,right =1 cm of S](G){G};
-\node[my node] at ($(S)!0.5!(G)-(0pt,1.5cm)$) (C) {C};
-\draw[->] (S) -- (C);
+\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
+\node[my node, fill=gray!30] (p){S};
+\node[my node, fill=gray!30] at ($(p)!0.5!(p)-(0pt,1.5cm)$) (G) {G};
+\node[my node,right = 1 cm of G, fill=gray!30](C){C};
+\draw[->] (p) -- (C);
 \draw[->] (G) -- (C);
@@
 
-One extension of policy intervention is off-policy intervention. 
-**Off-policy intervention** models different intervention that is not in our historical data.
-To accommodate the lack of data, we can estimate the intervention either based on model-based (e.g. regression),model-free estimates (propensity scores), or a mixture of these methods. 
-For further information, see e.g. @oberst2019counterfactual or [@creager].
+One way of the most prominent ways to estimate casual effects of our treatment is the average treatment effect.
+The average treatment effect looks at the difference between treatment and control group.
+For the case of simplicity in our example, we want to look at whether smoking at all has an impact on lung cancer.
+Mathematically, this could as follows:
 
-$do(t)$ -> replace function $T:=f_T(\pi)$ with different conditional probability (or constant) [@pearl2009causality].
+\begin{equation}
+f(C|S=1) - f(C|S=0)
+\end{equation}
 
-**Counterfactuals:**
+where $S = 1$ means the participant smokes and $S=0$ means the participant does not smoke.
+Furthermore, it is crucial that we hold all other factors constant so looking at our example, we would compare people with the same genetic features to truely isolate the effect of smoking.
+The ATE is a population level effect [@hardtrecht].
+
+**Confounders**
+
+Confounding is the circumstance where the observational information (conditional probabilities in our data) and intervention are different.
+Essentially, we want to express our do-intervention based on conditional probabilities, but have no direct information in our data.
+To still estimate a conditional probability, we can use a mixture of existing conditional probabilities and estimate the desired probability based on the adjustment formula.
+For further information, see  [@hardtrecht].
+
+## Counterfactuals
 
@@
 
-# SCMs and Time
+## Considerations
 
-largely ingore time
+**If you...**
 
-time in mechanical modeling crucial [@peters_elements_2017].
+* ..only have conditional probabilities 
+* ..are only interested in association 
+* ..have very limited knowledge of your actual underlying structure
 
-- Time in Social Sciences: Often Vague 
-- Time in Physical Sciences: Mechanical via **Differential equations** 
-- dependence on prior time point and change in time contribute to the value at time point
+$\rightarrow$ you should probably go for a probabilistic model, association based model or a model based on conditional probabilities
 
-Initial Value: 
-$$
-\mathbf{x}\left(t_{0}\right)=\mathbf{x}_{0}
-$$
+**But if you**
 
-Derivative of function x with respect to time t:
+* ..have an understanding of underlying relationships in the data
+* ..are dealing with high stake settings 
+* ..want to simulate different interventions
 
-$$
-\frac{d \mathbf{x}}{d t}=f(\mathbf{x}), \mathbf{x} \in \mathbb{R}^{d}
-$$
-
-Value of Function at time t + dt:
-
-$$
-\mathbf{x}(t+d t)=\mathbf{x}(t)+d t \cdot f(\mathbf{x}(t))
-$$
-
-$$
-\begin{array}{|l|l|l|l|l|}
-\hline \text { model } & \begin{array}{l}
-\text { IID setting }
-\end{array} & \begin{array}{l}
-\text { changing } \\
-\text { distributions } 
-\end{array} & \begin{array}{c}
-\text { counter- } \\
-\text { factual } \\
-\text { questions }
-\end{array} & \begin{array}{l}
-\text { physical } \\
-\text { insight }
-\end{array} \\
-\hline \begin{array}{l}
-\text { mechanistic } \\
-\text { model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} \\
-\hline \begin{array}{l}
-\text { structural } \\
-\text { causal model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} \\
-\hline \begin{array}{l}
-\text { causal } \\
-\text { graphical } \\
-\text { model }
-\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} \\
-\hline \begin{array}{l}
-\text { statistical } \\
-\text { model }
-\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} \\
-\hline
-\end{array}
-$$
-
-Table: Source: @peters_elements_2017
+$\rightarrow$ you could probably go for a SCM
 
@@
 There are many advantages of structural causal models:
-(1) We are able to model latent fators forcing us to reconsider existing assumptions about the relationship in our data.
+(1) We are able to model latent factors forcing us to reconsider existing assumptions about the relationship in our data.
 (2) Further, we get a underlying graphical representation including a mathematical language for this graphical systems to test causal assumptions that are otherwise untestable.
 (3) Additionally, one is able to model queries beyond mere association going as far as dealing with hyptoethical situations.
-Simultaneously,
+
+Apart from the assumptions, causal graphics can help reseachers avoiding mistakes.
+To determine this causal graph, we can use conditional independence testing and assess assumptions about faithfulness and jointly independent noise terms.
+
+Looking at the causal hierachy, structural causal modelling enables us to compute interventions efficienctly given certain condition hold. 
+These conditions defintely not always hold true in every instance. 
+Especially looking at the
 

### 2021-06-11 paper_scm.Rmd
,

@@
 I accomodate this outlook with various considerations for the usage of SCMs. 
-The aim of this paper is to provide an example based, intuitive understanding of different concepts within causality using manifold elements from core literature on causality.
+The aim of this paper is to provide an intuitive understanding of different concepts within causality using manifold elements from core literature on causality.
 
@@
 \footnotesize
+
 |Method          | Action |  Example | Usage | 
@@
 | Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO
+
 \normalsize
@@
 
-In **policy intervention** we specify a different conditional probability $(S=s)$ for an equation.
+In **policy intervention** (see figure 4) we specify a different conditional probability $(S=s)$ for an equation.
 We can derive s from S, because we include information on the intervention distribution. 
@@
 To still estimate a conditional probability, we can use a mixture of existing conditional probabilities and estimate the desired probability based on the adjustment formula.
-For further information, see  [@hardtrecht].
+For further information, see  [@barocas].
 
@@
 
-Process is described as follows:
+Counterfactuals deal with hypothetical situations. 
+Philosophically speaking, we interpret these outcomes as actual outcomes^[In other frameworks such as the Potential outcome framework we interpret them as potential outcomes rather than true outcomes [@hardtrecht]].
+The two most prominent models that can obtain counterfactuals are the structural causal model and the potential outcome framework. 
+This section focuses on counterfactuals in structural causal models. 
+For information on how to obtain counterfactuals in the potential outcome framework, see @cunningham2021causal.
+
+We can describe the process as follows:
 
@@
 **Stable Unit Treatment Value Assumption {SUTVA}** 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
-
 **Consistency** The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' 
-
 **Ignorability** The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. 
 As suggested by [@hardtrecht], this condition ensures that we are dealing with a perfect randomized controlled trial. 
-
-- First two hold for Counterfactuals in SCM
-- third not testable but can check via backdoor criterion in SCM
-- Source: [@hardtrecht]
+The first two conditions hold for counterfactuals in structural causal models. 
+The third condition is an untestable assumption.
+Due to the backdoor criterion, we can nevertheless verify, that our model specification is consistent with this assumption [@hardtrecht].
 
@@
 
-**If you...**
+Structural causal models are data driven models [@hernan_invited_2015]. 
+There is a dependence on data quality.
+Henceforth, if our insights based on data are limited, or perhaps we are not really able to clearly define function relationships beyond conditional probabilities, structural causal models can be problematic.
+Furthermore, sometimes association-based knowledge is sufficient. 
+inevitably, the entry barrier to higher order causal models is higher in terms of required knowledge. 
+Not every situation accomodates these conditions.
 
-* ..only have conditional probabilities 
-* ..are only interested in association 
-* ..have very limited knowledge of your actual underlying structure
-
-$\rightarrow$ you should probably go for a probabilistic model, association based model or a model based on conditional probabilities
-
-**But if you**
-
-* ..have an understanding of underlying relationships in the data
-* ..are dealing with high stake settings 
-* ..want to simulate different interventions
-
-$\rightarrow$ you could probably go for a SCM
+Nevertheless, especially in a situation where we have high-stake repercussions based on our model, causal modelling has been a complementary tool.
+When dealing with policy implementations, that might be controversial or unethical in the real world, simulations are pivotal.
+Recently, reserach has fairness has adopted various concepts and developed new characterisitics such as counterfactual fairness [@kusner_counterfactual_2018].
 
@@
 
-Structural causal models are flexible simulators to disentangle causality for manifold different queries.
+Structural causal models are flexible simulators to disentangle causality for manifold different queries such as interventions and counterfactuals.
+We make a number of different assumptions when building a structural causal model.
+The most important assumption is the indepedence of mechanisms and independence of noise terms which are the core of our simulation.
+
 There are many advantages of structural causal models:
@@
 
-Looking at the causal hierachy, structural causal modelling enables us to compute interventions efficienctly given certain condition hold. 
+Looking at the causal hierachy, structural causal modelling enables us to compute interventions efficiently given certain condition hold. 
 These conditions defintely not always hold true in every instance.
-Especially looking at the
+The entry barriers to causal modeling is high and in involves considerable investment in terms of data and time.
 

### 2021-06-11 paper_scm.Rmd
sq

@@
 The most popular causal model is the structural causal model or in short SCM.
-The SCM is the non-parametric counterpart to structural equation models.
-Researchers on SCMs tried to distance themselves from common practices for structural equation models [@pearl2009causality; @peters_elements_2017].
-The geneticist and statistican Sewall Wright introduced the first ancestor of the SCM, the path analysis [@pearl2009causality].
-Path analysis now falls into the broader class of structural equation models.
-Path analysis is a structural equation model with one variable per indicator.
-
 The SCM is an expressive simulator to estimate causal relationships, accounting for latent factors. 
@@
 These assignments describe our variables in our model.
-Compared to probabilistic models, where we only specify conditional probabilities, the SCM actually enables the combination of latent variables and observational data.
-Conditional probabilities cannot represent latent variables because there is no conditional probability in our observational data for unobserved variables [@pearl2009causality].
+Compared to probabilistic models, where we only specify conditional probabilities, the SCM actually enables the combination of different sources of knowledge.
+In association based learning and typical statistical learning we only model based on observational data and conditional probabilities. 
+Conditional probabilities alone, cannot represent latent variables because there is no conditional probability in our observational data for unobserved variables [@pearl2009causality].
 
@@
 Additionally, this paper provides an overview of various other statistical and causal methods within the grand scheme of pearls causal hierachy.
-I accomodate this outlook with various considerations for the usage of SCMs. 
-The aim of this paper is to provide an intuitive understanding of different concepts within causality using manifold elements from core literature on causality.
+I accommodate this outlook with various considerations for the usage of SCMs. 
+The aim of this paper is to provide an intuitive understanding of different concepts within causality using elements from core literature on causality.
 
@@
 # Assumptions in Structural Causal Models
+
+The SCM is the non-parametric counterpart to structural equation models.
+Researchers on SCMs tried to distance themselves from common practices for structural equation models [@pearl2009causality; @peters_elements_2017].
+The geneticist and statistican Sewall Wright introduced the first ancestor of the SCM, the path analysis [@pearl2009causality; @tarka2018overview].
+Path analysis now falls into the broader class of structural equation models.
+Path analysis is a structural equation model with one variable per indicator.
 
@@
 If there are any changes in our distribution, e.g. smoking changes its distribution from $s$ to $s_{new}$ we are unable to accomodate these changes [@peters_elements_2017].
-Association-based methods are not equipped to accodate such modifications.
+Association-based methods are not equipped to accomodate such modifications.
 
-**Graph**
-
+There are ways to also graphically depict associational models. 
 One example is the bayesian network, which uses conditional probabilities instead of functions to describe the relationship between variables [@pearl2009causality]
@@
 
+While in this simple 3 feature case, the graphical illustration is rather straight forward. 
+But especially in a high dimensional space, the graphical models become convoluted.
+Henceforth, finding the correct specification without reconsidering all modelling assumptions can led to mistakes as discussed in section 2.
+
 ## Intervention:
 
-Here we can use @pearl2009causality do-calculus.
-
+Intervention deals with questions like 'what happens if'. 
+E.g. in our smoking example, we could look at e.g. 'what happens if people smoke less?' or, 'what happens if people smoke more?'.
+This higher order methods opens many different applications and is predominately possible in causal bayesian networks, reinforcement learning or markov decision processes.
+For interventions, we can use @pearl2009causality do-calculus.
 The do-calculus enables us to study the manipulation of parent nodes.
-
 Instead of merely seeing the co-occurence of variables, we can actively manipulate the conditional distribution of one variable.
 Commonly, one could use this method to evaluate different policies [@creager].
-In our smoking example we can actively set the smoking behavior to a fixed value or a different conditional probability.
-
-Joint distribution post intervention: $P_{S=s}(c,g)= P(c|S=s,g) \times P(g)$
-
+In our cancer example one can actively set the smoking behavior to a fixed value or a different conditional probability.
+The post intervention joint distribution would look as follows: 
+$P_{S=s}(c,g)= P(c|S=s,g) \times P(g)$ 
+where $S=s$ is the new probability or atomic value.
 There are various types of intervention.
-
 I will illustrate atomic intervention and policy intervention based on our smoking example.
-
-Suppose the government wants to examine, a different smoking behavior distribution will lower the number of lung cancer cases and medical expenses.
-This different smoking behaviour could be the result of increased taxes, or e.g. finesup to a $1000 (as in the case of Singapore) for smoking in prohibited areas.
-Alternative, the government could also undertake treatment by imposing higher tobacco taxes.
-In this case we solely focus on the effect of a different smoking behavior on lung cancer for the sake of simplicity.
 
@@
 There is a dependence on data quality.
+Evaluating different sources of information into causal models accordingly is an open problem in causal modelling [@spirtes2010introduction].
 Henceforth, if our insights based on data are limited, or perhaps we are not really able to clearly define function relationships beyond conditional probabilities, structural causal models can be problematic.
@@
 When dealing with policy implementations, that might be controversial or unethical in the real world, simulations are pivotal.
-Recently, reserach has fairness has adopted various concepts and developed new characterisitics such as counterfactual fairness [@kusner_counterfactual_2018].
+Recently, research has fairness has adopted various concepts and developed new characteristics such as counterfactual fairness [@kusner_counterfactual_2018].
 

### 2021-06-11 paper_scm.Rmd
recent

@@
 SCMs cast a model based on  asymmetric assignments and extend probabilistic models by specifying the entire data generating process rather than solely utilizing conditional probabilities.
-A SCM is build on a number of different assumptions namely the independence of noise, autonomy of mechanisms, causal sufficiency and the causal markov condition.
-I also discuss the role of time in causality and how differential equations enable us to model SCMs similar to a mechanical mdoel.
+A SCM is built on a number of different assumptions namely the Independence of Noise, Autonomy of Mechanisms and Causal Sufficiency.
+I also discuss the role of time in causality, focusing on how differential equations enable us to develop a SCM like a mechanical model.
 Another difference between thes SCM and other model specifications is the ability  of SCMs to address different queries such as *predictions*, *interventions* and *counterfactuals*. 
@@
 The most popular causal model is the structural causal model or in short SCM.
+The geneticist and statistican Sewall Wright introduced the first ancestor of the SCM, the path analysis in the 1920s [@pearl2009causality; @tarka2018overview].
+Path analysis now falls into the broader class of structural equation models (SEM) for the specific case of using one variable per indicator.
+SEMs are popular in fields like economics, psychology and sociology [@pearl2009causality].
+The SCM is the non-parametric counterpart to structural equation models.
+SCMs specify an underlying data generating process without the computational effort of creating a parametric model.
+Many scholars challenge the parameter-specification in a SEM [@pearl2012causal].
+^[Inconsistency refers to inherent nature of creating parameters based of observational data.
+Observational data  is seldom consistent.
+Henceforth respective estimates are often questioned.
+For further information see @hernan2008does]
+Researchers on SCMs tried to distance themselves from common practices in structural equation modeling [@pearl2009causality; @peters_elements_2017] by using this different name, although frequently they are used interchargably.
+
 The SCM is an expressive simulator to estimate causal relationships, accounting for latent factors.
@@
 To accommodate existing literature, this paper provides a gentle introduction to SCMs, focusing on the underlying assumptions we make when building a structural causal model.
-Specifically, this paper focuses on the independence of noise, independence of mechanisms, causal sufficiency, the role of time and the causal markov condition.
-Additionally, this paper provides an overview of various other statistical and causal methods within the grand scheme of pearls causal hierachy.
+Specifically, this paper focuses on the independence of noise, independence of mechanisms, causal sufficiency, the role of time and the Causal Markov Condition.
+Additionally, this paper provides an overview of various other statistical and causal methods within the grand scheme of pearls causal hierarchy.
 I accommodate this outlook with various considerations for the usage of SCMs. 
@@
 
+Subsection 2.1 introduces a baseline example for both a simple set of assignments and the underlying causal graph.
+Subsection 2.2 provides further context to assumptions in causal modelling. 
+Section 3 discusses causal graphs and how we derive underlying graphical representations based on single observational datasets.
+Section 4 examines causality and the role of time.
+Section 5 studies the causal hierachy.
+Section 6 suggests instructions on when and when not to employ a structural causal model.
+
 # Assumptions in Structural Causal Models
 
-The SCM is the non-parametric counterpart to structural equation models.
-Researchers on SCMs tried to distance themselves from common practices for structural equation models [@pearl2009causality; @peters_elements_2017].
-The geneticist and statistican Sewall Wright introduced the first ancestor of the SCM, the path analysis [@pearl2009causality; @tarka2018overview].
-Path analysis now falls into the broader class of structural equation models.
-Path analysis is a structural equation model with one variable per indicator.
-
-## Mathematical Components
-
-The structural causal model consists of a set of equations.
+A SCM contains a (sub-)set of autonomous equations.
 These equations are asymmetric assignments, because they are not bi-directional.
+Equations in causal models did not always have a concise notation (Pearl, 2009).
+First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
+Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
+This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
+The problem is that the equation is symmetric.
+The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
+This misconception has caused a lot of challenges.^[For more information see @pearl2009causality].
 We cannot change the sides of the equation as one can do with regular equations [@pearl2012causal].
@@
 In their definition, C is the cause and E is the effect.
+
+## Mathematical Components
 
@@
 **Definition:** Independence
-*The causal generative process of a system’s variables is composed of autonomous modules that do not inform or influence each other.*
+
+*"The causal generative process of a system’s variables is composed of autonomous modules that do not inform or influence each other.*
 *In the probabilistic case, this means that the conditional distribution of each variable given its causes (i.e., its mechanism) does not inform or influence the other conditional distributions*,
-*In case we have only two variables, this reduces to an independence between the cause distribution and the mechanism producing the effect distribution* [@peters_elements_2017].
+*In case we have only two variables, this reduces to an independence between the cause distribution and the mechanism producing the effect distribution"* [@peters_elements_2017].
 
@@
 To ensure that we are correctly specifying mechanisms and our underlying model structure, one needs to ensure independence of these noise terms. 
-As mentioned in the definition for SCMs, we would mathematical notate this independence as follows:
-
-$N_{F} \perp \!\!\! \perp N_{C}$
+As mentioned in the definition for SCMs, we would mathematical notate this independence as follows: $N_{F} \perp \!\!\! \perp N_{C}$
 
@@
 
-There are two related dimensions to consider namely the informational aspect of independence  and implications for  modularity. 
+There are two related dimensions to consider, namely the informational aspect of independence  and implications for  modularity. 
 If the independence conditions holds, and the cause and effect variables are independent, the cause and effect variable do not contain information about each other. 
@@
 $$
-
-**Markov Condition**
-
-@pearl2009causality defines the markov condition as follows:
-
-**Definition:** Causal Markov Condition
-
-*Every Markovian causal model M induces a distribution P(x1,…, xn) that satisfies the parental Markov condition relative the causal diagram G associated with M; that is, each variable Xi is independent of all its nondescendants, given its parents PAi in G.*
 
@@
 Nodes represent our different variables.
+The square nodes represent the latent variables.
+The circle nodes represent the observed variables.
 Edges depict the assignment equations.
 All edges are directed in the DAG.
+If we have some edges without arrows, we call that graph semi-directed.
+If we have all edges without arrows, we call that graph un-directed.
 An acyclic graph has no roots that cause itself (directly and indirectly) [@morgan_winship_2014].
@@
 
-may not find unique solution if cyclic in equilbrium [@peters_elements_2017]
-
+The problem of feedback loops is that we cannot always find unique solutions for the equilibrium state [@peters_elements_2017].
 Variables have incoming paths from their parent nodes.
-
-These DAGs can be built on theory.
-Another way to determine the DAG structure is using observational data. 
-One of the more prominent algorithm to estimate the underlying dag structure is the pc-algorithm [@kalisch2012causal] .
 
@@
 
-Graphical models developed a mathematical langauge to assess certain conditions.
-
-We can create graphical models based on various algorithms.
-
 We can obtain a graphical model based on (a single) observational dataset through conditional independence testing.
-The graphical approach to recover a graph from an observaational dataset requires two assumptions namely jointly independent noise terms and faithfulness. 
+The graphical approach to recover a graph from an observational dataset requires two assumptions namely jointly independent noise terms and faithfulness. 
 Intuitively speaking, this method examines how noise spreads [@peters_elements_2017]. 
+One of the more prominent algorithm to estimate the underlying DAG structure is the pc-algorithm [@kalisch2012causal]. 
 Two other methods are ICM^[The intuition here is that the noise terms cultivate the footprint of the assignments.
@@
 For further information see @mooij2016distinguishing].
+
+Graphical models developed a mathematical language to assess certain conditions.
 
@@
 Initial Value:
-$$
+
+\begin{equation}
 \mathbf{x}\left(t_{0}\right)=\mathbf{x}_{0}
-$$
+\end{equation}
 
@@
 
-$$
+\begin{equation}
 \frac{d \mathbf{x}}{d t}=f(\mathbf{x}), \mathbf{x} \in \mathbb{R}^{d}
-$$
+\end{equation}
 
@@
 
-$$
+\begin{equation}
 \mathbf{x}(t+d t)=\mathbf{x}(t)+d t \cdot f(\mathbf{x}(t))
-$$
+\end{equation}
 
@@
 But once we encounter a changing distribution such as an covariate shift or a domain shift, these statistical models are not suitable anymore.
-Statistical models are also unable to answer any counterfactual questions about hypothetical settings outside of our observational data and are unable to provide physical insights because we do not model time as an explicity.
+Statistical models are unable to answer any counterfactual questions about hypothetical settings outside of our observational data and are unable to provide physical insights because we do not model time as an explicitly.
 Causal graphical models are able to deal with changing distributions and fair well in the IID setting. 
-Nevertheless, they are also unable to answer counterfactual questions on its own and cannot provide physical insights. 
+Nevertheless, they are unable to answer counterfactual questions on its own and cannot provide physical insights. 
 SCMs are able to deal with the first three queries, but largely ignore time.
-By commonly making assumptions such as that the effect is acyclic, SCMs simplify time and treat focus on equilbrium states.
+By commonly making assumptions such as that the effect is acyclic, SCMs simplify time and treat focus on equilibrium states.
 The mechanistic view, where we model time as an explicit factor, enables us to answer all those queries. 
@@
 Association based methods are most prevalent and contain the largest class of methods.
-Standard statistical tools such as regression analysis, supervised and unsupervsied learning and bayesian networks all fall into this category.[@bareinboim2020pearl]
+Standard statistical tools such as regression analysis, supervised and unsupervsied learning and Bayesian Networks all fall into this category.[@bareinboim2020pearl]
 The underlying action for association is co-occurence.
@@
 There are ways to also graphically depict associational models. 
-One example is the bayesian network, which uses conditional probabilities instead of functions to describe the relationship between variables [@pearl2009causality]
+One example is the Bayesian Network, which uses conditional probabilities instead of functions to describe the relationship between variables [@pearl2009causality]
 In the probabilistic representation, we ignore latent factors [@creager; @pearl2009causality].
@@
 E.g. in our smoking example, we could look at e.g. 'what happens if people smoke less?' or, 'what happens if people smoke more?'.
-This higher order methods opens many different applications and is predominately possible in causal bayesian networks, reinforcement learning or markov decision processes.
+This higher order methods opens many different applications and is predominately possible in Causal Bayesian Networks, Reinforcement Learning or Markov Decision Processes.
 For interventions, we can use @pearl2009causality do-calculus.
@@
 
-**Stable Unit Treatment Value Assumption {SUTVA}** 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
-**Consistency** The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' 
-**Ignorability** The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. 
+As one can see, we can only cast the probability for latent factors if we specify them in our model.
+Henceforth, models such as the Bayesian Causal Network or a Markov Decision Process cannot provide counterfactuals because they do not include information on these factors.
+
+**Stable Unit Treatment Value Assumption {SUTVA}**
+
+'The treatment that one unit receives does not change the effect of treatment for any other unit.'
+
+**Consistency**
+
+The outcome Y agrees with the potential outcome corresponding to the treatment indicator.'
+
+**Ignorability**
+
+The potential outcomes are conditionally independent of treatment given some set of de-confounding variables.
+
 As suggested by [@hardtrecht], this condition ensures that we are dealing with a perfect randomized controlled trial. 
 The first two conditions hold for counterfactuals in structural causal models. 
-The third condition is an untestable assumption.
+The third condition is an un-testable assumption.
 Due to the backdoor criterion, we can nevertheless verify, that our model specification is consistent with this assumption [@hardtrecht].
 
-## Considerations
+# Considerations
 
@@
 There are many advantages of structural causal models:
-(1) We are able to model latent factors forcing us to reconsider existing assumptions about the relationship in our data.
-(2) Further, we get a underlying graphical representation including a mathematical language for this graphical systems to test causal assumptions that are otherwise untestable.
-(3) Additionally, one is able to model queries beyond mere association going as far as dealing with hyptoethical situations.
+ We are able to model latent factors forcing us to reconsider existing assumptions about the relationship in our data.
+ Further, we get a underlying graphical representation including a mathematical language for this graphical systems to test causal assumptions that are otherwise untestable.
+ Additionally, one is able to model queries beyond mere association going as far as dealing with hyptoethical situations.
 
@@
 These conditions defintely not always hold true in every instance.
-The entry barriers to causal modeling is high and in involves considerable investment in terms of data and time.
+The entry barriers to causal modeling are high and in involve considerable investments in terms of data and time.
 

### 2021-06-11 paper_scm.Rmd
scm

@@
 For many research problems, we want to understand the relationship between variables beyond co-occurrence.
+Algorithmic decision making based on co-occurrence is insufficient  in high stake settings [@bareinboim2020pearl].
+For many problems we want to understand causal relationships between variables.
+There are different approaches on how to model causal relationships.
 The most popular causal model is the structural causal model or in short SCM.
@@
 The SCM is the non-parametric counterpart to structural equation models.
-SCMs specify an underlying data generating process without the computational effort of creating a parametric model.
-Many scholars challenge the parameter-specification in a SEM [@pearl2012causal].
-^[Inconsistency refers to inherent nature of creating parameters based of observational data.
-Observational data  is seldom consistent.
-Henceforth respective estimates are often questioned.
-For further information see @hernan2008does]
+SCMs specify an underlying data generating process without the computational effort of creating a parametric form ^[ Observational data  is seldom consistent.
+Scholars often question how informative these estimates are.
+For further information see @hernan2008does and @pearl2012causal].
 Researchers on SCMs tried to distance themselves from common practices in structural equation modeling [@pearl2009causality; @peters_elements_2017] by using this different name, although frequently they are used interchargably.
@@
 These assignments describe our variables in our model.
-Compared to probabilistic models, where we only specify conditional probabilities, the SCM actually enables the combination of different sources of knowledge.
-In association based learning and typical statistical learning we only model based on observational data and conditional probabilities. 
+Compared to probabilistic models, where we only specify full joint distribution, the SCM actually enables the combination of different sources of knowledge.
+In association based learning and typical statistical learning we only model based on observational data and conditional probabilities given a specific static dataset.
+Henceforth, these methods perform poorly in cases of covariate shifts or domain changes.
+To account for these changes, we can specify an intervention distribution and latent factors. 
 Conditional probabilities alone, cannot represent latent variables because there is no conditional probability in our observational data for unobserved variables [@pearl2009causality].
@@
 To accommodate existing literature, this paper provides a gentle introduction to SCMs, focusing on the underlying assumptions we make when building a structural causal model.
-Specifically, this paper focuses on the independence of noise, independence of mechanisms, causal sufficiency, the role of time and the Causal Markov Condition.
-Additionally, this paper provides an overview of various other statistical and causal methods within the grand scheme of pearls causal hierarchy.
-I accommodate this outlook with various considerations for the usage of SCMs. 
 The aim of this paper is to provide an intuitive understanding of different concepts within causality using elements from core literature on causality. 
-
-The rest of the paper is structured as follows:
-
-Subsection 2.1 introduces a baseline example for both a simple set of assignments and the underlying causal graph.
-Subsection 2.2 provides further context to assumptions in causal modelling. 
+Specifically, this paper focuses on the independence of noise, independence of mechanisms, causal sufficiency (section 2).
 Section 3 discusses causal graphs and how we derive underlying graphical representations based on single observational datasets.
 Section 4 examines causality and the role of time.
-Section 5 studies the causal hierachy.
-Section 6 suggests instructions on when and when not to employ a structural causal model.
+Section 5 studies the causal hierarchy.
+I further provide considerations for the usage of SCMs (section 6). 
+Section 6 concludes the paper.
 
@@
 
+## Mathematical Components
+
 A SCM contains a (sub-)set of autonomous equations.
-These equations are asymmetric assignments, because they are not bi-directional.
-Equations in causal models did not always have a concise notation (Pearl, 2009).
-First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
+These equations are asymmetric assignments.
+Regular equations are bidirectional.
+Assignments are not bi-directional.
+Equations in causal models did not always have a concise notation.
 Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
-This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
-The problem is that the equation is symmetric.
-The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
+An algebraic equation would imply that $E=F$ and $C=E$ because the order has no concrete meaning in algebraic equations.
+Cause and effect would be interchangeable which is undisireable for causal modelling. 
+The initial $'='$ sign was replaced with the ':=' which is asymmetric and entails causal information.
 This misconception has caused a lot of challenges.^[For more information see @pearl2009causality].
-We cannot change the sides of the equation as one can do with regular equations [@pearl2012causal].
+
+There are various definitions for structural causal models. 
+Regardless, there is a consensus that a SCM contains two components namely an underlying causal graph and a set of assigments. 
 [@peters_elements_2017] define a SCM as follows:
@@
 $$
-*where* $N_{F} \perp \!\!\! \perp N_{C}$ *that is* $N_{F}$ *is independent of* $N_{C}$
+*where* $N_{E} \perp \!\!\! \perp N_{C}$ *that is* $N_{E}$ *is independent of* $N_{C}$
 
 In their definition, C is the cause and E is the effect.
+$N_C$ is the random noise variable for the cause. 
+$N_E$ is the noise variable for the effect variable.
 
-## Mathematical Components
-
-We can set up an epidemiological example with three variables, looking at the impact of problem behavior and genetic code on lung cancer. 
-Problem behavior is a latent variable, but we can use observational data to characterize problem behavior.
+for the subsequent paper i will use a real world expample.
+One common example stems from epidemiology.
+This example looks at the impact of problem behavior such as smoking on lung cancer. 
+Problem behavior is a conceptual term and latent because we cannot observe pure 'problem behaviour' in observational data. 
+We can use variables that reflect problem behaviour to study problem behaviour. 
 One example of problem behavior is smoking.
-Henceforth,to estimate problem behaviour, we can for instance ask participants how frequently they smoke.
-For genetic code, we could look for specific genes and and examine whether the presence of specific genes has an impact on getting cancer. 
+We can for instance ask participants how frequently they smoke.
 As we can see, these functions are driven by underlying latent variables.
@@
 \begin{equation}
-G:= f_G(U_G)
-\end{equation}
-\begin{equation}
-C:= f_C(S,G,U_C)
+C:= f_C(S,U_C)
 \end{equation}
@@
 $\{S\}$ - Frequency of smoking
-$\{G\}$ - Presence of specific genes
 $\{C\}$ - Lung cancer
@@
 
-[@pearl2009causality] describes the SCM a process based tool, because it enables researchers to reflect on their underlying assumptions. 
+[@pearl2009causality] describes the SCM a process-based tool, because it enables researchers to reflect on their underlying assumptions. 
 The SCM requires more assumptions and thought.
@@
 
+we can use the definition provided by [@peters_elements_2017].
+
 **Definition:** Independence
@@
 *In the probabilistic case, this means that the conditional distribution of each variable given its causes (i.e., its mechanism) does not inform or influence the other conditional distributions*,
-*In case we have only two variables, this reduces to an independence between the cause distribution and the mechanism producing the effect distribution"* [@peters_elements_2017].
+*In case we have only two variables, this reduces to an independence between the cause distribution and the mechanism producing the effect distribution"*.
 
-If we specify the causal structure correctly:
+Suppose we have only two variables smoking and cancer. 
+Typically, we can express the full joint probability as either:
 
-- possible to undertake local intervention -> change f(smoking), regardless of f(cancer|smoking)
-- these components are autonomous objects -> set of autonomous assignments
+\begin{equation}
+p(s,c)=p(s|c)p(c)
+\end{equation}
+\begin{equation}
+p(s,c)=p(c|s)p(s)
+\end{equation}
+
+In the first example, we define the probability of smoking given one has lung cancer and in the second we look at the conditional probability of lung cancer given smoking.
+The central idea is that if we specify the causal effect correctly, we can keep the conditional probability of lung cancer given smoking constant while changing the smoking distribution.
+Clearly, these interventions need to be sensible and do not work for both cases.
+Writing the probability as smoking behavior given lung cancer and changing lung cancer is an unreasonable intervention.
+If we specify the causal structure correctly we can undertake local changes on smoking without changing the conditional probability of lung cancer conditional on smoking.
+One example is looking at smoking behavior in a different country (domain shift).
+The independence of our mechanisms means that the mechanisms underlying cancer and smoking are independent.
+If these terms are independent, we can look at local intervention without rephrasing the entire model and keep the other probabilities invariant.
 
@@
 
-The way we view noise in SCMs differs from the classical view in regression analysis.
-Therefore, I will briefly clarify essential differences. 
-Noise variables are our latent variables which are the parent nodes of our child nodes of interest [@hardtrecht].
-In classical regression analysis, such as an ordinary least squares regression, we employ the Gauss Markov Assumptions. 
+Note, that the way we view noise in SCMs differs from the classical view in regression analysis.
+Therefore, I will briefly clarify the differences. 
+Noise variables are our latent variables.
+These unobserved variables are the parent nodes of our child nodes of interest [@hardtrecht].
+In classical regression analysis, such as an ordinary least squares regression model, we employ the Gauss Markov Assumptions to ensure that the estimator is the best linear unbiased estimator (BLUE). 
 The exogeneity assumption, one of the five Gauss Markov Assumptions, suggests that the error terms are uncorrelated with our features [@wooldridge2010econometric].
@@
 To ensure that we are correctly specifying mechanisms and our underlying model structure, one needs to ensure independence of these noise terms.
-As mentioned in the definition for SCMs, we would mathematical notate this independence as follows: $N_{F} \perp \!\!\! \perp N_{C}$
+As mentioned in the definition for SCMs, we would mathematical notate this independence as follows: $U_{C} \perp \!\!\! \perp U_{E}$ 
+where the noise terms of the cause and effect variables are independent.
+This is pivotal, because we want to derive conditional probabilities for local changes without having to rephrase other variables. 
+There are two related dimensions to consider, namely the informational aspect of independence  and implications for  modularity. 
+If the independence conditions holds, and the cause and effect variables are independent, the cause and effect variable do not contain information about each other. 
+Modularity [@pearl2009causality] describes the advantage of being able to treat these variables as distinct modules.
+This means, even if we have a change in one variable (cause or effect), we can disentangle our variable as unique modular components.
+Especially looking at machine learning, domain shift and covariate shift are problematic for classical tools. [@peters_elements_2017]
 
-Scholarship in causality established the causal sufficiency condition:
+The central issue of independence of noise variables is that they are latent and henceforth this assumption is untestable.
+We can only examine this condition in totality but not in isolation^[via the d-seperation criterion [@hardtrecht]].
+Un-testable assumptions are the reason why many statisticians have distanced themselves from causality in the past [@pearl2012causal].
+
+To accommodate the fact that we can never truly ensure that latent variables are independent, scholarship has proposed the causal sufficiency condition.
 
@@
 
-There are two related dimensions to consider, namely the informational aspect of independence  and implications for  modularity. 
-If the independence conditions holds, and the cause and effect variables are independent, the cause and effect variable do not contain information about each other. 
-Modularity [@pearl2009causality] describes the advantage of being able to treat these variables as distinct modules.
-This means, even if we have a change in one variable (cause or effect), we can disentangle our variable as unique modular components.
-Especially looking at machine learning, domain shift and covariate shift are problematic for classical tools. [@peters_elements_2017]
-By using these independence assumptions in SCMs, we can accommodate these settings.
-
-Looking at our smoking example:
-
-$$
-P(Cancer|Smoking)=P(Smoking|Cancer) \times P(Cancer)
-$$
-
-**Mechanism**
-
-\begin{equation}
-\begin{aligned}
-p(a, t) &=p(a \mid t) p(t) \\
-&=p(t \mid a) p(a)
-\end{aligned}
-\end{equation}
-
-\begin{equation}
-p^{\mathrm{o}}(a, t)=p(t \mid a) p^{\mathrm{o}}(a) \text { and } p^{\mathrm{s}}(a, t)=p(t \mid a) p^{\mathrm{s}}(a) .
-\end{equation}
-
-$$
-\begin{array}{l}
-P_{x}(v)=\prod_{\left\{i \mid V_{i} \notin X\right\}} P\left(v_{i} \mid p a_{i}\right)\\
-\text { for all } \boldsymbol{v} \text { consistent with } x
-\end{array}
-$$
-
 # Directed Acyclic Graphs
@@
 The most popular graphical model is the directed acyclic graph or in short DAG.
-A causal graph for a SCM contains endogenous and exogenous variables. 
-A DAG entails nodes and edges.
+Note that a DAG is a specific graphic model but not every SCM has an underlying DAG.
+Every SCM has an underlying graphical model.
+A causal graph is a more general graphical model, also including e.g. causal cyclic models.
+DAGs are predominately used because they are straightforward but simultaneously only apply if our model is truly directed and acyclic.
+A DAG entails nodes (endogenous and exogenous variables) and edges. 
+Nodes represent our different variables.
+Edges depict the assignment equations.
+All edges are directed in the DAG.
+If we have some edges without arrows, we call that causal graph semi-directed.
+If we have all edges without arrows, we call that causal graph un-directed.
+An acyclic graph has no roots that cause itself (directly and indirectly) [@morgan_winship_2014].
+This acyclic structure is important for the conditional probabilities [@forre2020causal].
+The problem of feedback loops is that we cannot always find unique solutions for the equilibrium state [@peters_elements_2017].
 
-Nodes represent our different variables.
+Figure 1 provides a basic structural causal model.
+The graph provides an extension of our real world example with another feature namely genetic features.
+As one can see, genetic features and smoking are autonous equations and we can derive their conditional distributions in a modular manner. 
+Further, the $U_s$, $U_g$ and $U_c$ are the expressive noise variables. 
 The square nodes represent the latent variables.
 The circle nodes represent the observed variables.
-Edges depict the assignment equations.
-All edges are directed in the DAG.
-If we have some edges without arrows, we call that graph semi-directed.
-If we have all edges without arrows, we call that graph un-directed.
-An acyclic graph has no roots that cause itself (directly and indirectly) [@morgan_winship_2014].
-
-This acyclic structure is important for the conditional probabilities [@forre2020causal].
-
-The problem of feedback loops is that we cannot always find unique solutions for the equilibrium state [@peters_elements_2017].
-Variables have incoming paths from their parent nodes.
+All edges are directed and there are no variables causing itself. 
+We are dealing with a DAG.
 
@@
 We can obtain a graphical model based on (a single) observational dataset through conditional independence testing.
-The graphical approach to recover a graph from an observational dataset requires two assumptions namely jointly independent noise terms and faithfulness. 
+The graphical approach to recover a graph from an observational dataset requires two assumptions namely jointly independent noise terms (which is assessed via the markov condition) and faithfulness.^[For further information, see [@morgan_winship_2014:1; @peters_elements_2017]].
+
 Intuitively speaking, this method examines how noise spreads [@peters_elements_2017].
@@
 For further information see @mooij2016distinguishing].
-
 Graphical models developed a mathematical language to assess certain conditions. 
+Two prominent examples are the causal markov condition and the backdoor criterion.
+The backdoor criterion allow us to select an admissible subset of variables [@pearl2009causality].
+The causal markov condition allows us to test independence in a graph, ensuring us that we are able to utilize the causal factorization [@peters_elements_2017].
 
@@
 
-largely ingore time
+In causal modelling we largely ignore time. 
+Most SCMs focus on specifying an equilibrium states post-intervention.
+Notions of time differ in different disciplines.
+Time in social sciences is often vague and less exact.
+Natural sciences such as physics treat time in a concise manner, by explicitly including time in models via differential equations [@peters_elements_2017;[@mooij2013ordinary].
+This is beneficial if we study non-equilibrium states because we can actual observe the react at different time points rather than only modelling the post intervention outcome.
+Dynamical modelling is becoming more relevant in social sciences.
+E.g. [@creager] suggest that static concepts in fair-algorithm decision making policies perform very differently if we model these policies as dynamic systems.
 
-time in mechanical modeling crucial [@peters_elements_2017].
-
-- Time in Social Sciences: Often Vague 
-- Time in Physical Sciences: Mechanical via **Differential equations** 
-- dependence on prior time point and change in time contribute to the value at time point
-
-Initial Value:
+We can recast structural causal models based on differential equations, too. 
+The first step of casting a SCM with time as a explicit factor is defining an initial state.
 
@@
 
-Derivative of function x with respect to time t:
+Note, that here the initial state is pivotal because there is a dependence on the prior time points.
+Subsequently, we need to define our function x at time point t.
+We can take a partial derivative with respect to time t of our function x.
 
@@
 \end{equation}
+
+Subsequently, we combine these elements and look at time t + change in t:
 
@@
 
-[@peters_elements_2017] provide a taxonomy of different methods (see table 1)
+Henceforth, there is a dependence on the initial state and then subsequently we combine this effect with the change.
+In our smoking example, one could perhaps think of a situation where we want to examine whether smoking at certain time points is more harmful (impact of smoking during cancer treatment looking at exact time points).
+
+For further context, [@peters_elements_2017] provide a taxonomy of different methods (see table 1).
+This table looks at different modelling approaches and how they perform in different settings.
 
@@
 
-Pearl [@bareinboim2020pearl; @pearl2009causality] introduced the hierachy of causation to categorize different statistical and causal tools. 
+@pearl2009causality] introduced the hierarchy of causation to categorize different statistical and causal tools. 
 The hierarchy of causation contains (see table 1) three levels, where the high methods on the hierarchy, the more information the method requires.
 This section discusses each of these methods and their respective advantages.
+I also included the usage of different methods and briefly discuss how they relate to the hierarchy based on [@bareinboim2020pearl].
 
@@
 The first level, association, requires the least information.
+This query deals with questions like 'what happend?'.
+In our smoking example, a possible question is e.g. 'what was the impact of smoking on lung cancer'?
 Association based methods are most prevalent and contain the largest class of methods.
-Standard statistical tools such as regression analysis, supervised and unsupervsied learning and Bayesian Networks all fall into this category.[@bareinboim2020pearl]
-The underlying action for association is co-occurence.
+Standard statistical tools such as regression analysis, supervised and unsupervised learning and Bayesian Networks all fall into this category [@bareinboim2020pearl].
+The underlying action for association is co-occurrence.
 As prominently criticized by @bender2021dangers, this reduces the number of questions we can answer, because methods are heavily dependent on the observational data. 
 In the context of deep learning and the advancement of natural language processing, @bender2021dangers suggest that many association-based methods results in stochastic parrots as opposed to natural langauge understanding.
-
 Association-based methods ignore external changes outside of our data.
@@
 Note, that the intervention distribution is only defined in high order methods. 
-
 Reiterating our lung cancer example, the full joint distribution looks as follows:
 
-$P(c,s,g)= P(s) \times P(c|s,g) \times P(g)$
+$P(c,s)= P(s) \times P(c|s)$
 
-If there are any changes in our distribution, e.g. smoking changes its distribution from $s$ to $s_{new}$ we are unable to accomodate these changes [@peters_elements_2017].
-Association-based methods are not equipped to accomodate such modifications.
+If there are any changes in our distribution, e.g. smoking changes its distribution from $s$ to $s_{new}$ we are unable to accommodate these changes [@peters_elements_2017].
+These conditional probabilities are typically not derived from functions $do(S=s)$ in these methods. 
+They are based on observational data for the given methods. 
+As mentioned in section 1, the amount of things we can derive based on a specific observational dataset is limited [@hernan_causal_nodate; @hernan2008does].
+Association-based methods are not equipped to accommodate such modifications.
 
-There are ways to also graphically depict associational models. 
-One example is the Bayesian Network, which uses conditional probabilities instead of functions to describe the relationship between variables [@pearl2009causality]
-In the probabilistic representation, we ignore latent factors [@creager; @pearl2009causality].
+There are ways to also graphically depict associations-based models. 
+One example is the Bayesian Network [@pearl2009causality].
+In the probabilistic representation (see figure 2), we ignore latent factors [@creager; @pearl2009causality].
+We can see, that this model is technically speaking also a directed acyclic graph.
 
@@
 
-While in this simple 3 feature case, the graphical illustration is rather straight forward. 
-But especially in a high dimensional space, the graphical models become convoluted.
-Henceforth, finding the correct specification without reconsidering all modelling assumptions can led to mistakes as discussed in section 2.
+While in this simple 2 feature case, the graphical illustration is rather straight forward. 
+But especially in a high dimensional space, these specifications become convoluted.
 
@@
 
-In **policy intervention** (see figure 4) we specify a different conditional probability $(S=s)$ for an equation.
+In **policy intervention** (see figure 4) we specify a different conditional probability $do(S=s)$ for an equation.
 We can derive s from S, because we include information on the intervention distribution. 
@@
 where $s$ is our new conditional probability. 
-
-One way of the most prominent ways to estimate casual effects of our treatment is the average treatment effect.
+One way of the most prominent ways to estimate casual effects of our treatment is the **average treatment effect**.
 The average treatment effect looks at the difference between treatment and control group.
@@
 where $S = 1$ means the participant smokes and $S=0$ means the participant does not smoke.
-Furthermore, it is crucial that we hold all other factors constant so looking at our example, we would compare people with the same genetic features to truely isolate the effect of smoking.
-The ATE is a population level effect [@hardtrecht].
+Furthermore, it is crucial that we hold all other factors constant (e.g. age, gender, genetic code).
 
@@
 Counterfactuals deal with hypothetical situations.
+Typically questions are: 'what would have happend if'.
+In our smoking case, suppose we want to see what would have happened if we wanted to include a different problem behavior that was not included in the observational data.
 Philosophically speaking, we interpret these outcomes as actual outcomes^[In other frameworks such as the Potential outcome framework we interpret them as potential outcomes rather than true outcomes [@hardtrecht]].
@@
 
+SCMs derive Counterfactuals.
 We can describe the process as follows:
 
-(a) Abduction: Cast probability $P(u)$ as conditional probability $P(u|\epsilon)$ 
-(b) Action: Exchange $(X = x)$ 
-(c) Prediction: Compute $(Y = y)$
+(a) Abduction: Cast probability $P(u_s)$ as conditional probability $P(u_s|\epsilon)$ 
+(b) Action: Exchange $(S = s')$ 
+(c) Prediction: Compute $(C = c')$
 
+In the first step, the abduction step, we cast the new conditional probability for our latent factor.
+This probabilities is based on the events in our data which are denoted as $\epsilon$.
 As one can see, we can only cast the probability for latent factors if we specify them in our model.
 Henceforth, models such as the Bayesian Causal Network or a Markov Decision Process cannot provide counterfactuals because they do not include information on these factors.
+In the second step, we exchange S for s'.
+In the third step, we compute the outcome c' for this hypothetical setting.
 
-**Stable Unit Treatment Value Assumption {SUTVA}**
+Regardless of whether dealing with counterfactuals in SCMs or in the potential-outcome-Framework, we need to ensure three assumptions.
+Here, I am using the definitions by [@hardtrecht].
 
+**Stable Unit Treatment Value Assumption {SUTVA}**: 
 'The treatment that one unit receives does not change the effect of treatment for any other unit.'
 
-**Consistency**
-
+**Consistency**: 
 The outcome Y agrees with the potential outcome corresponding to the treatment indicator.'
 
-**Ignorability**
+The first two conditions hold for counterfactuals in structural causal models.
+As suggested by [@hardtrecht], the third condition ensures that we are dealing with a perfect randomized controlled trial.
 
+**Ignorability**:
 The potential outcomes are conditionally independent of treatment given some set of de-confounding variables.
 
-As suggested by [@hardtrecht], this condition ensures that we are dealing with a perfect randomized controlled trial. 
-The first two conditions hold for counterfactuals in structural causal models. 
-The third condition is an un-testable assumption.
+We can never truly ensure that we are undertaking an experiment under perfect conditions, because we cannot ensure that we include all pivotal variables. 
+Therefore, the third condition is an un-testable assumption.
 Due to the backdoor criterion, we can nevertheless verify, that our model specification is consistent with this assumption [@hardtrecht].
+Note, that this is merely consistency and not a test that allows us to corroborate this assumptions entirely. 
+The backdoor criterion is unique to causal graphs.
+E.g. the potential outcome framework cannot utilize this test because they neglect graphical implementations of their underlying mathematical models.
 
@@
 Structural causal models are data driven models [@hernan_invited_2015]. 
-There is a dependence on data quality.
+There is a strong dependence on data quality.
+In a situation, in which we are certain that the quality of observational data is low, using a SCM is not advisable.
+Alternatively, if we have a very strong theoretical understanding, we might be more interested in other modelling approaches  such as Markov Decision Processes [@hernan_invited_2015].
 Evaluating different sources of information into causal models accordingly is an open problem in causal modelling [@spirtes2010introduction].
-Henceforth, if our insights based on data are limited, or perhaps we are not really able to clearly define function relationships beyond conditional probabilities, structural causal models can be problematic.
+If our insights based on data are limited, or perhaps we are not really able to clearly define function relationships beyond conditional probabilities, structural causal models can be problematic.
 Furthermore, sometimes association-based knowledge is sufficient.
 inevitably, the entry barrier to higher order causal models is higher in terms of required knowledge. 
-Not every situation accomodates these conditions.
-
+Not every situation accommodates these conditions. 
 Nevertheless, especially in a situation where we have high-stake repercussions based on our model, causal modelling has been a complementary tool.
@@
 
-Structural causal models are flexible simulators to disentangle causality for manifold different queries such as interventions and counterfactuals.
+Structural causal models are flexible simulators to disentangle causality for higher order queries such as interventions and counterfactuals.
 We make a number of different assumptions when building a structural causal model.
 The most important assumption is the indepedence of mechanisms and independence of noise terms which are the core of our simulation. 
-
 There are many advantages of structural causal models:
@@
 Additionally, one is able to model queries beyond mere association going as far as dealing with hyptoethical situations. 
-
 Apart from the assumptions, causal graphics can help reseachers avoiding mistakes.
-To determine this causal graph, we can use conditional independence testing and assess assumptions about faithfulness and jointly independent noise terms.
-
+To determine this causal graph.
+We can use conditional independence testing and assess assumptions about faithfulness and jointly independent noise terms in the graphical approach. 
 Looking at the causal hierachy, structural causal modelling enables us to compute interventions efficiently given certain condition hold. 
@@
 \newpage
+
+# Appendix
+
+\tiny
+
+ |Method         | CBN |  SCM
+|----------------|---------------------------|------------------------|
+| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
+|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
+|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
+|                |                                                                                                                      | 
+| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
+|                | $\boldsymbol{\cdot}$ Unstable(Nature Conditional prob.)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Equation)
+|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Equation)
+|                |                                                                                                                      | 
+|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
+|                | $\boldsymbol{\cdot}$ No information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors
+
+Table: Source: Pearl (2009)

### 2021-06-11 paper_scm.Rmd
final

@@
 bibliography: [ref.bib]
-nocite: '@*'
+#nocite: '@*'
 ---
@@
 SCMs cast a model based on asymmetric assignments and extend probabilistic models by specifying the entire data generating process rather than solely utilizing conditional probabilities.
+We build a SCM based on a set of assignments and an underlying causal graph. 
 A SCM is built on a number of different assumptions namely the Independence of Noise, Autonomy of Mechanisms and Causal Sufficiency.
-I also discuss the role of time in causality, focusing on how differential equations enable us to develop a SCM like a mechanical model.
-Another difference between thes SCM and other model specifications is the ability  of SCMs to address different queries such as *predictions*, *interventions* and *counterfactuals*. 
+I also discuss the role of time in causality, focusing on how differential equations for SCMs.
+Another difference between SCMs and other model specifications is the ability  of SCMs to address different queries such as *predictions*, *interventions* and *counterfactuals*. 
 These queries are part of Pearl's causal hierarchy (2009).
@@
 
-For many research problems, we want to understand the relationship between variables beyond co-occurrence.
 Algorithmic decision making based on co-occurrence is insufficient in high stake settings [@bareinboim2020pearl].
@@
 The most popular causal model is the structural causal model or in short SCM.
-The geneticist and statistican Sewall Wright introduced the first ancestor of the SCM, the path analysis in the 1920s [@pearl2009causality; @tarka2018overview].
-Path analysis now falls into the broader class of structural equation models (SEM) for the specific case of using one variable per indicator.
+The geneticist and statistician Sewall Wright introduced the first ancestor of the SCM,the path analysis in the 1910s-1920s [@pearl2009causality; @tarka2018overview].
+Path analysis now falls into the broader class of structural equation models (SEM)^[for the specific case of using one variable per indicator.]
 SEMs are popular in fields like economics, psychology and sociology [@pearl2009causality].
 The SCM is the non-parametric counterpart to structural equation models.
-SCMs specify an underlying data generating process without the computational effort of creating a parametric form ^[ Observational data  is seldom consistent.
+SCMs specify an underlying data generating process without the  burden of creating a correct parametric form^[ Observational data  is seldom consistent.
 Scholars often question how informative these estimates are.
 For further information see @hernan2008does and @pearl2012causal].
-Researchers on SCMs tried to distance themselves from common practices in structural equation modeling [@pearl2009causality; @peters_elements_2017] by using this different name, although frequently they are used interchargably.
-
-The SCM is an expressive simulator to estimate causal relationships, accounting for latent factors.
-Latent factors are unobserved factors [@pearl2009causality].
+The SCM is a more flexible version of the SEM.
+Researchers on SCMs tried to distance themselves from common practices in structural equation modeling [@pearl2009causality] by using this different name.
+Nowadays, scholars often used these terms interchangeably.
+The underlying purpose of the SCM is to simulate causal relationships.
+This includes underlying latent variables.
+Latent noise variables in the SCM are the core of the simulation.
+These latent factors define our variables of interest.
+In regression analysis we treat noise variables as ignorable factors. 
+We assume noise is un-related to our variable of interest.
+These perspectives are polar opposites.
 SCMs entail endogenous and exogenous variables.
-Exogenous variables should not be confused with latent variables.
-For reference, @peters_elements_2017 define endogenous and exogenous as follows: *Endogeneous variables are those that the modeler tries to understand, while exogenous ones are determined by factors outside the model, and are taken as given.*
-The basis of these SCMs is a set of equations, or more precisely assignments, providing functions to derive the conditional probabilities for our model [@hardtrecht].
+@peters_elements_2017 define endogenous and exogenous variables as follows: *"Endogeneous variables are those that the modeler tries to understand, while exogenous ones are determined by factors outside the model, and are taken as given."*
+The basis of these SCMs is a set of functional assignments.
+We can use these functions to derive the conditional probabilities for our model [@hardtrecht].
 These assignments describe our variables in our model.
-Compared to probabilistic models, where we only specify full joint distribution, the SCM actually enables the combination of different sources of knowledge.
-In association based learning and typical statistical learning we only model based on observational data and conditional probabilities given a specific static dataset.
-Henceforth, these methods perform poorly in cases of covariate shifts or domain changes.
-To account for these changes, we can specify an intervention distribution and latent factors. 
-Conditional probabilities alone, cannot represent latent variables because there is no conditional probability in our observational data for unobserved variables [@pearl2009causality].
-
-To accommodate existing literature, this paper provides a gentle introduction to SCMs, focusing on the underlying assumptions we make when building a structural causal model.
-The aim of this paper is to provide an intuitive understanding of different concepts within causality using elements from core literature on causality. 
-Specifically, this paper focuses on the independence of noise, independence of mechanisms, causal sufficiency (section 2).
+Probabilistic models only specify full joint distribution. 
+the SCM actually enables the combination of different sources of knowledge.
+Sources of information can include observational data but also theory.
+In association-based learning we typically only use observational data.
+These association-based methods perform poorly in cases of covariate shifts or domain changes.
+To account for these changes,  the SCM includes a specification for different interventions (intervention distribution).
+Conditional probabilities alone, cannot represent latent variables.
+There is no conditional probability in our observational data for unobserved variables [@pearl2009causality].
+Hence, we cannot account for changes in our data.
+To accommodate existing literature, this paper provides a gentle introduction to SCMs.
+The aim of this paper is to provide an intuitive understanding of fundamental concepts within causality.
+I focus on the underlying assumptions in SCMs.
+This paper focuses on the independence of noise, independence of mechanisms, causal sufficiency (section 2).
 Section 3 discusses causal graphs and how we derive underlying graphical representations based on single observational datasets.
@@
 Section 5 studies the causal hierarchy.
-I further provide considerations for the usage of SCMs (section 6). 
-Section 6 concludes the paper.
+I provide considerations for the usage of SCMs in section 6.
 
 # Assumptions in Structural Causal Models
+
+There are various components in SCMs.
+When we notate a SCM, we make different assumptions based on the design of these systems and the definition of a SCM.
+This sub-section focuses on the mathematical elements in a SCM, the system of equations.
+The second sub-section looks at the difference assumptions related to independence.
 
@@
 Equations in causal models did not always have a concise notation.
-Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
+Treating an equation as an algebraic equation led to confusion because those have no causal information.
 An algebraic equation would imply that $E=F$ and $C=E$ because the order has no concrete meaning in algebraic equations.
-Cause and effect would be interchangeable which is undisireable for causal modelling. 
-The initial $'='$ sign was replaced with the ':=' which is asymmetric and entails causal information.
+Cause and effect would be interchangeable which is undesirable for causal modelling. 
+The initial '=' sign was replaced with the ':=' which is asymmetric and entails causal information.
 This misconception has caused a lot of challenges.^[For more information see @pearl2009causality].
@@
 
-for the subsequent paper i will use a real world expample.
-One common example stems from epidemiology.
-This example looks at the impact of problem behavior such as smoking on lung cancer. 
+For the subsequent sections, I will use a real world expample.
+This example stems from epidemiology.
+This study looks at the impact of problem behavior such as smoking on lung cancer. 
 Problem behavior is a conceptual term and latent because we cannot observe pure 'problem behaviour' in observational data. 
-We can use variables that reflect problem behaviour to study problem behaviour. 
+We can use variables that reflect problem behavior to study problem behaviour. 
 One example of problem behavior is smoking.
@@
 Every structural causal model contains an underlying graphical model [@hardtrecht]. 
-This is one important feature that differentiates SCMs from other frameworks^[e.g. the Potential Outcome framework[@pearl2009causality]].
+This is one important feature that differentiates SCMs from other frameworks^[e.g. the Potential Outcome framework [@pearl2009causality]].
 
@@
 @hirano2001estimation suggest a method for covariate selection that according to @pearl2009myth favours bias-enhancing features in the propensity score.
-Further @bollen2013eight (2013) argue that @rosenbaum2002overt and @rubin2007design falsely declared that 'there is no reason to avoid adjustment for a variable describing subjects before treatment'.
+Further @bollen2013eight (2013) state that @rosenbaum2002overt and @rubin2007design falsely declared that *'there is no reason to avoid adjustment for a variable describing subjects before treatment'* which also is a severe error.
 
@@
 
-we can use the definition provided by [@peters_elements_2017].
+For the definition of independence, we can use the definition provided by [@peters_elements_2017]:
 
-**Definition:** Independence
+**Definition 2:** Independence
 
@@
 To ensure that we are correctly specifying mechanisms and our underlying model structure, one needs to ensure independence of these noise terms.
-As mentioned in the definition for SCMs, we would mathematical notate this independence as follows: $U_{C} \perp \!\!\! \perp U_{E}$ 
+As mentioned in the definition for SCMs, we would notate this independence as follows: $U_{C} \perp \!\!\! \perp U_{E}$ 
 where the noise terms of the cause and effect variables are independent.
@@
 Especially looking at machine learning, domain shift and covariate shift are problematic for classical tools. [@peters_elements_2017]
-
 The central issue of independence of noise variables is that they are latent and henceforth this assumption is untestable.
@@
 Un-testable assumptions are the reason why many statisticians have distanced themselves from causality in the past [@pearl2012causal].
+To accommodate the fact that we can never truly ensure that latent variables are independent, @spirtes2000causation proposed the causal sufficiency condition.
 
-To accommodate the fact that we can never truly ensure that latent variables are independent, scholarship has proposed the causal sufficiency condition.
+**Definition 3:** Causal Sufficiency
 
-**Definition:** Causal Sufficiency
-
-*A set of variables X is usually said to be causally sufficient if there is no hidden common cause $C \notin X$ that is causing more than one variable in X* [@peters_elements_2017; @spirtes2010introduction]
+*"A set of variables X is usually said to be causally sufficient if there is no hidden common cause $C \notin X$ that is causing more than one variable in X"* [@peters_elements_2017]
 
@@
 This acyclic structure is important for the conditional probabilities [@forre2020causal].
+Obtaining conditional probabilities in feedback-loops is challenging [@forre2020causal].
 The problem of feedback loops is that we cannot always find unique solutions for the equilibrium state [@peters_elements_2017].
+Most DAGs assume the effect is invariant to change over time.[@morgan_winship_2014]
+For further information on the role of time in causal modelling, see section 3.
 
@@
 We can obtain a graphical model based on (a single) observational dataset through conditional independence testing.
-The graphical approach to recover a graph from an observational dataset requires two assumptions namely jointly independent noise terms (which is assessed via the markov condition) and faithfulness.^[For further information, see [@morgan_winship_2014:1; @peters_elements_2017]].
+The graphical approach to recover a graph from an observational dataset requires two assumptions namely jointly independent noise terms (which is assessed via the markov condition) and faithfulness.^[For further information, see [@morgan_winship_2014; @peters_elements_2017]].
 
@@
 
-Subsequently, we combine these elements and look at time t + change in t:
-
-Value of Function at time t + dt:
+Thereafter, we combine these elements and look at time t + change in dt:
 
@@
 
-In the table we can see that traditional statistical models fair well in IID settings, where our observations are independent and identically distributed.
-But once we encounter a changing distribution such as an covariate shift or a domain shift, these statistical models are not suitable anymore.
+In table 1 we can see that traditional statistical models fair well in IID settings, where our observations are independent and identically distributed.
+Once we encounter a changing distribution such as an covariate shift or a domain shift, these statistical models are not suitable anymore.
 Statistical models are unable to answer any counterfactual questions about hypothetical settings outside of our observational data and are unable to provide physical insights because we do not model time as an explicitly.
@@
 Nevertheless, they are unable to answer counterfactual questions on its own and cannot provide physical insights.
+Note that we can derive the underlying causal graphs from a SCM.
+This table examines graphical models in isolation apart from the SCM.
 SCMs are able to deal with the first three queries, but largely ignore time.
@@
 In our smoking example, a possible question is e.g. 'what was the impact of smoking on lung cancer'?
-Association based methods are most prevalent and contain the largest class of methods.
+association-based methods are most prevalent and contain the largest class of methods.
 Standard statistical tools such as regression analysis, supervised and unsupervised learning and Bayesian Networks all fall into this category [@bareinboim2020pearl].
@@
 Note, that the intervention distribution is only defined in high order methods. 
-Reiterating our lung cancer example, the full joint distribution looks as follows:
-
-$P(c,s)= P(s) \times P(c|s)$
+Reiterating our lung cancer example, the full joint distribution looks as follows: $P(c,s)= P(s) \times P(c|s)$
 
@@
 E.g. in our smoking example, we could look at e.g. 'what happens if people smoke less?' or, 'what happens if people smoke more?'.
-This higher order methods opens many different applications and is predominately possible in Causal Bayesian Networks, Reinforcement Learning or Markov Decision Processes.
+This higher order methods opens many different applications and is predominately possible in Causal Bayesian Networks (CBN), Reinforcement Learning (RL) or Markov Decision Processes (MDP).
+Note that while CBNs are capable of computing interventions, they are computationally more costly compared to SCMs [@pearl2009causality].
+SCMs are built on functions which are inherently better capable of computing changes.
+For further informaation on the computational difference, see the appendix. 
 For interventions, we can use @pearl2009causality do-calculus.
@@
 \begin{equation}
-S:= s
+S:= c
 \end{equation}
@@
 In **policy intervention** (see figure 4) we specify a different conditional probability $do(S=s)$ for an equation.
-We can derive s from S, because we include information on the intervention distribution. 
+We can derive s from S, because we include information on the intervention distribution in our latent variables.
 This information cannot be obtained, if we directly specify our model as conditional probabilities.
@@
 where $s$ is our new conditional probability. 
-One way of the most prominent ways to estimate casual effects of our treatment is the **average treatment effect**.
+One of the most prominent ways to estimate casual effects of our treatment is the **average treatment effect**.
 The average treatment effect looks at the difference between treatment and control group.
@@
 
-where $S = 1$ means the participant smokes and $S=0$ means the participant does not smoke.
+Where $S = 1$ means the participant smokes and $S=0$ means the participant does not smoke.
 Furthermore, it is crucial that we hold all other factors constant (e.g. age, gender, genetic code).
 
-**Confounders**
-
+**Confounders:**
 Confounding is the circumstance where the observational information (conditional probabilities in our data) and intervention are different.
@@
 In the first step, the abduction step, we cast the new conditional probability for our latent factor.
-This probabilities is based on the events in our data which are denoted as $\epsilon$.
+These probabilities is based on the events in our data which are denoted as $\epsilon$.
 As one can see, we can only cast the probability for latent factors if we specify them in our model.
@@
 Regardless of whether dealing with counterfactuals in SCMs or in the potential-outcome-Framework, we need to ensure three assumptions.
-Here, I am using the definitions by [@hardtrecht].
+Here, I am using the definitions by [@hardtrecht]:
 
-**Stable Unit Treatment Value Assumption {SUTVA}**: 
-'The treatment that one unit receives does not change the effect of treatment for any other unit.'
+**Stable Unit Treatment Value Assumption {SUTVA}:**
+*"The treatment that one unit receives does not change the effect of treatment for any other unit."*
 
 **Consistency**: 
-The outcome Y agrees with the potential outcome corresponding to the treatment indicator.'
+*"The outcome Y agrees with the potential outcome corresponding to the treatment indicator."*
 
@@
 **Ignorability**:
-The potential outcomes are conditionally independent of treatment given some set of de-confounding variables.
+*"The potential outcomes are conditionally independent of treatment given some set of de-confounding variables."*
 
@@
 The backdoor criterion is unique to causal graphs.
-E.g. the potential outcome framework cannot utilize this test because they neglect graphical implementations of their underlying mathematical models.
+E.g. the potential outcome framework cannot utilize this test because the framework neglects underlying graphical models.
 
@@
 Furthermore, sometimes association-based knowledge is sufficient.
-inevitably, the entry barrier to higher order causal models is higher in terms of required knowledge. 
-Not every situation accommodates these conditions. 
-Nevertheless, especially in a situation where we have high-stake repercussions based on our model, causal modelling has been a complementary tool.
+Not every setting is a high-stake decision making setting.
+inevitably, the entry barrier to higher order causal models is higher in terms of required knowledge and resources. 
+Not every situation accommodates these conditions nor requires them.
+If we have cyclic structures, such as feedback loops, SCM are usable but very complicated.
+For further work on this issue see [@mooij2013ordinary; @forre2020causal].
+
+Nevertheless, in a situation where we have high-stake repercussions based on our model, causal modelling has been a complementary tool.
 When dealing with policy implementations, that might be controversial or unethical in the real world, simulations are pivotal.
-Recently, research has fairness has adopted various concepts and developed new characteristics such as counterfactual fairness [@kusner_counterfactual_2018].
+Recently, research on fairness has adopted various concepts and developed new characteristics such as counterfactual fairness [@kusner_counterfactual_2018].
+Further fairness research also looks at latent sensitive features such as race and allows us to model bias [@creager].
+Another interested area is simulations for situations where our historical data was bias. 
+One example is the use-cases by @ensign2018runaway, examining bias in predictive policing based on algorithmic decision making.
+Their research suggest that historical bias is reinforced through further racial targeting of drug-related crimes in Oakland based on historical records.
+In these cases, we can use SCMs to simulate domain shift and escape reinforcing biases.
 
@@
 
-Structural causal models are flexible simulators to disentangle causality for higher order queries such as interventions and counterfactuals.
-We make a number of different assumptions when building a structural causal model.
-The most important assumption is the indepedence of mechanisms and independence of noise terms which are the core of our simulation. 
-There are many advantages of structural causal models:
-We are able to model latent factors forcing us to reconsider existing assumptions about the relationship in our data.
-Further, we get a underlying graphical representation including a mathematical language for this graphical systems to test causal assumptions that are otherwise untestable.
-Additionally, one is able to model queries beyond mere association going as far as dealing with hyptoethical situations. 
-Apart from the assumptions, causal graphics can help reseachers avoiding mistakes.
-To determine this causal graph.
-We can use conditional independence testing and assess assumptions about faithfulness and jointly independent noise terms in the graphical approach. 
-Looking at the causal hierachy, structural causal modelling enables us to compute interventions efficiently given certain condition hold. 
-These conditions defintely not always hold true in every instance.
-The entry barriers to causal modeling are high and in involve considerable investments in terms of data and time.
+Association-based learning is insufficient to answer advanced queries in high-stake situations. 
+Causal Modeling provides the necessary toolkit to deal with questions of  higher order causal understanding. 
+The most prominent causal model is the structural causal model, containing mathematical and graph-theoretic elements. 
+SCMs are flexible simulators to disentangle causality for higher order queries (e.g. interventions, counterfactuals).
+SCMs enable us to consider latent factors, forcing us to re-evaluate existing assumptions in our model.
+We make a number of  assumptions when building a structural causal model.
+The most important assumption is the independence of mechanisms and independence of noise terms.
+Note, that noise terms are the elementary core in SCMs. 
+Independence of mechanisms ensures that are mechanisms we can make local changes without re-specifying the entire model.
+Independence of Noise ensures that we do not misspecify common causes.
+Independence of noise is an un-testable assumption.
+We can use the causal sufficiency condition to at least verify our set of variables is consistent with these assumptions.
+Every SCM also entails a causal graph.
+We can reconstruct graphical models from our observational based on further assumptions.
+One can use conditional independence testing and assess assumptions about faithfulness and jointly independent noise terms in the graphical approach. 
+These graphs come with a mathematical language to test causal assumptions that are otherwise un-testable and merely assumed to hold true(e.g. backdoor-criterion).
+The entry barriers to causal modeling are high .
+SCMs demand considerable knowledge about the model.
+Further, there are also still many open questions in causal modelling [@spirtes2010introduction].
+One example is how to weight different data sources and their relative value.
+This issue is of great importance when mixing different data sources.  
+Regardless, SCMs provide the building blocks for many different applications.
+While their entry requirements are high, the potential of these models is fruitful.
+Advances in algorithmic decision making suggest potential, especially looking at the intersection of  causal modeling and machine learning.
 
@@
 
-\tiny
+$$\text{Table 3: Source: Pearl (2009)}$$
+
+\footnotesize
 
@@
 
-Table: Source: Pearl (2009)
+\normalsize
